---
title: "Predicting time of auto accidents"
author: Francis Lauder
date: 2026-09-10
format:
  html:
    toc: true
    toc-depth: 3
    toc_float: true
    number-sections: false
    toc-location: left
    embed-resources: true
---

# Introduction
<br><br>


# Data Dictionary

<br><br>


# Libraries

<br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"

library(dplyr)
library(ggplot2)
#library(plotly)
library(RColorBrewer)
library(forcats)
library(tidycensus)

library(sf)

library(jsonlite)

library(tigris)

library(stringr)

library(leaflet)
library(htmlwidgets)
library(htmltools)
library(gt)
library(DT)
library(GGally)
library(lme4)
library(lmerTest)
library(rstanarm)
library(brms)
library(broom)
library(lubridate)
library(tidymodels)
library(performance)
library(boot.pval)
library(bayesplot)
library(units)
library(tidybayes)
library(marginaleffects)
library(collapse)
library(ggfortify)
library(bayesplot)
library(tidyr)






```

<br><br>


# Data Inspection and Cleaning

## US Accident Data


```{r}
#| echo: false
#| include: false
#| eval: false


US_accidents <- read.csv("/Users/francislauder/Documents/Data Sets/US Acciidents/US_Accidents_March23.csv")


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(US_accidents, "US_accidents.rds")

```

```{r}
#| echo: false
#| include: false
#| eval: false


US_accidents <- readRDS("US_accidents.rds")


```



```{r}
#| echo: false
#| include: false
#| eval: false



glimpse(US_accidents)


```



```{r}
#| echo: false
#| include: false
#| eval: false


MA_accidents <- US_accidents |> 
    filter(State == "MA")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


MA_accidents <- MA_accidents %>% 
  mutate(Weather_Timestamp = as.Date(Weather_Timestamp))


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(MA_accidents, "MA_accidents.rds")



```

```{r}
#| include: false


MA_accidents <- readRDS("MA_accidents.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



options(browser="false", scipen = 999, max.print = 20000)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


dim(MA_accidents)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



MA_accidents |> glimpse()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


summary(MA_accidents)


```




```{r}
#| code-fold: true
#| code-summary: "Show Code"



MA_accidents |> 
    count(Severity) |> 
    mutate(perc = round(n / sum(n),3)) |> 
    arrange(desc(perc))


```



Some features are not in the correct data types for use in our analysis.  We will tranform them to the required data types. 



First, Sverity,  will be transformed to a factor from an integer data type. 

```{r}
#| code-fold: true
#| code-summary: "Show Code"




MA_accidents_2 <- MA_accidents |> 
    mutate(Severity = as.factor(Severity))


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"


class(MA_accidents_2$Severity)


```




We will now extract the year, month,day, and hourm of the day from the Start_Time feature. 





```{r}
#| code-fold: true
#| code-summary: "Show Code"



MA_accidents_2 <- MA_accidents_2 |> 
    mutate(Year = year(Start_Time)) |> 
    mutate(Month = month(Start_Time)) |> 
    mutate(day_of_week = wday(Start_Time)) |>
    mutate(day_type = if_else(wday(Start_Time) %in% c(1,7), "Weekend", "Weekday")) |> 
    mutate(hour_of_day = hour(Start_Time))


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"


MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day, day_type) |> 
    head()


```

Day of week starts with Sunday (1) and ends with Saturday (7)

Hour of day starts at 1:00 am (1) and ends at 12:00 am (24)




```{r}
#| code-fold: true
#| code-summary: "Show Code"


MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day) |> 
    glimpse()



```



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


MA_accidents_2 <- MA_accidents_2 |> 
    mutate(Year = as.factor(Year)) |> 
    mutate(Month = as.factor(Month)) |> 
    mutate(day_of_week = as.factor(day_of_week)) |> 
    mutate(day_type = as.factor(day_type))



```



```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(MA_accidents_2, "MA_accidents_2.rds")


```

```{r}
#| include: false


MA_accidents_2 <- readRDS("MA_accidents_2.rds")



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"

MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day) |> 
    glimpse()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


MA_accidents_2 |> 
    count(Year)


```


Before continuing, we will filter our data into the years 2016-2021 for training and testing, and the year 2022 for predicting on unseen data.  Since the year 2023 appears to be incomplete, we will omit it from our analysis. 

<br>

Years 2016 through 2021


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_accidents_16_21 <- MA_accidents_2 |> 
    filter(Year %in% c("2016", "2017", "2018", "2019", "2020","2021"))


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_accidents_16_21, "ma_accidents_16_21.rds")

```

```{r}
#| include: false


ma_accidents_16_21 <- readRDS("ma_accidents_16_21.rds")



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


dim(ma_accidents_16_21)


```

Year 2022


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_accidents_22 <- MA_accidents_2 |> 
    filter(Year == "2022")


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_accidents_22, "ma_accidents_22.rds")


```

```{r}
#| include: false


ma_accidents_22 <- readRDS("ma_accidents_22.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


dim(ma_accidents_22)


```


For us to be able to calculate the time duration of accidents we'll first need to convert the start and endtimes to a date time object that includes year, month, day, hours, minutes, and seconds. 


```{r}
#| code-fold: true
#| code-summary: "Show Code"

ma_accidents_16_21_2 <- ma_accidents_16_21 |> 
    mutate(Start_Time = ymd_hms(Start_Time)) |> 
    mutate(End_Time = ymd_hms(End_Time))
```

```{r}
#| code-fold: true
#| code-summary: "Show Code"

ma_accidents_16_21_2 |> 
    select(Start_Time, End_Time) |> 
    glimpse()


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_2 |> 
    select(Start_Time, End_Time) |> 
    head()


```

Next, we'll create a new feature accident_duration, which will calcualte the duration of an accident in minutes. 


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_2 <- ma_accidents_16_21_2 |> 
    mutate(accident_duration = round(difftime(End_Time,Start_Time, units = "mins"), 0))


```

 We also will create LATTITUDE and LONGITUDE features from Start_Lat and Start_Lng. These features will be useful for retieving US Census data and mapping. 




```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_accidents_16_21_2$LATITUDE <- ma_accidents_16_21_2$Start_Lat

ma_accidents_16_21_2$LONGITUDE <- ma_accidents_16_21_2$Start_Lng


```


```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_accidents_16_21_2, "ma_accidents_16_21_2.rds")


```


```{r}
#| include: false


 ma_accidents_16_21_2 <- readRDS("ma_accidents_16_21_2.rds")


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_2 |> select(accident_duration) |>  head()


```

From the output, we see the accident duration is in minutes. 


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_2 |> select(LATITUDE, LONGITUDE) |> head()


```

Our output shows the two new featurres. 





## US Census Data











```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| include: false



Sys.getenv("CENSUS_API_KEY")


```





```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(long_lat_1 , "long_lat_1.rds")


```

```{r}
#| echo: false
#| include: false
#| eval: false

long_lat_1  <- readRDS("long_lat_1.rds")


```



```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(long_lat_2, "long_lat_2.rds")

```

```{r}
#| echo: false
#| include: false
#| eval: false


long_lat_2 <- readRDS("long_lat_2.rds")


```






```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(long_lat_3, "long_lat_3.rds")


```


```{r}
#| echo: false
#| include: false
#| eval: false


long_lat_3 <- readRDS("long_lat_3.rds")



```




```{r}
#| echo: false
#| include: false
#| eval: false



saveRDS(long_lat_4, "long_lat_4.rds")



```


```{r}
#| echo: false
#| include: false
#| eval: false


long_lat_4 <- readRDS("long_lat_4.rds")


```



```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(long_lat_5, "long_lat_5.rds")



```


```{r}
#| echo: false
#| include: false
#| eval: false



long_lat_5 <- readRDS("long_lat_5.rds")


```






```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(long_lat_6, "long_lat_6.rds")



```


```{r}
#| echo: false
#| include: false
#| eval: false

long_lat_6 <- readRDS("long_lat_6.rds")


```







```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



long_lat$census_code <- apply(long_lat_, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



long_lat$GEOID <- substr(long_lat$census_code, 1,11)


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(long_lat, "long_lat.rds")

```


```{r}
#| include: false


long_lat <- readRDS("long_lat.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


long_lat |>  head()


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


census_geo <- long_lat |> 
    select(-LATITUDE, -LONGITUDE)


```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(census_geo, "census_geo.rds")


```


```{r}
#| include: false

census_geo <- readRDS("census_geo.rds")



```


```{r}
#| code-fold: true
#| code-summary: "Show Code"




census_geo |> head()
```


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_accidents_16_21_3 <- cbind(ma_accidents_16_21_2, census_geo)



```

```{r}
#| echo: false
#| include: false
#| eval: false



saveRDS(ma_accidents_16_21_3, "ma_accidents_16_21_3.rds")


```


```{r}
#| include: false


ma_accidents_16_21_3 <- readRDS("ma_accidents_16_21_3.rds")


```




```{r}



ma_accidents_16_21_3 |> 
    summary()

```



There are five mssing values in the newly created GEOID.  These missing values (NAs) will be removed. 


```{r}


ma_accidents_16_21_4 <- ma_accidents_16_21_3 |> 
    filter(!is.na(GEOID))


```

```{r}


ma_accidents_16_21_4 |> 
    select(GEOID, census_code) |> 
    summary()


```


```{r}


ma_accidents_16_21_3 |> 
    select(GEOID,County, Wind_Speed.mph.) |> 
    filter(is.na(Wind_Speed.mph.)) |> 
    head(72)


```

```{r}


ma_accidents_16_21_4 |> 
    filter(!is.na(Wind_Speed.mph.)) |> 
    group_by(Severity) |> 
    summarise(median_wind_speed = median(Wind_Speed.mph.))


```


```{r}
#| eval: false


ma_accidents_16_21_4 <- ma_accidents_16_21_4 |> 
    filter(!is.na(Wind_Speed.mph.))


```

```{r}

ma_accidents_16_21_4 |> 
    summary()


```

```{r}


ma_accidents_16_21_4 |> 
    filter(!is.na(Visibility.mi.)) |> 
    group_by(Severity) |> 
    summarise(median_visibility = median(Visibility.mi.))


```

```{r}


ma_accidents_16_21_4|> 
    filter(!is.na(Precipitation.in.)) |> 
    group_by(Severity) |> 
    summarise(median_visibility = median(Precipitation.in.))


```



```{r}


ma_accidents_16_21_4  |> filter(accident_duration > 1000)  |> head()


```

```{r}

ma_accidents_16_21_4  |> filter(accident_duration > 1000)  |> tail()


```

```{r}


ma_accidents_16_21_4  |> filter(accident_duration <=1000) |> dim()


```


```{r}


ma_accidents_16_21_4 <- ma_accidents_16_21_4 |> 
    select(-Start_Lat, -End_Lat, -Start_Lng, -End_Lng, -State, -Zipcode, -Country, -Timezone, -Airport_Code, -Weather_Timestamp, -Wind_Chill.F., -Visibility.mi., -Precipitation.in., -Wind_Speed.mph.)


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_accidents_16_21_4, "ma_accidents_16_21_4.rds")

```

```{r}
#| include: false


ma_accidents_16_21_4 <- readRDS("ma_accidents_16_21_4.rds")


```

```{r}


ma_accidents_16_21_4 |> 
    summary()


```

Remove NAs fron Temperature feature.

```{r}


ma_accidents_16_21_5 <- ma_accidents_16_21_4|>
    filter(!is.na(Temperature.F.)) 


```

Remove NAs from Humidity feature. 

```{r}


ma_accidents_16_21_5 <- ma_accidents_16_21_5 |> 
    filter(!is.na(Humidity...))


```


Transform accident_duration from difftime to numeric. 

```{r}



ma_accidents_16_21_5$accident_duration <- as.numeric(ma_accidents_16_21_5$accident_duration)


```


```{r}


ma_accidents_16_21_5 |> 
    count(County)




```


Remove Dukes and Nantucket levels from Counties. 


```{r}


ma_accidents_16_21_5 <- ma_accidents_16_21_5 |> 
    filter(County  %in% c("Barnstable", "Berkshire", "Bristol", "Essex", "Franklin", "Hampden", "Hampshire", "Middlesex", "Norfolk", "Plymouth", "Suffolk", "Worcester"))


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_accidents_16_21_5, "ma_accidents_16_21_5.rds")


```




```{r}
#| include: false


ma_accidents_16_21_5  <- readRDS("ma_accidents_16_21_5.rds")


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"

ma_accidents_16_21_5 |> 
    count(County)


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"




ma_accidents_16_21_5 |> select(accident_duration, Severity, Start_Time,End_Time, City, LONGITUDE, LATITUDE, Street, Sunrise_Sunset, Civil_Twilight,Street)  |> 
  filter(accident_duration > 5000)  |> head()



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_accidents_16_21_5  |> select(accident_duration, Severity, Start_Time,End_Time, City, LONGITUDE, LATITUDE, Street, Sunrise_Sunset, Civil_Twilight,Street)  |> 
  filter(City == "Milton" & Street == "Adams St")  |> head()


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_accidents_16_21_5  |> select(accident_duration, Severity, Start_Time,End_Time, Year, Month, day_of_week, City, LONGITUDE, LATITUDE, Street, Sunrise_Sunset, Civil_Twilight,Street)  |> 
  filter(Year == "2021" & Month == "4" & day_of_week == "4")  |> tail()

```



```{r}
#| code-fold: true
#| code-summary: "Show Code"




ma_accidents_16_21_5 |> 
    select_if(is.numeric) |> 
    glimpse()


```





 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_6 <- ma_accidents_16_21_5 |> 
    rename(temperature = Temperature.F.)


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 


ma_accidents_16_21_5 |> 
    count(Wind_Direction)



 ```


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

 

ma_accidents_16_21_6$Wind_Direction[trimws(ma_accidents_16_21_6$Wind_Direction) == ""] <- "NR"


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

 


ma_accidents_16_21_6 <- ma_accidents_16_21_6 |> 
    mutate(Wind_Direction = as.factor(Wind_Direction))


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 




ma_accidents_16_21_6 <- ma_accidents_16_21_6 |> 
    mutate(Wind_Direction = recode(Wind_Direction,  "CALM" = "Calm", "NR" = "Calm", "ENE" = "E", "ESE" = "E", "East" = "E", "NE" = "N", "NNE" = "N", "NNW" = "N", "NW" = "N", "North" = "N", "SE" = "S", 
                                                    "SSE" = "S", "SSW" ="S", "SW" = "S", "South" = "S", "VAR" = "Variable", "WNW" = "W", "WSW" = "W", "West" = "W"))



 ```



 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_6 <- ma_accidents_16_21_6 |> 
    select(-Source, -Start_Time, -End_Time, -Distance.mi., -Pressure.in.,-Humidity... )


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_6 <- ma_accidents_16_21_6 |> 
    mutate(Wind_Direction = as.factor(Wind_Direction),
            Weather_Condition = as.factor(Weather_Condition),
            Amenity = as.factor(Amenity),
            Bump = as.factor(Bump),
            Crossing = as.factor(Crossing),
            Give_Way = as.factor(Give_Way),
            Junction = as.factor(Junction),
            No_Exit = as.factor(No_Exit),
            Railway = as.factor(Railway),
            Roundabout = as.factor(Roundabout),
            Station = as.factor(Station),
            Stop = as.factor(Stop),
            Traffic_Calming = as.factor(Traffic_Calming),
            Traffic_Signal = as.factor(Traffic_Signal),
            Turning_Loop = as.factor(Turning_Loop),
            Sunrise_Sunset = as.factor(Sunrise_Sunset),
            Civil_Twilight = as.factor(Civil_Twilight),
            Nautical_Twilight = as.factor(Nautical_Twilight),
            Astronomical_Twilight = as.factor(Astronomical_Twilight))


 ```

 ```{r}
#| echo: false
#| include: false
#| eval: false
 

saveRDS(ma_accidents_16_21_6 , "ma_accidents_16_21_6.rds")


 ```

 ```{r}
 #| include: false 
 

ma_accidents_16_21_6  <- readRDS("ma_accidents_16_21_6.rds")



 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"

 

ma_accidents_16_21_6 |> 
    select_if(is.factor) |> 
    summary()


 ```


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 


ma_accidents_16_21_6 |> 
    count(Crossing) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_6 |> 
    count(Wind_Direction) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_6 |> 
    count(County, Bump)



 ```




 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 


ma_accidents_16_21_7 <- ma_accidents_16_21_6 |> 
    mutate(Street = recode(Street,  "Mass Pike" = "I-90"))


 ```

 

 

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 



# Force-create the category column directly onto your existing dataframe
# (Replace 'Street' below if your column is lowercase 'street')
col_to_check <- ma_accidents_16_21_7$Street

ma_accidents_16_21_7$road_category <- "Uncategorized"

# Apply regex flags directly
ma_accidents_16_21_7$road_category[grepl("\\b(I|Interstate)[-\\s]?\\d+", col_to_check, ignore.case = TRUE)] <- "Interstate"
ma_accidents_16_21_7$road_category[grepl("\\b(Route|Rte|Rt|State Route|SR|State Hwy|MA)[. -]*\\d+", col_to_check, ignore.case = TRUE)] <- "MA Highway"
ma_accidents_16_21_7$road_category[grepl("\\b(US)[-\\s]?\\d+", col_to_check, ignore.case = TRUE)] <- "US Route"
ma_accidents_16_21_7$road_category[grepl("\\b(Rd|Road)\\b", col_to_check, ignore.case = TRUE)] <- "Road"
ma_accidents_16_21_7$road_category[grepl("way(\\b|[ .]+(N|S|E|W|NE|NW|SE|SW|North|South|East|West)?$)", col_to_check, ignore.case = TRUE)] <- "Way"
ma_accidents_16_21_7$road_category[grepl("\\b(St|Street)\\b", col_to_check, ignore.case = TRUE)] <- "Street"
ma_accidents_16_21_7$road_category[grepl("\\b(Ave|Avenue)\\b", col_to_check, ignore.case = TRUE)] <- "Avenue"
ma_accidents_16_21_7$road_category[grepl("\\b(Blvd|Boulevard)\\b", col_to_check, ignore.case = TRUE)] <- "Boulevard"
ma_accidents_16_21_7$road_category[grepl("\\b(Dr|Drive)\\b", col_to_check, ignore.case = TRUE)] <- "Drive"
ma_accidents_16_21_7$road_category[grepl("\\b(Ln|Lane)\\b", col_to_check, ignore.case = TRUE)] <- "Lane"
ma_accidents_16_21_7$road_category[grepl("\\b(Hwy|Highway)\\b", col_to_check, ignore.case = TRUE)] <- "Highway"
ma_accidents_16_21_7$road_category[grepl("\\b(Pkwy|Parkway)\\b", col_to_check, ignore.case = TRUE)] <- "Parkway"
ma_accidents_16_21_7$road_category[grepl("\\b(Brg|Bridge)\\b", col_to_check, ignore.case = TRUE)] <- "Bridge"
ma_accidents_16_21_7$road_category[grepl("\\b(Cir|Circle)\\b", col_to_check, ignore.case = TRUE)] <- "Circle"
ma_accidents_16_21_7$road_category[grepl("\\b(Trl|Trail)\\b", col_to_check, ignore.case = TRUE)] <- "Trail"
ma_accidents_16_21_7$road_category[grepl("\\b(Ct|Cor|Court)\\b", col_to_check, ignore.case = TRUE)] <- "Court"
ma_accidents_16_21_7$road_category[grepl("\\b(Ter|Terrace)\\b", col_to_check, ignore.case = TRUE)] <- "Terrace"
ma_accidents_16_21_7$road_category[grepl("\\b(Pl|Place)\\b", col_to_check, ignore.case = TRUE)] <- "Place"
ma_accidents_16_21_7$road_category[grepl("\\b(Tpke|Turnpike)\\b", col_to_check, ignore.case = TRUE)] <- "Turnpike"
ma_accidents_16_21_7$road_category[grepl("\\b(Sq|Square)\\b", col_to_check, ignore.case = TRUE)] <- "Square"
ma_accidents_16_21_7$road_category[grepl("\\b(Pth|Path)\\b", col_to_check, ignore.case = TRUE)] <- "Path"
ma_accidents_16_21_7$road_category[grepl("\\b(Conn|Connector)\\b", col_to_check, ignore.case = TRUE)] <- "Connector"
ma_accidents_16_21_7$road_category[grepl("\\b(Ext|Exit)\\b", col_to_check, ignore.case = TRUE)] <- "Exit"
ma_accidents_16_21_7$road_category[grepl("\\b(Rty|Rotary)\\b", col_to_check, ignore.case = TRUE)] <- "Rotary"
ma_accidents_16_21_7$road_category[grepl("\\b(Expy|Expressway)\\b", col_to_check, ignore.case = TRUE)] <- "Expressway"
ma_accidents_16_21_7$road_category[grepl("\\b(Tunl|Tunnel)\\b", col_to_check, ignore.case = TRUE)] <- "Tunnel"


 ```

 

 ```{r}
#| code-fold: true
#| code-summary: "Show Code" 
 
ma_accidents_16_21_6 |> count(Weather_Condition)


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code" 
 

ma_accidents_16_21_6 |> select(Weather_Condition, City) |> 
    filter(Weather_Condition == "") |> 
    head(15)


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 
ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    mutate(Weather_Condition = as.character(Weather_Condition))


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_7$Weather_Condition[trimws(ma_accidents_16_21_7$Weather_Condition) == ""] <- "NR"


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    mutate(Weather_Condition = as.factor(Weather_Condition))


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    mutate(Weather_Condition = recode(Weather_Condition,  "NR" = "Fair Partley Cloudy", "Light Rain / Windy" = "Light Rain", "Light Rain Shower" = "Light Rain", "Light Rain with Thunder" = "Light Rain", "Light Drizzle" = "Light Rain","Rain / Windy" = "Rain", "Thunderstorms and Rain" = "Rain", "Thunderstorm" = "Rain", "T-Storm" = "Rain","Thunder" = "Overcast", "Cloudy" = "Overcast", "Thunder / Windy" = "Overcast", "Thunder in the Vicinity" = "Overcast","Cloudy / Windy" = "Overcast", "Mostly Cloudy" = "Overcast", "Mostly Cloudy / Windy" = "Overcast","Light Thunderstorms and Rain" = "Rain","Heavy Rain / Windy" = "Heavy Rain", "Heavy Thunderstorms and Rain" = "Heavy Rain","Heavy T-Storm / Windy" = "Heavy Rain","Heavy T-Storm" = "Heavy Rain", "Light Snow / Windy" = "Light Snow", "Light Snow and Sleet" = "Light Snow", "Clear" = "Fair Partley Cloudy","Blowing Snow" = "Heavy Snow", "Heavy Snow / Windy" = "Heavy Snow", "Light Drizzle / Windy" = "Light Rain", "Mist" = "Light Rain", "Drizzle and Fog" = "Light Rain", "Fair / Windy" = "Fair Partley Cloudy", "Partly Cloudy" = "Fair Partley Cloudy", "Partly Cloudy / Windy" = "Fair Partley Cloudy", "Fair" = "Fair Partley Cloudy", "Scattered Clouds" = "Fair Partley Cloudy",
   "Fog / Windy" = "Fog", "Patches of Fog" = "Fog","Shallow Fog" = "Fog","Wintry Mix / Windy" = "Wintry Mix", "Light Freezing Drizzle" = "Wintry Mix" , "Light Freezing Rain" = "Wintry Mix","Smoke" = "Haze", "Squalls" = "Snow",
   "Light Freezing Fog" = "Wintry Mix", "Ice Pellets" = "Wintry Mix","Small Hail" = "Wintry Mix",  "Snow / Windy" = "Snow", "N/A Precipitation" = "Fair Partley Cloudy") )


 ```




   Now we can perform our binning.  Essentiall we will place hour of day into eight ranges.




 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 



ma_accidents_16_21_7 <- ma_accidents_16_21_7|> 
    mutate(hour_day_sections = case_when(
        between(hour_of_day,5,6) ~ "Dawn",
        between(hour_of_day,6,9) ~ "Early Morning",
        between(hour_of_day,9,12) ~ "Late Morning",
        between(hour_of_day,12,16) ~ "Early Afternoon",
        between(hour_of_day,16,18) ~ "Late Afternoon",
        between(hour_of_day,18,21) ~ "Evening",
        between(hour_of_day,21,24) ~ "Night",
        TRUE ~ "Middle of Night"

    ))



 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 


ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    mutate(hour_day_sections = as.factor(hour_day_sections))


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    select(-hour_of_day)


 ```

 ```{r}
#| echo: false
#| include: false
#| eval: false
 

saveRDS(ma_accidents_16_21_7, "ma_accidents_16_21_7.rds")



 ```


 ```{r}
#| include: false
 

ma_accidents_16_21_7 <- readRDS("ma_accidents_16_21_7.rds")


 ```

We now have eight categories of different parts of the day. 

 ```{r}
 

ma_accidents_16_21_7 |> 
    count(hour_day_sections) |> 
    mutate(perc = round(n/sum(n),3)*100) |> 
    arrange(desc(perc))



 ```



 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_7 |> 
    count(road_category)


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_7 |> 
    count(Weather_Condition)



 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_7 |> 
    select_if(is.factor) |> 
    summary()

 ```


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_7 |> 
    count(No_Exit) |> 
    mutate(perc = round(n/sum(n),4)) |> 
    arrange(desc(perc))


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 
ma_accidents_16_21_7 |> 
    count(County, No_Exit)

 ```


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_7|> 
    group_by(No_Exit ) |> 
    summarise(median_acc_dur = median(accident_duration))

```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_7 |> 
    count(Crossing) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 
ma_accidents_16_21_7 |> 
    count(Railway) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))



 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 


 ma_accidents_16_21_7 |> 
    count(County, Railway)

 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 
ma_accidents_16_21_7 |> 
    count(Give_Way) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 


ma_accidents_16_21_7 |> 
    count(County, Give_Way)


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_7 |> 
    count(Amenity) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 
ma_accidents_16_21_7 |> 
    count(County, Amenity)


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_7 |> 
    count(Junction) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_7 |> 
    count(County, Junction)


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_7 |> 
    count(Station) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))

 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 


ma_accidents_16_21_7 |> 
    count(County, Station)

 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_7 |> 
    count(Stop) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

 ma_accidents_16_21_7 |> 
    count(County, Stop)
 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_7 |> 
    count(Traffic_Calming) |> 
    mutate(perc = round(n/sum(n),3)) |> 
    arrange(desc(perc))

 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_7|> 
    group_by(Traffic_Calming) |> 
    summarise(median_acc_dur = median(accident_duration))



 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_7 |> 
    count(Stop) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_7 |> 
    count(Traffic_Signal) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

 ma_accidents_16_21_7 |> 
    count(County, Traffic_Signal)


 ```


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 


ma_accidents_16_21_7 |> 
    count(Wind_Direction)




 ```



 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_8 <- ma_accidents_16_21_7 |> 
    select(-No_Exit, -Traffic_Calming, -Nautical_Twilight, -Astronomical_Twilight, -Give_Way, -Railway, -Description, -Street, -Bump, -Roundabout, -Turning_Loop)


 ```

 ```{r}
#| echo: false
#| include: false
#| eval: false
 
saveRDS(ma_accidents_16_21_8, "ma_accidents_16_21_8.rds")


 ```

 ```{r}
 #| include: false
 


ma_accidents_16_21_8 <- readRDS("ma_accidents_16_21_8.rds")

 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_8 |> 
    glimpse()


 ```

 

 <br><br>

 

  <br><br>


 ```{r}
 
class(ma_accidents_16_21_8$hour_of_day)


 ```

  <br><br>

 



  <br><br>




Main Sections of the DayMorning (5:00 AM – 12:00 PM):Dawn / Sunrise: First light and when the sun comes up (approx. 5:00 AM – 6:00 AM).Early Morning: The start of typical morning routines (approx. 6:00 AM – 9:00 AM).Late Morning: Approaching midday (approx. 10:00 AM – 12:00 PM).Midday / Noon (12:00 PM): The exact middle of the day when the sun is highest.Afternoon (12:00 PM – 5:00 PM / 6:00 PM):Early Afternoon: Right after lunch (1:00 PM – 3:00 PM).Late Afternoon: Leading toward sunset (4:00 PM – 6:00 PM).Evening (5:00 PM / 6:00 PM – 9:00 PM):Sunset / Dusk: The sun goes down and twilight begins.Early/Late Evening: Transition time for relaxation and dinner.Night (9:00 PM – 4:00 AM / 5:00 AM):Late Night: Hours leading up to midnight (10:00 PM – 12:00 AM).Midnight: 12:00 AM, the middle of the night.Predawn / Middle of the Night: The quiet hours before the new day starts (1:00 AM – 4:00 AM).

 <br><br>




  <br><br>


  ```{r}
  
saveRDS(ma_accidents_16_21_9, "ma_accidents_16_21_9.rds")


  ```

  ```{r}
#| include: false
  

ma_accidents_16_21_9 <- readRDS("ma_accidents_16_21_9.rds")


  ```


  <br><br>




  <br><br>






```{r}





```

```{r}



ma_accidents_16_21_3  <- ma_accidents_16_21_2  |> 
    mutate(id = row_number())


```





# Census

## County




```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_county <- counties("MA", year = 2024)

```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_county, "ma_county.rds")


```

```{r}
#| include: false


ma_county <- readRDS("ma_county.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_county |> 
    glimpse()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_county |> 
    summary()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


print(ma_county |> select(NAME), n=14)


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_county_shape <- ma_county |> 
    filter(NAME != "Nantucket") |> 
    filter(NAME != "Dukes")


```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_county_shape , "ma_county_shape.rds")


```


```{r}
#| include: false


ma_county_shape  <- readRDS("ma_county_shape.rds")



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



print(ma_county_shape |> select(NAME), n=14) 


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_county_shape_2 <- ma_county_shape |> 
    select(COUNTYFP, GEOID, NAME, NAMELSAD, INTPTLAT, INTPTLON, geometry)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

ma_county_shape_2 <- ma_county_shape_2 |> 
    rename("GEOID_COUNTY" = "GEOID")


```

```{r}
#| echo: false
#| include: false
#| eval: false



saveRDS(ma_county_shape_2, "ma_county_shape_2.rds")

```

```{r}
#| include: false



ma_county_shape_2 <- readRDS("ma_county_shape_2.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_county_shape_2 |> glimpse()



```



```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(census_county_urban, "census_county_urban.rds")


```

```{r}
#| include: false

census_county_urban <- readRDS("census_county_urban.rds")


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"


census_county_urban |> glimpse()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


census_county_urban |> 
    count(STATE)


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_county_urban <- census_county_urban |> 
    filter(STATE_NAME == "Massachusetts")


```


```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_county_urban , "ma_county_urban.rds")


```

```{r}
#| include: false


ma_county_urban  <- readRDS("ma_county_urban.rds")

```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_county_urban |> summary()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



print(ma_county_urban |> select(COUNTY_NAME), n=14 )


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_county_urban_2 <- ma_county_urban |> 
    filter(COUNTY_NAME != "Dukes") |> 
    filter(COUNTY_NAME != "Nantucket")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_county_urban_2 <- ma_county_urban_2 |> 
    select(COUNTY, POP_COU, HOU_COU, `ALAND_Mi²_COU`, POP_URB, POPPCT_URB, HOU_URB, HOUPCT_URB, `ALAND_Mi²_URB`, ALAND_PCT_URB, POP_RUR, POPPCT_RUR, HOU_URB, HOUPCT_RUR, `ALAND_Mi²_RUR`, ALAND_PCT_RUR)



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_county_urban_2 <- ma_county_urban_2 |> 
    rename("ALAND_MI_COU" = "ALAND_Mi²_COU", "ALAND_MI_URB" = "ALAND_Mi²_URB", "ALAND_MI_RUR" = "ALAND_Mi²_RUR", "COUNTYFP" = "COUNTY")

```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_county_urban_2, "ma_county_urban_2.rds")



```

```{r}
#| include: false


ma_county_urban_2 <- readRDS("ma_county_urban_2.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_county_urban_2 |> glimpse()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



print(ma_county_urban_2 |> select(COUNTYFP), n=14)




```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_county_urban_3 <- ma_county_urban_2 |> 
    mutate(county_pop_density = round(POP_COU / ALAND_MI_COU,0), county_house_density = round(HOU_COU / ALAND_MI_COU,0))


```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_county_urban_3, "ma_county_urban_3.rds")


```

```{r}
#| include: false


ma_county_urban_3 <- readRDS("ma_county_urban_3.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


print(ma_county_urban_3 |> select(county_house_density, county_pop_density),n=14)



```



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


county_geo <- ma_county_shape_2 |> 
    full_join(ma_county_urban_3, by="COUNTYFP")



```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(county_geo , "county_geo.rds")


```

```{r}
#| include: false



county_geo  <- readRDS("county_geo.rds")



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



county_geo |> glimpse()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


county_geo_2 <- county_geo |> 
    select(COUNTYFP, NAME, POP_COU, HOU_COU, POPPCT_URB, HOU_URB, county_pop_density) |> 
    st_drop_geometry()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

county_geo_2 <- county_geo_2 |> 
    rename("County" = "NAME", "Cou_POPPCT_URB" = "POPPCT_URB", "COU_HOU_URB" = "HOU_URB")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



county_geo_2 |> glimpse()

```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(county_geo_2, "county_geo_2.rds")

```

```{r}
#| include: false


county_geo_2 <- readRDS("county_geo_2.rds")


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_county_urban |> 
    select(COUNTY_NAME, POPDEN_COU, POPPCT_URB, HOUDEN_COU, HOUDEN_URB,HOUDEN_RUR, HOUPCT_URB)


```

## Tracts

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tracts <- tracts("MA", year = 2024)


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tracts, "ma_tracts.rds")


```

```{r}
#| include: false

ma_tracts <- readRDS("ma_tracts.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tracts


```


Define the census tract variables we want to retrieve from the acs. 

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


variables_to_get <- c(
    total_tract_housing_units = "B25001_001",
    total_vehicles_tract = "B25046_001",
    total_households_tract = "B25044_001",
    total_population_tract = "B01003_001"


)


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data <- get_acs(
    geography = "tract",
    variables = variables_to_get,
    state = "MA",
    geometry = TRUE,
    output = "wide",
    year = 2021
)



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data |> 
    glimpse()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data |> summary()


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_geometry <- ma_tract_data |> 
    select(GEOID, geometry)


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tract_geometry, "ma_tract_geometry.rds")

```

```{r}
#| include: false


ma_tract_geometry <- readRDS("ma_tract_geometry.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_tract_cls <- ma_tract_geometry |> 
    st_transform(5070) 


```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_tract_cls, "ma_tract_cls.rds")


```

```{r}
#| include: false


ma_tract_cls <- readRDS("ma_tract_cls.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_data |> 
    select(NAME) |> 
    head(10)



```



The output of the NAME column shows it contains Census Tract, Coumty, and State. Will will use the separate function to split the NAME column into three new columns.



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

ma_tract_data_2 <- ma_tract_data |> 
    separate(NAME, into=c("census_tract", "County", "state"), sep=", ") |> 
    st_drop_geometry()



```

Remove State

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_2 <- ma_tract_data_2 |> 
    select(-state)


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tract_data_2 , "ma_tract_data_2.rds")

```


```{r}
#| include: false

ma_tract_data_2  <- readRDS("ma_tract_data_2.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_3 <- ma_tract_geometry |> 
    full_join(ma_tract_data_2, by = "GEOID")


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tract_data_3 , "ma_tract_data_3.rds")

```

```{r}
#| include: false


ma_tract_data_3  <- readRDS("ma_tract_data_3.rds")


```





```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data_3 |> summary()




```


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_4 <- ma_tract_data_3 |> 
    filter(GEOID != "25001990000") |> 
    filter(GEOID != "25023990003")



```


```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data_4 |> 
    filter(GEOID == "25001990000" | GEOID == "25023990003") |> 
    head()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"

ma_tract_data_4 |> 
    filter(GEOID == "25001990000") |> 
    head()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_data_3 |> 
    select(GEOID, total_vehicles_tractE) |> 
    filter(is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summary()



```


```{r}
#| code-fold: true
#| code-summary: "Show Code"




ma_tract_data_3 |> 
    select(GEOID, census_tract,County, total_vehicles_tractE) |> 
    filter(is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    head(72)

```


```{r}
#| code-fold: true
#| code-summary: "Show Code"





ma_tract_data_3 |> 
    filter(County == "Suffolk County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))




```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Suffolk County" & is.na(total_vehicles_tractE), 1310,total_vehicles_tractE ))




```



```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data_3 |> 
    filter(County == "Bristol County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Bristol County" & is.na(total_vehicles_tractE), 2785,total_vehicles_tractE ))



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"




ma_tract_data_3 |> 
    filter(County == "Hampshire County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Hampshire County" & is.na(total_vehicles_tractE), 3340,total_vehicles_tractE ))


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data_3 |> 
    filter(County == "Essex County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Essex County" & is.na(total_vehicles_tractE), 2989,total_vehicles_tractE ))


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data_3 |> 
    filter(County == "Middlesex County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Middlesex County" & is.na(total_vehicles_tractE), 2828,total_vehicles_tractE))
```


```{r}
#| code-fold: true
#| code-summary: "Show Code"




ma_tract_data_3 |> 
    filter(County == "Plymouth County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))

```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Plymouth County" & is.na(total_vehicles_tractE), 3471,total_vehicles_tractE))



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"




ma_tract_data_3 |> 
    filter(County == "Worcester County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Worcester County" & is.na(total_vehicles_tractE), 2898,total_vehicles_tractE))



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data_3 |> 
    filter(County == "Barnstable County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))

```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Barnstable County" & is.na(total_vehicles_tractE), 3166,total_vehicles_tractE))


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_data_3 |> 
    filter(County == "Hampden County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Hampden County" & is.na(total_vehicles_tractE), 2736,total_vehicles_tractE))


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data_3 |> 
    filter(County == "Berkshire County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Berkshire County" & is.na(total_vehicles_tractE), 2248,total_vehicles_tractE))


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_data_3 |> 
    filter(County == "Norfolk County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```





```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Norfolk County" & is.na(total_vehicles_tractE), 3080,total_vehicles_tractE))



```



```{r}
#| code-fold: true
#| code-summary: "Show Code"






ma_tract_data_3 |> 
    select(GEOID, census_tract,County, total_vehicles_tractE) |> 
    filter(is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    head(72)


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tract_data_4, "ma_tract_data_4.rds")

```

```{r}
#| include: false


ma_tract_data_4 <- readRDS("ma_tract_data_4.rds")



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_data_4 |> 
    summary()


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_5 <- ma_tract_data_4 |> 
    filter(total_population_tractE != 0)


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_5 <- ma_tract_data_5 |> 
    filter(County != "Nantucket County") |> 
    filter(County != "Dukes County")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_5 <- ma_tract_data_5 |> 
    filter(total_households_tractE != 0)



```



```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tract_data_5, "ma_tract_data_5.rds")


```

```{r}
#| include: false


ma_tract_data_5 <- readRDS("ma_tract_data_5.rds")



```

```{r}

ma_tract_data_5 |> count(County)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"

ma_tract_data_5 |> 
    filter(is.na(total_vehicles_tractE)) |> 
    count(County)


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_data_5 |> 
    summary()


```




### Feature Engineering



From our retrieved census tract variables, we will create new variables for housing density of census tract, population density of census tract, and average cars per houshold of census tract. 







```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_6 <- ma_tract_data_5 |> 
    mutate(
        tract_house_density_sqmi = round(as.numeric(set_units(total_tract_housing_unitsE/ st_area(geometry),"1/mi2")),2),
        tract_pop_density_sqmi = round(as.numeric(set_units(total_population_tractE/ st_area(geometry),"1/mi2")),2),
        avg_cars_per_household = round(total_vehicles_tractE / total_households_tractE, 2)  
    #select(!ends_with("M"))
    #rename_with(.fn = ~str_remove(.x, "E$"))


    )



```



```{r}
#| code-fold: true
#| code-summary: "Show Code"




print(ma_tract_data_5 |> 
    filter(total_households_tractE == 0), n=70)


```






```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_6 <- ma_tract_data_6 |> 
    select(-total_tract_housing_unitsM, -total_vehicles_tractM, -total_households_tractM, -total_population_tractM)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_6 <- ma_tract_data_6 |> 
    select(-County)


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tract_data_6 , "ma_tract_data_6.rds")


```

```{r}
#| include: false


ma_tract_data_6 <- readRDS("ma_tract_data_6.rds")


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"





ma_tract_data_6|> glimpse()



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data_6 |> summary()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_7 <- ma_tract_data_6 |> 
    select(-census_tract, -County) |> 
    st_drop_geometry()




```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tract_data_7, "ma_tract_data_7.rds")


```

```{r}
#| include: false

ma_tract_data_7 <- readRDS("ma_tract_data_7.rds")


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data_7|> 
    glimpse()


```






## Joining


# Exploratory Data Analysis


## Features


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_accidents_16_21_6 |> 
    ggplot(aes(x=accident_duration, y=Temperature.F.)) +
    geom_point()



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"




ma_accidents_16_21_6 |> 
    ggplot(aes(x=log(accident_duration), y=Temperature.F.)) +
    geom_point()


```

## Maps

### County


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


county_map <- ggplot()+
    geom_sf(data=ma_county, fill=NA)+
    ggtitle("Counties of Massachusetts") +
    theme(legend.position = "none",plot.title = element_text(color="blue", size=14, face="bold.italic", hjust=0.5)) +
    theme_bw()


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_county_geo_4326 <- county_geo |> 
    st_transform(crs=4326)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_county_stcrs <- st_crs(ma_county_geo_4326)


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_county_stcrs$input



```









```{r}
#| code-fold: true
#| code-summary: "Show Code"



bins<-c(0,71029,129026,162308,228996,465825,530819,579200,725981,797936,809829,862111,Inf)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"




pal<-colorBin(palette="RdYlBu",
            domain=ma_county_geo_4326$POP_COU,bins=bins)
```

```{r}
#| code-fold: true
#| code-summary: "Show Code"




ma_county_formated <- ma_county_geo_4326 |> 
    mutate(county_pop_density = format(county_pop_density, big.mark = ",", big.interval = 3, digits=12) ) |> 
    mutate(POPPCT_URB = round(POPPCT_URB * 100,2))


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"





mytext <- paste(
  "County: ", ma_county_formated$NAME,"<br/>",
  "Population Density: ", ma_county_formated$county_pop_density, "<br/>",
  "County Population Percent Urban: ", ma_county_formated$POPPCT_URB,"<br/>", 
    sep="") %>%
  lapply(htmltools::HTML)


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"



map_county_pop_density_plt<- ma_county_geo_4326 |> leaflet(options = leafletOptions(minZoom = 9.4, maxZoom = 18)) |> 
  addProviderTiles(providers$Esri.WorldStreetMap) |> 
  setView(lng = -071.9402820, lat = 42.3116934, zoom = 9.4) |> 
  addPolygons(
 fillColor=~pal(POP_COU),
 weight=2,
 opacity=0.3,
 color= "white",
 label=~mytext,
 dashArray = "3",
 fillOpacity=0.3,
 highlightOptions = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.5,
    bringToFront = FALSE)) 
    
  




```



```{r}
#| code-fold: true
#| code-summary: "Show Code"


county_map_pop_density<-map_county_pop_density_plt %>% 
  addLegend(pal = pal, values = ~county_pop_density, opacity = 0.6, title = "Massachusetts County Population Density",
  position = "bottomright")


```

```{r}

map_county_pop_density_plt



```



### Census Tracts


```{r}
#| code-fold: true
#| code-summary: "Show Code"




ma_tracts_plot <- ggplot(ma_tracts_2) +
    geom_sf()+
    ggtitle("Census Tracts of Massachusetts") +
    theme(legend.position = "none",plot.title = element_text(color="blue", size=14, face="bold.italic", hjust=0.5)) +
    theme_void()



```

```{r}

ma_tracts_plot


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

tract_accident_count <- ma_accidents_16_21_8 |> 
    count(GEOID, name = "tract_accident_count")


```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(tract_accident_count, "tract_accident_count.rds")

```

```{r}
#| include: false


tract_accident_count <- readRDS("tract_accident_count.rds")



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"





tract_accident_count |> 
    slice_max(tract_accident_count, n=20) 

```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_8 |> 
    filter(GEOID == "25017334100")



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_accidents_16_21_8 |> 
    filter(accident_duration >=30 & accident_duration <=60)

```


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_accidents_16_21_10 <- ma_accidents_16_21_8 |> 
    full_join(tract_accident_count, by = "GEOID")



```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_accidents_16_21_10, "ma_accidents_16_21_10.rds")


```

```{r}
#| include: false



ma_accidents_16_21_10 <- readRDS("ma_accidents_16_21_10.rds")


```

 ```{r}
 #| code-fold: true
#| code-summary: "Show Code"

 
ma_accidents_16_21_10 |> 
 count(GEOID) |> 
 arrange(desc(n))


 ```

 ```{r}
 



 ```

 ```{r}
 #| code-fold: true
#| code-summary: "Show Code"

 

ma_accidents_16_21_10 |> 
    summary()



 ```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_tract_accident <- ma_tract_data_6 |> 
    full_join(ma_accidents_16_21_10, by = "GEOID")


```

```{r}
#| echo: false
#| include: false
#| eval: false



saveRDS(ma_tract_accident, "ma_tract_accident.rds")


```

```{r}
#| include: false


ma_tract_accident <- readRDS("ma_tract_accident.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"




ma_tract_accident |> 
    summary()



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_accident |> 
    filter(is.na(total_households_tractE)) |> 
    summary()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"





ma_tract_accident |> 
    filter(is.na(total_households_tractE)) |> 
    count(GEOID)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_accident |> 
    filter(is.na(total_households_tractE)) |> 
    head(1000)



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_accident_2 <- ma_tract_accident |> 
    mutate(total_households_tractE = replace_na(total_households_tractE, 0), 
           total_population_tractE = replace_na(total_population_tractE, 0),
           tract_house_density_sqmi = replace_na(tract_house_density_sqmi,0),
           tract_pop_density_sqmi = replace_na(tract_pop_density_sqmi,0),
           total_vehicles_tractE = replace_na(total_vehicles_tractE,0),
           total_tract_housing_unitsE = replace_na(total_tract_housing_unitsE,0) )


```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_tract_accident_2, "ma_tract_accident_2.rds")

```

```{r}
#| include: false


ma_tract_accident_2 <- readRDS("ma_tract_accident_2.rds")

```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_accident_2 |> 
    summary()

```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false




ma_tract_accident_3 <- ma_tract_accident_2 |> 
    filter(!is.na(day_of_week))
```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_tract_accident_3, "ma_tract_accident_3.rds")

```

```{r}
#| include: false

ma_tract_accident_3 <- readRDS("ma_tract_accident_3.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

ma_tract_accident_3 |> 
    summary()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_cls <- st_crs(ma_tract_accident_3)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

ma_tract_cls


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

ma_census4326 <- ma_tract_accident_3 |> 
    st_transform(crs=4326)


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_census4326, "ma_census4326.rds")


```

```{r}
#| include: false



ma_census4326 <- readRDS("ma_census4326.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_cls_2 <- st_crs(ma_census4326)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_cls_2$input


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


bins<-c(0,500,2000,3000,4000,5000,6500,8500,9500,10500,Inf)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



pal<-colorBin(palette="RdYlBu",
            domain=ma_census4326$total_population_tractE,bins=bins)



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

       
       
ma_tract_formated <- ma_tract_accident_3 |> 
    mutate(total_tract_housing_unitsE = format(total_tract_housing_unitsE, big.mark = ",", big.interval = 3, digits=12) ) |> 
    mutate(total_vehicles_tractE = format(total_vehicles_tractE, big.mark = ",", big.interval = 3, digits=12) ) 
   


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



mytext <- paste(
  "County: ", ma_tract_formated$County,"<br/>",
  "Census Tract: ", ma_tract_formated$GEOID, "<br/>",
  "Total Vehicles Census Tract: ", ma_tract_formated$total_vehicles_tractE, "<br/>",
  "Total Housing Census Tract: ", ma_tract_formated$total_tract_housing_unitsE,"<br/>", 
  "Census Tract Accident County: ", ma_tract_formated$tract_accident_count,"<br/>",
    sep="") %>%
  lapply(htmltools::HTML)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"




map_tract_population_plt<- ma_census4326 |> leaflet(options = leafletOptions(minZoom = 9.4, maxZoom = 18)) |> 
  addProviderTiles(providers$Esri.WorldStreetMap) |> 
  setView(lng = -071.9402820, lat = 42.3116934, zoom = 9.4) |> 
  addPolygons(
 fillColor=~pal(total_population_tractE),
 weight=2,
 opacity=0.8,
 color= "white",
 label=~mytext,
 dashArray = "3",
 fillOpacity=0.3,
 highlightOptions = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.5,
    bringToFront = FALSE)) 


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"




tract_map_population<-map_tract_population_plt %>% 
  addLegend(pal = pal, values = ~total_population_tractE, opacity = 0.6, title = "Massachusetts Census Tract Population",
  position = "bottomright")

```

```{r}


tract_map_population
  

```


```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_census4326 |> 
    filter(GEOID == "25017368102")


```

```{r}






```



```{r}






```

```{r}





```

```{r}





```












# Preprocessing


```{r}



```


#rows / columns 
10:1 rule 
You should have at least 10 rows of data for every 1 column (feature) in your dataset. 

`

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


census_tract_df <- ma_accidents_16_21_8 |> 
    left_join(ma_tract_data_7, by = "GEOID")

```

```{r}
#| echo: false
#| include: false
#| eval: false



saveRDS(census_tract_df, "census_tract_df.rds")


```


```{r}
#| include: false


census_tract_df <- readRDS("census_tract_df.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



census_tract_df |> 
   summary()


```

```{r}

census_tract_df  |> filter(is.na(total_population_tractE))  |> count(GEOID)


```

```{r}

census_tract_df |> filter(total_tract_housing_unitsE == 0) |> head(10)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


census_tract_county_df <- census_tract_df |> 
    left_join(county_geo_2, by = "County")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


census_tract_df_2 <- census_tract_df |> 
    select(-ID, -Severity, -City, -County, -Year, -LATITUDE, -LONGITUDE, -census_code, -tract_house_density_sqmi, -Stop,-Amenity, -Crossing,-Junction ,-Station,-Traffic_Signal,-Sunrise_Sunset,-Civil_Twilight, -day_of_week, -total_tract_housing_unitsE, -total_vehicles_tractE, -total_population_tractE, -avg_cars_per_household, -total_households_tractE, -tract_pop_density_sqmi)


```

```{r}
#| echo: false
#| include: false
#| eval: false



saveRDS(census_tract_df_2, "census_tract_df_2.rds")


```

```{r}
#| include: false


census_tract_df_2 <- readRDS("census_tract_df_2.rds")



```

```{r}


census_tract_df_3 <- census_tract_df |> 
    select(-ID, -Severity, -City, -Year, -LATITUDE, -LONGITUDE, -census_code, -tract_house_density_sqmi, -Stop,-Amenity, -Crossing,-Junction ,-Station,-Traffic_Signal,-Sunrise_Sunset,-Civil_Twilight, -day_of_week, -total_tract_housing_unitsE, -total_vehicles_tractE, -avg_cars_per_household, -total_households_tractE, -tract_pop_density_sqmi, -total_population_tractE)



```




```{r}


census_tract_df_3 |> summary()

```



```{r}


census_tract_df_2 |> 
    glimpse()


```

```{r}
#| echo: false
#| include: false
#| eval: false



census_tract_df |> select(GEOID, County, total_households_tractE, total_population_tractE, total_vehicles_tractE) |> filter(is.na(total_households_tractE)) |> head()

```



```{r}
#| echo: false
#| include: false
#| eval: false


census_tract_df_4 <- census_tract_df_3 |> 
    filter(!is.na(total_population_tractE))


```

```{r}
#| echo: false
#| include: false
#| eval: false

census_tract_df_2 <- census_tract_df_2 |> 
    filter(total_tract_housing_unitsE != 0)


```


```{r}
#| echo: false
#| include: false
#| eval: false


census_tract_county_df_2 <- census_tract_county_df |> 
    select(-ID, -Severity, -City, -County, -Year, -LATITUDE, -LONGITUDE, -census_code, -tract_house_density_sqmi, -Stop,-Amenity, -Crossing,-Junction ,-Station,-Traffic_Signal,-Sunrise_Sunset,-Civil_Twilight, -day_of_week, -total_tract_housing_unitsE, -total_vehicles_tractE, -total_population_tractE, -avg_cars_per_household, -total_households_tractE, -tract_pop_density_sqmi, -County, 
    -POP_COU, -HOU_COU, -COU_HOU_URB)



```

```{r}
#| echo: false
#| include: false
#| eval: false


census_tract_county_df_2 |> summary()


```



```{r}
#| echo: false
#| include: false
#| eval: false



census_tract_df_2 |> 
    ggplot(aes(temperature)) +
    geom_histogram(bins=25)


```

```{r}
#| echo: false
#| include: false
#| eval: false


census_tract_df_2 |> 
    ggplot(aes(log(wind_speed))) +
    geom_histogram(bins=25)


```


```{r}
#| echo: false
#| include: false
#| eval: false


census_tract_df_2 |> 
    filter(accident_duration < 5000) |> 
    ggplot(aes(accident_duration,log(temperature))) +
    geom_point()


```

```{r}
#| echo: false
#| include: false
#| eval: false



census_tract_df_2 |> 
    mutate(GEOID = as.factor(GEOID)) |> 
    count(GEOID) |>
    arrange(desc(n)) 



```





## Train/Test Split

```{r}
#| code-fold: true
#| code-summary: "Show Code"


acci_census_df <- initial_split(census_tract_df_2, prop = 0.80)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"

print(acci_census_df)



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


acci_census_df_2 <- initial_split(census_tract_df_3, prop = 0.80)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


acci_count_tract_df <- initial_split(census_tract_county_df_2, prop = 0.80)


```






```{r}
#| code-fold: true
#| code-summary: "Show Code"



train_census_data <- training(acci_census_df)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"

dim(train_census_data)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


test_census_data <- testing(acci_census_df)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"

dim(test_census_data)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"

train_census_data_2 <- training(acci_census_df_2)


```

```{r}


test_census_data_2 <- testing(acci_census_df_2)


```

```{r}


train_tract_cou <- training(acci_count_tract_df)

```

```{r}



test_tract_coun <- testing(acci_count_tract_df)


```





```{r}
#| code-fold: true
#| code-summary: "Show Code"



census_rec <- recipe(accident_duration ~ ., data = train_census_data) |> 
    step_center(all_numeric_predictors()) |> 
    step_scale(all_numeric_predictors())


```

```{r}


census_rec_2<- recipe(accident_duration ~ ., data = train_census_data_2) |> 
    step_center(all_numeric_predictors()) |> 
    step_scale(all_numeric_predictors())


```

```{r}



census_cou_rec <- recipe(accident_duration ~ ., data = train_tract_cou) |> 
    step_center(all_numeric_predictors()) |> 
    step_scale(all_numeric_predictors())

```




```{r}
#| code-fold: true
#| code-summary: "Show Code"


train_census_rec <- prep(census_rec, training= train_census_data)
scaled_census_train <- bake(train_census_rec, new_data = train_census_data)


```

```{r}

train_census_rec_2 <- prep(census_rec_2, training= train_census_data_2)
scaled_census_train_2 <- bake(train_census_rec_2, new_data = train_census_data_2)



```


```{r}


train_tract_cou_rec <- prep(census_cou_rec, training=train_tract_cou)
scaled_tract_cou_train <- bake(train_tract_cou_rec, new_data = train_tract_cou)


```

# Models

## Linear Regression

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



lr <- lm(log(accident_duration) ~ . -GEOID, data = train_census_data)


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(lr, "lr.rds")

```

```{r}
#| include: false


lr <- readRDS("lr.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



summary(lr)



```


```{r}


lr_resid <- data.frame(residuals = residuals(lr))

```



```{r}


lr_qq <- ggplot(lr_resid, aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(
    title = "LR Model Normal Q-Q Plot of Residuals",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  )
```

```{r}


lr_qq

```

```{r}



lr_resid_data <- data.frame(
    Fitted = fitted(lr),
    Residuals = resid(lr)
)


```

```{r}


lr_residual_plot <- ggplot(lr_resid_data, aes(x = Fitted, y = Residuals)) +
    geom_point(color = "steelblue", size =2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    labs(title = "lr Model Residual Plot", x = "Fitted Values", y = "Residuals" ) +
    theme_minimal()

```

```{r}


lr_residual_plot

```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


lr_2 <- lm(log(accident_duration) ~ . -GEOID, data = train_census_data_2)


```



```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(lr_2, "lr_2.rds")



```

```{r}
#| include: false


lr_2 <- readRDS("lr_2.rds")

```

```{r}

summary(lr_2)

```

```{r}


lr_2_resid <- data.frame(residuals = residuals(lr_2))


```

```{r}

lr_2_qq <- ggplot(lr_2_resid, aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(
    title = "LR_2 Model Normal Q-Q Plot of Residuals",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  )

```

```{r}

lr_2_qq

```

```{r}

lr_2_resid_data <- data.frame(
    Fitted = fitted(lr_2),
    Residuals = resid(lr_2)
)

```

```{r}

lr_2_residual_plot <- ggplot(lr_2_resid_data, aes(x = Fitted, y = Residuals)) +
    geom_point(color = "steelblue", size =2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    labs(title = "lr Model Residual Plot", x = "Fitted Values", y = "Residuals" ) +
    theme_minimal()

```

```{r}

lr_2_residual_plot 

```






## Fixed Effects



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


model_lme_1 <- lmer(log(accident_duration) ~ . + temperature*Month -GEOID -Wind_Direction -Weather_Condition + (temperature | GEOID), data = scaled_census_train)



```

```{r}




```

```{r}





```

```{r}
#| code-fold: true
#| code-summary: "Show Code"




summary(model_lme_1)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



performance::check_singularity(model_lme_1)



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

model_lme_2 <- lmer(log(accident_duration) ~ . -GEOID  + (1 | GEOID), data = scaled_census_train)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



summary(model_lme_2)

```


```{r}


icc(model_lme_2)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"




performance::check_singularity(model_lme_2)


```

The result of TRUE from the check_singularity function informs us that the  lme4 model has some variance components that equal zero and thus random effect variances can not be computed. 

```{r}
#| code-fold: true
#| code-summary: "Show Code"



performance::check_heteroscedasticity(model_lme_2)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



performance::check_heteroskedasticity(model_lme_2)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



summary(model_lme_1)


```





```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


stan_model_1 <- stan_lmer(
    log(accident_duration) ~ . -GEOID + (temperature | GEOID),
  data = scaled_census_train,
  chains = 4,
  iter = 8000,
  cores = 30, 
  seed = 42
)


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



brms_model <- brm(
  log(accident_duration) ~ . -GEOID + (temperature | GEOID),
  data = scaled_census_train,
  chains = 4,
  iter = 12000,
  family = student(),
  cores = 40, 
  seed = 42
)



```


```{r}




```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(brms_model, "brms_model.rds")


```


```{r}
#| include: false


brms_model <- readRDS("brms_model.rds")


```





```{r}
#| code-fold: true
#| code-summary: "Show Code"


summary(brms_model)



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"




performance::check_singularity(brms_model)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

performance::check_collinearity(brms_model)


```



```{r}


icc(brms_model, nrep=50)


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


brms_model_2 <- brm(
  log(accident_duration) ~ . -GEOID -Wind_Direction -Weather_Condition + (temperature | GEOID),
  data = scaled_census_train,
  chains = 4,
  iter = 8000,
  family = student(),
  cores = 40, 
  seed = 42
)



```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(brms_model_2, "brms_model_2.rds")


```

```{r}
#| include: false


brms_model_2 <- readRDS("brms_model_2.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



summary(brms_model_2)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



performance::check_collinearity(brms_model_2)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



brms_model_3 <- brm(
  log(accident_duration) ~ . -GEOID -Wind_Direction -Weather_Condition + temperature * Month + (temperature | GEOID),
  data = scaled_census_train,
  chains = 4,
  iter = 8000,
  family = student(),
  cores = 16, 
  seed = 42
)


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(brms_model_3, "brms_model_3.rds")


```

```{r}
#| include: false


brms_model_3 <- readRDS("brms_model_3.rds")



```

```{r}


summary(brms_model_3)


```

```{r}
#| echo: false
#| include: false
#| eval: false



dp <- default_prior(
    log(accident_duration) ~ temperature + day_type + road_category + hour_day_sections + (temperature|GEOID),
    data = scaled_census_train,
    family=student()
)

```

```{r}
#| echo: false
#| include: false
#| eval: false

dp |> 
    filter(coef == "" | is.na(coef)) |> 
    select(class, group, resp, dpar, prior)



```

```{r}
#| echo: false
#| include: false
#| eval: false


prior_b <- prior(normal(0.3, 0.2), class = "b")

```

```{r}
#| echo: false
#| include: false
#| eval: false


prior_sim_model <- brm(
    formula = log(accident_duration) ~ temperature + day_type + road_category + hour_day_sections + (temperature|GEOID),
               data = scaled_census_train,
               family=student(), 
               prior = prior_b,
               sample_prior = "only",
               chains = 2,
               iter = 1000)


```

```{r}
#| echo: false
#| include: false
#| eval: false


pp_check(prior_sim_model, type = "dens_overlay", ndraws = 50)


```

```{r}
#| echo: false
#| include: false
#| eval: false



tamed_priors <- c(
  # Baseline log-price around 13.5 (~$700k) with room to range from ~10 to 17
  prior(normal(3.81, 1.0), class = "Intercept"),
  
  # Slopes on standardized predictors: 
  # A normal(0, 0.5) prior says a 1-SD change rarely alters price by more than +/- 100%
  prior(normal(0, 0.5), class = "b"),
  
  # Group-level variability across census tracts
  prior(exponential(2), class = "sd"),
  
  # Student-t residual variance and degrees of freedom
  prior(exponential(2), class = "sigma"),
  prior(gamma(2, 0.1), class = "nu")
)




```

```{r}
#| echo: false
#| include: false
#| eval: false



prior_sim_model_2 <- brm(
    formula = log(accident_duration) ~ temperature + day_type + road_category + hour_day_sections + (temperature|GEOID),
               data = scaled_census_train,
               family=student(), 
               prior = tamed_priors,
               sample_prior = "only",
               chains = 2,
               iter = 1000)



```

```{r}
#| echo: false
#| include: false
#| eval: false


pp_check(prior_sim_model_2, type = "dens_overlay", ndraws = 50)



```

```{r}
#| echo: false
#| include: false
#| eval: false


brms_model_3 <- brm(
  log(accident_duration) ~ . -GEOID -Wind_Direction -Weather_Condition + temperature * Month + hour_day_sections * day_type + (temperature | GEOID),
  data = scaled_census_train,
  chains = 4,
  iter = 8000,
  family = student(),
  cores = 16, 
  seed = 42
)


```

```{r}
#| echo: false
#| include: false
#| eval: false


summary(brms_model_3)


```

```{r}




```

```{r}



summary(brms_model_4)


```

```{r}



brms_model_cat_plot_top_60 <- brms_model_2 %>%
  spread_draws(r_GEOID[GEOID, Intercept]) %>%
  median_qi(.width = 0.95) %>%
  slice_max(order_by = abs(r_GEOID), n = 60) %>%
  ggplot(aes(x = r_GEOID, y = reorder(GEOID, r_GEOID), xmin = .lower, xmax = .upper)) +
  geom_pointinterval(color = "#1f78b4") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title = "Top 60 Most Extreme Census Tract Effects (brms_model)",
    x = "Log Accident Duration Adjustment",
    y = "Census Tract (GEOID)"
  )


```



```{r}


brms_model_cat_plot_top_60_2 <- brms_model_2 %>%
  spread_draws(r_GEOID[GEOID, Intercept]) %>%
  median_qi(.width = 0.95) %>%
  slice_max(order_by = abs(r_GEOID), n = 100) %>%
  ggplot(aes(x = r_GEOID, y = reorder(GEOID, r_GEOID), xmin = .lower, xmax = .upper)) +
  geom_pointinterval(color = "#1f78b4") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title = "Top 60 Most Extreme Census Tract Effects (brms_model)",
    x = "Log Accident Duration Adjustment",
    y = "Census Tract (GEOID)"
  )



```

```{r}



brms_model_cat_plot_top_60


```

```{r}


brms_model_cat_plot_top_60_2


```

```{r}


ma_census4326 |> 
    filter(GEOID %in% c("25021418004",  "25001011802","25005644200", "25017321500", "25021416101", "25021421200","25023507101", "25017312000", "25005630202","25017315401", "25021421100", "25017373902",
                         "25025130402", "25017358700", "25009211201","25025101001", "25025100800", "25025140300","25023520100", "25009208401", "25025981100", "25023502101")) |> 
                            count(County) |> 
                            arrange(desc(n)) |> 
                            st_drop_geometry()


```

```{r}


ma_census4326 |> 
    filter(GEOID %in% c("25021418004",  "25001011802","25005644200", "25017321500", "25021416101", "25021421200","25023507101", "25017312000", "25005630202","25017315401", "25021421100", "25017373902",
                         "25025130402", "25017358700", "25009211201","25025101001", "25025100800", "25025140300","25023520100", "25009208401", "25025981100", "25023502101")) |> 
                            count(County,GEOID) |> 
                            arrange(desc(n)) |> 
                            st_drop_geometry()



```

```{r}


ma_census4326 |> 
    filter(GEOID %in% c( "25017320105",  "25021419700", "25021419800", "25009254401", "25025010600", "25027744103", "25017368400","25013802101", "25025170301", "25025170400", "25017350105", "25013810100",
                        "25017320106", "25021411200", "25025081001", "25027732901","25017358600","25017374800", "25005614102", "25017337300", "25027710200", "25021403100", "25017342501", "25023501204",
                          "25017383700", "25017384003", "25017374200", "25025050901", "25021410100","25021414100")) |> 
                            count(County) |> 
                            arrange(desc(n)) |> 
                            st_drop_geometry()


```

```{r}


ma_census4326 |> 
    filter(GEOID %in% c( "25017320105",  "25021419700", "25021419800", "25009254401", "25025010600", "25027744103", "25017368400","25013802101", "25025170301", "25025170400", "25017350105", "25013810100",
                        "25017320106", "25021411200", "25025081001", "25027732901","25017358600","25017374800", "25005614102", "25017337300", "25027710200", "25021403100", "25017342501", "25023501204",
                          "25017383700", "25017384003", "25017374200", "25025050901", "25021410100","25021414100")) |> 
                            count(County, GEOID) |> 
                            arrange(desc(n)) |> 
                            st_drop_geometry()

```


```{r}

ma_census_middle <- ma_census4326 |> 
    filter(County == "Middlesex")


```

```{r}


saveRDS(ma_census_middle, "ma_census_middle.rds")


```

```{r}
#| include: false


ma_census_middle <- readRDS("ma_census_middle.rds")


```

```{r}

ma_extreme_middle_higher <- ma_census_middle |> 
    select(GEOID, ID, accident_duration, County, City, hour_day_sections, day_type, Month,total_vehicles_tractE,road_category, LONGITUDE, LATITUDE) |> 
    filter(GEOID %in% c("25017321500","25017358700", "25017315401", "25017373902", "25017312000" )) |> 
        st_drop_geometry()

```

```{r}


saveRDS(ma_extreme_middle_higher, "ma_extreme_middle_higher.rds")


```

```{r}
#| include: false

ma_extreme_middle_higher <- readRDS("ma_extreme_middle_higher.rds")


```

```{r}

bins<-c(0,200,4000,5400,6500,7500,8500,Inf)

```





```{r}

pal<-colorBin(palette="RdYlBu",
            domain=ma_census_middle$total_population_tractE,bins=bins)



```












```{r}


ma_tract_middle_extreme_formatted <- ma_extreme_middle_higher |> 
    mutate(accident_duration = format(accident_duration, big.mark = ",", big.interval = 3, digits=12) ) 
     


```

```{r}

saveRDS(ma_tract_middle_extreme_formatted, "ma_tract_middle_extreme_formatted.rds")

```

```{r}

ma_tract_middle_extreme_formatted <- readRDS("ma_tract_middle_extreme_formatted.rds")


```

```{r}



mytext <- paste(
  "Census Tract: ", ma_census_middle$GEOID, "<br/>",
  "Total Vehicles Census Tract: ", ma_census_middle$total_vehicles_tractE, "<br/>",
  "Total Housing Census Tract: ", ma_census_middle$total_tract_housing_unitsE,"<br/>", 
  "Census Tract Accident County: ", ma_census_middle$tract_accident_count,"<br/>",
    sep="") %>%
  lapply(htmltools::HTML)


```

```{r}




map_tract_extreme_middle_plt <- ma_census_middle |> leaflet(options = leafletOptions(minZoom = 11, maxZoom = 18)) |> 
  addProviderTiles(providers$Esri.WorldStreetMap) |> 
  setView(lng = -071.23928, lat = 42.38980, zoom = 11.5) |> 
  addPolygons(
 fillColor=~pal(total_population_tractE),
 weight=2,
 opacity=0.8,
 color= "white",
 label=~mytext,
 dashArray = "3",
 fillOpacity=0.05,
 highlightOptions = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.2,
    bringToFront = FALSE)) |> 
  addMarkers(
    data = ma_tract_middle_extreme_formatted,
    lng = ~LONGITUDE,
    lat = ~LATITUDE,
    popup = ~paste("Accident ID: ", ID,
                   "<br>Accident Duration: ", accident_duration, 
                   "<br>City: ", City, 
                   "<br>Weekday/Weekend: ", day_type,
                   "<br>Time of Day: ", hour_day_sections,
                   "<br>Road Type: ", road_category) )     

```

```{r}


tract_map_extreme_middle<-map_tract_extreme_middle_plt %>% 
  addLegend(pal = pal, values = ~total_population_tractE, opacity = 0.6, title = "Massachusetts Extreme Census Tract Population",
  position = "bottomright")


```

```{r}

tract_map_extreme_middle


```

```{r}


bins_2<-c(0,200,4000,5400,6500,7500,8500,Inf)


```

```{r}


ma_census_norfolk <- ma_census4326 |> 
    filter(County == "Norfolk")


```

```{r}

saveRDS(ma_census_norfolk, "ma_census_norfolk.rds")

```

```{r}
#| include: false


ma_census_norfolk <- readRDS("ma_census_norfolk.rds")


```

```{r}



ma_extreme_norfolk_higher <- ma_census_norfolk |> 
    select(GEOID, ID, accident_duration, County, City, hour_day_sections, day_type, Month,total_vehicles_tractE,road_category, LONGITUDE, LATITUDE) |> 
    filter(GEOID %in% c("25021421200","25021421100","25021416101", "25021418004")) |> 
        st_drop_geometry()


```

```{r}

saveRDS(ma_extreme_norfolk_higher, "ma_extreme_norfolk_higher.rds")

```

```{r}
#| include: false

ma_extreme_norfolk_higher <- readRDS("ma_extreme_norfolk_higher.rds")


```

```{r}

pal_2 <- colorBin(palette="RdYlBu",
            domain=ma_census_norfolk$total_population_tractE,bins=bins_2)


```

```{r}


ma_tract_norfolk_extreme_formatted <- ma_extreme_norfolk_higher |> 
    mutate(accident_duration = format(accident_duration, big.mark = ",", big.interval = 3, digits=12) )


```

```{r}

saveRDS(ma_tract_norfolk_extreme_formatted, "ma_tract_norfolk_extreme_formatted.rds")


```

```{r}
#| include: false


ma_tract_norfolk_extreme_formatted <- readRDS("ma_tract_norfolk_extreme_formatted.rds")


```

```{r}


mytext_2 <- paste(
  "Census Tract: ", ma_census_norfolk$GEOID, "<br/>",
  "Total Vehicles Census Tract: ", ma_census_norfolk$total_vehicles_tractE, "<br/>",
  "Total Housing Census Tract: ", ma_census_norfolk$total_tract_housing_unitsE,"<br/>", 
  "Census Tract Accident County: ", ma_census_norfolk$tract_accident_count,"<br/>",
    sep="") %>%
  lapply(htmltools::HTML)


```

```{r}



map_tract_extreme_norfolk_plt <- ma_census_norfolk |> leaflet(options = leafletOptions(minZoom = 9, maxZoom = 18)) |> 
  addProviderTiles(providers$Esri.WorldStreetMap) |> 
  setView(lng = -71.17127  , lat = 42.17051 , zoom = 11.5) |> 
  addPolygons(
 fillColor=~pal_2(total_population_tractE),
 weight=2,
 opacity=0.5,
 color= "white",
 label=~mytext_2,
 dashArray = "3",
 fillOpacity=0.05,
 highlightOptions = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.2,
    bringToFront = FALSE)) |> 
  addMarkers(
    data = ma_tract_norfolk_extreme_formatted,
    lng = ~LONGITUDE,
    lat = ~LATITUDE,
    popup = ~paste("Accident ID: ", ID,
                   "<br>Accident Duration: ", accident_duration, 
                   "<br>City: ", City, 
                   "<br>Weekday/Weekend: ", day_type,
                   "<br>Time of Day: ", hour_day_sections,
                   "<br>Road Type: ", road_category) ) 


```

```{r}


tract_map_extreme_norfolk<-map_tract_extreme_norfolk_plt %>% 
  addLegend(pal = pal_2, values = ~total_population_tractE, opacity = 0.6, title = "Norfolk Extreme Census Tract Population",
  position = "bottomright")


```

```{r}

tract_map_extreme_norfolk


```

```{r}



print(ma_census_norfolk  |> select(GEOID, County), n=1500)


```


```{r}




brms_extreme_df_2 <- brms_model_2 %>%
  spread_draws(r_GEOID[GEOID, Intercept]) %>%
  median_qi(.width = 0.95) %>%
  slice_max(order_by = abs(r_GEOID), n = 100)


```

```{r}


print(brms_extreme_df_2, n=100)


```


```{r}


brms_model_2 <- brm(
  log(accident_duration) ~ . -GEOID + (temperature | GEOID),
  data = scaled_census_train_2,
  chains = 4,
  iter = 12000,
  family = student(),
  cores = 40, 
  seed = 42
)




```



```{r}



# Extract fixed-effect formula without random effects
ff <- nobars(formula(your_model)
X <- model.matrix(ff, data = your_data)
)
# Check for linear combinations
caret::findLinearCombos(X)


```

```{r}


library(caret)
```

```{r}

# Extract fixed-effect formula without random effects
ff <- nobars(formula(model_lme_1))
X <- model.matrix(ff, data = train_data) 


# Check for linear combinations
caret::findLinearCombos(X)



```

```{r}


summary(model_lme_1)



```

```{r}


icc(model_lme_1, nrep=50)

```


