---
title: "Predicting time of auto accidents"
author: Francis Lauder
date: 2026-09-10
format:
  html:
    toc: true
    toc-depth: 3
    toc_float: true
    number-sections: false
    toc-location: left
    embed-resources: true
---

# Introduction
<br><br>


# Data Dictionary

<br><br>


# Libraries

<br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"

library(dplyr)
library(ggplot2)
library(plotly)
library(RColorBrewer)
library(forcats)
library(tidycensus)

library(sf)

library(jsonlite)

library(tigris)

library(stringr)

library(leaflet)
library(htmlwidgets)
library(htmltools)
library(gt)
library(DT)
library(GGally)
library(lme4)
library(lmerTest)
library(rstanarm)
library(brms)
library(broom)
library(lubridate)
library(tidymodels)
library(performance)






```

<br><br>


# Data Inspection and Cleaning

## US Accident Data


```{r}
#| echo: false
#| include: false
#| eval: false


US_accidents <- read.csv("/Users/francislauder/Documents/Data Sets/US Acciidents/US_Accidents_March23.csv")


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(US_accidents, "US_accidents.rds")

```

```{r}
#| echo: false
#| include: false
#| eval: false


US_accidents <- readRDS("US_accidents.rds")


```



```{r}
#| echo: false
#| include: false
#| eval: false



glimpse(US_accidents)


```



```{r}
#| echo: false
#| include: false
#| eval: false


MA_accidents <- US_accidents |> 
    filter(State == "MA")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


MA_accidents <- MA_accidents %>% 
  mutate(Weather_Timestamp = as.Date(Weather_Timestamp))


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(MA_accidents, "MA_accidents.rds")



```

```{r}
#| echo: false


MA_accidents <- readRDS("MA_accidents.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



options(browser="false", scipen = 999, max.print = 20000)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



MA_accidents |> glimpse()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


summary(MA_accidents)


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"


dim(MA_accidents)


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"



MA_accidents |> 
    count(Severity) |> 
    mutate(perc = round(n / sum(n),3)) |> 
    arrange(desc(perc))


```



Some features are in the correct data types for use in our analysis.  We will tranform them to the required data types. 



First, Sverity, which is our outcome feature, will be transformed to a factor from an integer data type. 

```{r}
#| code-fold: true
#| code-summary: "Show Code"




MA_accidents_2 <- MA_accidents |> 
    mutate(Severity = as.factor(Severity))


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"


class(MA_accidents_2$Severity)


```




We will now extract the year, month,day, and hourm of the day from the Start_Time feature. 





```{r}
#| code-fold: true
#| code-summary: "Show Code"



MA_accidents_2 <- MA_accidents_2 |> 
    mutate(Year = year(Start_Time)) |> 
    mutate(Month = month(Start_Time)) |> 
    mutate(day_of_week = wday(Start_Time)) |> 
    mutate(hour_of_day = hour(Start_Time))


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"


MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day) |> 
    head()


```

Day of week starts with Sunday (1) and ends with Saturday (7)

Hour of day starts at 1:00 am (1) and ends at 12:00 am (24)




```{r}
#| code-fold: true
#| code-summary: "Show Code"


MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day) |> 
    glimpse()



```



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


MA_accidents_2 <- MA_accidents_2 |> 
    mutate(Year = as.factor(Year)) |> 
    mutate(Month = as.factor(Month)) |> 
    mutate(day_of_week = as.factor(day_of_week))



```



```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(MA_accidents_2, "MA_accidents_2.rds")


```

```{r}
#| echo: false


MA_accidents_2 <- readRDS("MA_accidents_2.rds")



```

```{r}

MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day) |> 
    glimpse()


```

```{r}


MA_accidents_2 |> 
    count(Year)


```


Before continuing, we will filter our data into the years 2016-2021 for training and testing, and the year 2022 for predicting on unseen data.  Since the year 2023 appears to be incomplete, we will omit it from our analysis. 

<br>

Years 2016 through 2021


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_accidents_16_21 <- MA_accidents_2 |> 
    filter(Year %in% c("2016", "2017", "2018", "2019", "2020","2021"))


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_accidents_16_21, "ma_accidents_16_21.rds")

```

```{r}
#| echo: false


ma_accidents_16_21 <- readRDS("ma_accidents_16_21.rds")



```

```{r}


dim(ma_accidents_16_21)


```

Year 2022


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_accidents_22 <- MA_accidents_2 |> 
    filter(Year == "2022")


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_accidents_22, "ma_accidents_22.rds")


```

```{r}
#| echo: false


ma_accidents_22 <- readRDS("ma_accidents_22.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


dim(ma_accidents_22)


```


For us to be able to calculate the time duration of accidents we'll first need to convert the start and endtimes to a date time object that includes year, month, day, hours, minutes, and seconds. 


```{r}
#| code-fold: true
#| code-summary: "Show Code"

ma_accidents_16_21_2 <- ma_accidents_16_21 |> 
    mutate(Start_Time = ymd_hms(Start_Time)) |> 
    mutate(End_Time = ymd_hms(End_Time))
```

```{r}
#| code-fold: true
#| code-summary: "Show Code"

ma_accidents_16_21_2 |> 
    select(Start_Time, End_Time) |> 
    glimpse()


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_2 |> 
    select(Start_Time, End_Time) |> 
    head()


```

Next, we'll create a new feature accident_duration, which will calcualte the duration of an accident in minutes. 


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_2 <- ma_accidents_16_21_2 |> 
    mutate(accident_duration = round(difftime(End_Time,Start_Time, units = "mins"), 0))


```

 We also will create LATTITUDE and LONGITUDE features from Start_Lat and Start_Lng. These features will be useful for retieving US Census data and mapping. 




```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_accidents_16_21_2$LATITUDE <- ma_accidents_16_21_2$Start_Lat

ma_accidents_16_21_2$LONGITUDE <- ma_accidents_16_21_2$Start_Lng


```


```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_accidents_16_21_2, "ma_accidents_16_21_2.rds")


```


```{r}
#| echo: false


 ma_accidents_16_21_2 <- readRDS("ma_accidents_16_21_2.rds")


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_2 |> select(accident_duration) |>  head()


```

From the output, we see the accident duration is in minutes. 


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_2 |> select(LATITUDE, LONGITUDE) |> head()


```

Our output shows the two new featurres. 





## US Census Data


```{r}





long_lat_1 <- ma_accidents_16_21_3 |> 
    filter(id <=5000) |> 
    select(LATITUDE, LONGITUDE)



```


```{r}


long_lat_2 <- ma_accidents_16_21_3 |> 
    filter(id > 5000 & id <=15000) |> 
    select(LATITUDE, LONGITUDE)



```


```{r}




long_lat_3 <- ma_accidents_16_21_3 |> 
    filter(id > 15000 & id <=25000) |> 
    select(LATITUDE, LONGITUDE)

```


```{r}


long_lat_4 <- ma_accidents_16_21_3 |> 
    filter(id > 25000 & id <=35000) |> 
    select(LATITUDE, LONGITUDE)


```

```{r}


long_lat_5 <- ma_accidents_16_21_3 |> 
    filter(id > 35000 & id <=45000) |> 
    select(LATITUDE, LONGITUDE)



```

```{r}


long_lat_6 <- ma_accidents_16_21_3 |> 
    filter(id > 45000 & id <=55587) |> 
    select(LATITUDE, LONGITUDE)



```


```{r}


Sys.getenv("CENSUS_API_KEY")


```


```{r}


long_lat_1$census_code <- apply(long_lat_1, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))




```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(long_lat_1 , "long_lat_1.rds")


```

```{r}


long_lat_1  <- readRDS("long_lat_1.rds")


```


```{r}

long_lat_2$census_code <- apply(long_lat_2, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))


```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(long_lat_2, "long_lat_2.rds")

```

```{r}


long_lat_2 <- readRDS("long_lat_2.rds")


```


```{r}



long_lat_3$census_code <- apply(long_lat_3, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))


```



```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(long_lat_3, "long_lat_3.rds")


```


```{r}


long_lat_3 <- readRDS("long_lat_3.rds")



```


```{r}


long_lat_4$census_code <- apply(long_lat_4, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))


```


```{r}
#| echo: false
#| include: false
#| eval: false



saveRDS(long_lat_4, "long_lat_4.rds")



```


```{r}


long_lat_4 <- readRDS("long_lat_4.rds")


```


```{r}

long_lat_5$census_code <- apply(long_lat_5, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(long_lat_5, "long_lat_5.rds")



```


```{r}



long_lat_5 <- readRDS("long_lat_5.rds")


```


```{r}


long_lat_6$census_code <- apply(long_lat_6, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))




```



```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(long_lat_6, "long_lat_6.rds")



```


```{r}

long_lat_6 <- readRDS("long_lat_6.rds")


```


```{r}




long_lat_7$census_code <- apply(long_lat_7, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))



```




```{r}



long_lat <- rbind(long_lat_1, long_lat_2, long_lat_3, long_lat_4, long_lat_5, long_lat_6)



```


```{r}



long_lat$GEOID <- substr(long_lat$census_code, 1,11)


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(long_lat, "long_lat.rds")

```


```{r}


long_lat <- readRDS("long_lat.rds")


```

```{r}


long_lat |>  head()


```


```{r}


census_geo <- long_lat |> 
    select(-LATITUDE, -LONGITUDE)


```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(census_geo, "census_geo.rds")


```


```{r}

census_geo <- readRDS("census_geo.rds")



```


```{r}




census_geo |> head()
```


```{r}



ma_accidents_16_21_3 <- cbind(ma_accidents_16_21_2, census_geo)



```

```{r}
#| echo: false
#| include: false
#| eval: false



saveRDS(ma_accidents_16_21_3, "ma_accidents_16_21_3.rds")


```


```{r}


ma_accidents_16_21_3 <- readRDS("ma_accidents_16_21_3.rds")


```




```{r}



ma_accidents_16_21_3 |> 
    summary()

```

There are five mssing values in the newly created GEOID.  These missing values (NAs) will be removed. 


```{r}


ma_accidents_16_21_4 <- ma_accidents_16_21_3 |> 
    filter(!is.na(GEOID))


```

```{r}


ma_accidents_16_21_4 |> 
    select(GEOID, census_code) |> 
    summary()


```

```{r}


ma_accidents_16_21_4 |> 
    filter(!is.na(Wind_Speed.mph.)) |> 
    group_by(Severity) |> 
    summarise(median_wind_speed = median(Wind_Speed.mph.))


```


```{r}


ma_accidents_16_21_4 <- ma_accidents_16_21_4 |> 
    filter(!is.na(Wind_Speed.mph.))


```

```{r}

ma_accidents_16_21_4 |> 
    summary()


```

```{r}


ma_accidents_16_21_4 |> 
    filter(!is.na(Visibility.mi.)) |> 
    group_by(Severity) |> 
    summarise(median_visibility = median(Visibility.mi.))


```

```{r}


ma_accidents_16_21_4|> 
    filter(!is.na(Precipitation.in.)) |> 
    group_by(Severity) |> 
    summarise(median_visibility = median(Precipitation.in.))


```

```{r}


ma_accidents_16_21_4 |> 
    select(id, Description) |> 
    filter(id == "51113") 


```

```{r}


ma_accidents_16_21_4  |> filter(accident_duration > 1000)  |> head()


```

```{r}

ma_accidents_16_21_4  |> filter(accident_duration > 1000)  |> tail()


```

```{r}


ma_accidents_16_21_4  |> filter(accident_duration <=1000) |> dim()


```


```{r}


ma_accidents_16_21_4 <- ma_accidents_16_21_4 |> 
    select(-Start_Lat, -End_Lat, -Start_Lng, -End_Lng, -State, -Zipcode, -Country, -Timezone, -Airport_Code, -Weather_Timestamp, -Wind_Chill.F., -Visibility.mi., -Precipitation.in.)


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_accidents_16_21_4, "ma_accidents_16_21_4.rds")

```

```{r}
#| echo: false


ma_accidents_16_21_4 <- readRDS("ma_accidents_16_21_4.rds")


```

```{r}


ma_accidents_16_21_4 |> 
    summary()


```

Remove NAs fron Temperature feature.

```{r}


ma_accidents_16_21_5 <- ma_accidents_16_21_4|>
    filter(!is.na(Temperature.F.)) 


```

Remove NAs from Humidity feature. 

```{r}


ma_accidents_16_21_5 <- ma_accidents_16_21_5 |> 
    filter(!is.na(Humidity...))


```


Transform accident_duration from difftime to numeric. 

```{r}



ma_accidents_16_21_5$accident_duration <- as.numeric(ma_accidents_16_21_5$accident_duration)


```


```{r}


ma_accidents_16_21_5 |> 
    count(County)




```


Remove Dukes and Nantucket levels from Counties. 


```{r}


ma_accidents_16_21_5 <- ma_accidents_16_21_5 |> 
    filter(County  %in% c("Barnstable", "Berkshire", "Bristol", "Essex", "Franklin", "Hampden", "Hampshire", "Middlesex", "Norfolk", "Plymouth", "Suffolk", "Worcester"))


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_accidents_16_21_5, "ma_accidents_16_21_5.rds")


```




```{r}
#| echo: false


ma_accidents_16_21_5  <- readRDS("ma_accidents_16_21_5.rds")


```


```{r}

ma_accidents_16_21_5 |> 
    count(County)


```


```{r}




ma_accidents_16_21_5 |> select(accident_duration, Severity, Start_Time,End_Time, City, LONGITUDE, LATITUDE, Street, Sunrise_Sunset, Civil_Twilight,Street)  |> 
  filter(accident_duration > 5000)  |> head()



```

```{r}



ma_accidents_16_21_5  |> select(accident_duration, Severity, Start_Time,End_Time, City, LONGITUDE, LATITUDE, Street, Sunrise_Sunset, Civil_Twilight,Street)  |> 
  filter(City == "Milton" & Street == "Adams St")  |> head()


```



```{r}



ma_accidents_16_21_5  |> select(accident_duration, Severity, Start_Time,End_Time, Year, Month, day_of_week, City, LONGITUDE, LATITUDE, Street, Sunrise_Sunset, Civil_Twilight,Street)  |> 
  filter(Year == "2021" & Month == "4" & day_of_week == "4")  |> tail()

```



```{r}




ma_accidents_16_21_5 |> 
    select_if(is.numeric) |> 
    glimpse()


```

hour_of_day is numeric datatype.  We'll need to transform this feature to a factor for our analysis and modeling. 


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_accidents_16_21_6 <- ma_accidents_16_21_5 |> 
    mutate(hour_of_day = as.factor(hour_of_day))


```

 ```{r}
 #| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_6 <- ma_accidents_16_21_6 |> 
    rename(temperature = Temperature.F., humidity = Humidity..., wind_speed = Wind_Speed.mph.)


 ```

 ```{r}
 #| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_6 <- ma_accidents_16_21_6 |> 
    select(-Source, -Start_Time, -End_Time, -Distance.mi., -Pressure.in. )


 ```

 ```{r}
 #| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_6 <- ma_accidents_16_21_6 |> 
    mutate(Wind_Direction = as.factor(Wind_Direction),
            Weather_Condition = as.factor(Weather_Condition),
            Amenity = as.factor(Amenity),
            Bump = as.factor(Bump),
            Crossing = as.factor(Crossing),
            Give_Way = as.factor(Give_Way),
            Junction = as.factor(Junction),
            No_Exit = as.factor(No_Exit),
            Railway = as.factor(Railway),
            Roundabout = as.factor(Roundabout),
            Station = as.factor(Station),
            Stop = as.factor(Stop),
            Traffic_Calming = as.factor(Traffic_Calming),
            Traffic_Signal = as.factor(Traffic_Signal),
            Turning_Loop = as.factor(Turning_Loop),
            Sunrise_Sunset = as.factor(Sunrise_Sunset),
            Civil_Twilight = as.factor(Civil_Twilight),
            Nautical_Twilight = as.factor(Nautical_Twilight),
            Astronomical_Twilight = as.factor(Astronomical_Twilight))


 ```

 ```{r}
#| echo: false
#| include: false
#| eval: false
 

saveRDS(ma_accidents_16_21_6 , "ma_accidents_16_21_6.rds")


 ```

 ```{r}
 #| echo: false 
 

ma_accidents_16_21_6  <- readRDS("ma_accidents_16_21_6.rds")



 ```

 ```{r}
 

ma_accidents_16_21_6 |> 
    select_if(is.factor) |> 
    summary()


 ```


 ```{r}
 


ma_accidents_16_21_6 |> 
    count(Crossing) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
 

ma_accidents_16_21_6 |> 
    count(Wind_Direction) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 


 ```{r}
 

ma_accidents_16_21_6 |> 
    count(County, Bump)



 ```


 
 ```{r}
 #| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 


ma_accidents_16_21_7 <- ma_accidents_16_21_6 |> 
    select(-Bump, -Roundabout, -Turning_Loop)

 
 ```

 ```{r}
 #| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 


ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    mutate(Street = recode(Street,  "Mass Pike" = "I-90"))


 ```

 

 

 ```{r}
 #| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 



# Force-create the category column directly onto your existing dataframe
# (Replace 'Street' below if your column is lowercase 'street')
col_to_check <- ma_accidents_16_21_7$Street

ma_accidents_16_21_7$road_category <- "Uncategorized"

# Apply regex flags directly
ma_accidents_16_21_7$road_category[grepl("\\b(I|Interstate)[-\\s]?\\d+", col_to_check, ignore.case = TRUE)] <- "Interstate"
ma_accidents_16_21_7$road_category[grepl("\\b(Route|Rte|Rt|State Route|SR|State Hwy|MA)[. -]*\\d+", col_to_check, ignore.case = TRUE)] <- "MA Highway"
ma_accidents_16_21_7$road_category[grepl("\\b(US)[-\\s]?\\d+", col_to_check, ignore.case = TRUE)] <- "US Route"
ma_accidents_16_21_7$road_category[grepl("\\b(Rd|Road)\\b", col_to_check, ignore.case = TRUE)] <- "Road"
ma_accidents_16_21_7$road_category[grepl("way(\\b|[ .]+(N|S|E|W|NE|NW|SE|SW|North|South|East|West)?$)", col_to_check, ignore.case = TRUE)] <- "Way"
ma_accidents_16_21_7$road_category[grepl("\\b(St|Street)\\b", col_to_check, ignore.case = TRUE)] <- "Street"
ma_accidents_16_21_7$road_category[grepl("\\b(Ave|Avenue)\\b", col_to_check, ignore.case = TRUE)] <- "Avenue"
ma_accidents_16_21_7$road_category[grepl("\\b(Blvd|Boulevard)\\b", col_to_check, ignore.case = TRUE)] <- "Boulevard"
ma_accidents_16_21_7$road_category[grepl("\\b(Dr|Drive)\\b", col_to_check, ignore.case = TRUE)] <- "Drive"
ma_accidents_16_21_7$road_category[grepl("\\b(Ln|Lane)\\b", col_to_check, ignore.case = TRUE)] <- "Lane"
ma_accidents_16_21_7$road_category[grepl("\\b(Hwy|Highway)\\b", col_to_check, ignore.case = TRUE)] <- "Highway"
ma_accidents_16_21_7$road_category[grepl("\\b(Pkwy|Parkway)\\b", col_to_check, ignore.case = TRUE)] <- "Parkway"
ma_accidents_16_21_7$road_category[grepl("\\b(Brg|Bridge)\\b", col_to_check, ignore.case = TRUE)] <- "Bridge"
ma_accidents_16_21_7$road_category[grepl("\\b(Cir|Circle)\\b", col_to_check, ignore.case = TRUE)] <- "Circle"
ma_accidents_16_21_7$road_category[grepl("\\b(Trl|Trail)\\b", col_to_check, ignore.case = TRUE)] <- "Trail"
ma_accidents_16_21_7$road_category[grepl("\\b(Ct|Cor|Court)\\b", col_to_check, ignore.case = TRUE)] <- "Court"
ma_accidents_16_21_7$road_category[grepl("\\b(Ter|Terrace)\\b", col_to_check, ignore.case = TRUE)] <- "Terrace"
ma_accidents_16_21_7$road_category[grepl("\\b(Pl|Place)\\b", col_to_check, ignore.case = TRUE)] <- "Place"
ma_accidents_16_21_7$road_category[grepl("\\b(Tpke|Turnpike)\\b", col_to_check, ignore.case = TRUE)] <- "Turnpike"
ma_accidents_16_21_7$road_category[grepl("\\b(Sq|Square)\\b", col_to_check, ignore.case = TRUE)] <- "Square"
ma_accidents_16_21_7$road_category[grepl("\\b(Pth|Path)\\b", col_to_check, ignore.case = TRUE)] <- "Path"
ma_accidents_16_21_7$road_category[grepl("\\b(Conn|Connector)\\b", col_to_check, ignore.case = TRUE)] <- "Connector"
ma_accidents_16_21_7$road_category[grepl("\\b(Ext|Exit)\\b", col_to_check, ignore.case = TRUE)] <- "Exit"
ma_accidents_16_21_7$road_category[grepl("\\b(Rty|Rotary)\\b", col_to_check, ignore.case = TRUE)] <- "Rotary"
ma_accidents_16_21_7$road_category[grepl("\\b(Expy|Expressway)\\b", col_to_check, ignore.case = TRUE)] <- "Expressway"
ma_accidents_16_21_7$road_category[grepl("\\b(Tunl|Tunnel)\\b", col_to_check, ignore.case = TRUE)] <- "Tunnel"


 ```

 

 ```{r}
 
ma_accidents_16_21_6 |> count(Weather_Condition)


 ```

 ```{r}
 

ma_accidents_16_21_6 |> select(Weather_Condition, City) |> 
    filter(Weather_Condition == "") |> 
    head(15)


 ```

 ```{r}
 #| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 
ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    mutate(Weather_Condition = as.character(Weather_Condition))


 ```

 ```{r}
 #| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_7$Weather_Condition[trimws(ma_accidents_16_21_7$Weather_Condition) == ""] <- "NR"


 ```

 ```{r}
 #| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    mutate(Weather_Condition = as.factor(Weather_Condition))


 ```

 ```{r}
 #| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    mutate(Weather_Condition = recode(Weather_Condition,  " " = "No Report", "Light Rain / Windy" = "Light Rain", "Light Rain Shower" = "Light Rain", "Light Rain with Thunder" = "Light Rain", "Light Drizzle" = "Light Rain","Rain / Windy" = "Rain", "Thunderstorms and Rain" = "Rain", "Thunderstorm" = "Rain", "T-Storm" = "Rain","Thunder" = "Overcast", "Cloudy" = "Overcast", "Thunder / Windy" = "Overcast", "Thunder in the Vicinity" = "Overcast","Cloudy / Windy" = "Overcast", "Mostly Cloudy" = "Overcast", "Mostly Cloudy / Windy" = "Overcast","Light Thunderstorms and Rain" = "Rain","Heavy Rain / Windy" = "Heavy Rain", "Heavy Thunderstorms and Rain" = "Heavy Rain","Heavy T-Storm / Windy" = "Heavy Rain","Heavy T-Storm" = "Heavy Rain", "Light Snow / Windy" = "Light Snow", "Light Snow and Sleet" = "Light Snow", "Clear" = "Fair Partley Cloudy","Blowing Snow" = "Heavy Snow", "Heavy Snow / Windy" = "Heavy Snow", "Light Drizzle / Windy" = "Light Rain", "Mist" = "Light Rain", "Drizzle and Fog" = "Light Rain", "Fair / Windy" = "Fair Partley Cloudy", "Partly Cloudy" = "Fair Partley Cloudy", "Partly Cloudy / Windy" = "Fair Partley Cloudy", "Fair" = "Fair Partley Cloudy", "Scattered Clouds" = "Fair Partley Cloudy",
   "Fog / Windy" = "Fog", "Patches of Fog" = "Fog","Shallow Fog" = "Fog","Wintry Mix / Windy" = "Wintry Mix", "Light Freezing Drizzle" = "Wintry Mix" , "Light Freezing Rain" = "Wintry Mix","Smoke" = "Haze", "Squalls" = "Snow",
   "Light Freezing Fog" = "Wintry Mix", "Ice Pellets" = "Wintry Mix","Small Hail" = "Wintry Mix",  "Snow / Windy" = "Snow", "N/A Precipitation" = "NR") )


 ```

 ```{r}
 #| echo: false
#| include: false
#| eval: false
 

saveRDS(ma_accidents_16_21_7, "ma_accidents_16_21_7.rds")



 ```


 ```{r}
#| echo: false
 

ma_accidents_16_21_7 <- readRDS("ma_accidents_16_21_7.rds")


 ```

 ```{r}
 

ma_accidents_16_21_7 |> 
    count(road_category)


 ```

 ```{r}
 

ma_accidents_16_21_7 |> 
    count(Weather_Condition)



 ```


 ```{r}
 

ma_accidents_16_21_7 |> 
    count(No_Exit) |> 
    mutate(perc = round(n/sum(n),4)) |> 
    arrange(desc(perc))


 ```

 ```{r}
 
ma_accidents_16_21_7 |> 
    count(County, No_Exit)

 ```

 ```{r}
 

ma_accidents_16_21_7 |> 
    count(Crossing) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
 
ma_accidents_16_21_7 |> 
    count(Railway) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))



 ```

 ```{r}
 


 ma_accidents_16_21_7 |> 
    count(County, Railway)

 ```

 ```{r}
 
ma_accidents_16_21_7 |> 
    count(Give_Way) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
 


ma_accidents_16_21_7 |> 
    count(County, Give_Way)


 ```

 ```{r}
 

test_df <- test_df |> 
    mutate(Wind_Direction = recode(Wind_Direction,  "VAR" = "Variable", "S" = "South", "N" = "North", "E" = "East", "W" = "West"))



 ```

 ```{r}
 





 ```



```{r}


ma_accidents_16_21_7 <- ma_accidents_16_21_6 |> 
    select(-No_Exit, -Bump)



```



  No_Exit 

 ```{r}
 

ma_accidents_16_21_6 |> 
    group_by(Traffic_Calming) |> 
    summarise(median_acc_dur = median(accident_duration))


 ```


 ```{r}
 


 ma_accidents_16_21_6 |> 
    group_by(No_Exit ) |> 
    summarise(median_acc_dur = median(accident_duration))
 ```


 ```{r}
 



ma_accidents_16_21_7 <- ma_accidents_16_21_6 |> 
    select(-No_Exit, -Bump, -Traffic_Calming)


 ```






```{r}





ny_housing_census_2 <- ny_housing_census_2 |> 
    mutate(TYPE = recode_values(TYPE, "Condop for sale" ~ "Condo for sale"))


```

```{r}


class(ma_accidents_16_21_7$hour_of_day)


```


```{r}





```

```{r}



ma_accidents_16_21_3  <- ma_accidents_16_21_2  |> 
    mutate(id = row_number())


```





# Census

## County




```{r}


ma_county <- counties("MA", year = 2024)

```

```{r}

saveRDS(ma_county, "ma_county.rds")


```

```{r}


ma_county <- readRDS("ma_county.rds")


```

```{r}


ma_county


```


```{r}


census_county_urban <- readxl::read_xlsx("/Users/francislauder/Documents/census/2020_UA_COUNTY.xlsx")


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(census_county_urban, "census_county_urban.rds")


```

```{r}
#| echo: false

census_county_urban <- readRDS("census_county_urban.rds")


```


```{r}


census_county_urban |> glimpse()


```

```{r}


census_county_urban |> 
    count(STATE)


```


```{r}


ma_county_urban <- census_county_urban |> 
    filter(STATE_NAME == "Massachusetts")


```


```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_county_urban , "ma_county_urban.rds")


```

```{r}
#| echo: false


ma_county_urban  <- readRDS("ma_county_urban.rds")

```

```{r}


ma_county_urban |> summary()


```

```{r}


ma_county_urban |> glimpse()


```

```{r}


ma_county_urban |> 
    select(COUNTY_NAME, POPDEN_COU, POPPCT_URB, HOUDEN_COU, HOUDEN_URB,HOUDEN_RUR, HOUPCT_URB)


```

## Tracts

```{r}


ma_tracts <- tracts("MA", year = 2024)


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tracts, "ma_tracts.rds")


```

```{r}

ma_tracts <- readRDS("ma_tracts.rds")


```

```{r}

ma_tracts


```

## Joining


# Exploratory Data Analysis


## Features


```{r}


ma_accidents_16_21_6 |> 
    ggplot(aes(x=accident_duration, y=Temperature.F.)) +
    geom_point()



```

```{r}



ma_accidents_16_21_6 |> 
    ggplot(aes(x=log(accident_duration), y=Temperature.F.)) +
    geom_point()


```



# Preprocessing


```{r}



```


#rows / columns 
10:1 rule 
You should have at least 10 rows of data for every 1 column (feature) in your dataset. 

```{r}


# Allocates 80% to training, 20% to testing
data_split <- initial_split(your_dataframe, prop = 0.80) 


```

```{r}


ma_county_urban_test <- ma_county_urban |> 
    select(COUNTY, COUNTY_NAME, POP_COU, HOU_COU, POPDEN_COU, HOUDEN_COU, POPPCT_URB, HOU_URB, HOUPCT_URB)


```

```{r}


test_df <- ma_accidents_16_21_6 |> 
    select(-Severity, -Bump, -Roundabout, -Turning_Loop, -Nautical_Twilight, -Astronomical_Twilight, -Description, -Street, -City, -LATITUDE, -LONGITUDE, -census_code, -ID, -GEOID, -Weather_Condition)



```



```{r}



test_df <- test_df |> 
    mutate(Wind_Direction = recode(Wind_Direction,  "VAR" = "Variable", "S" = "South", "N" = "North", "E" = "East", "W" = "West"))

```


```{r}

test_df <- test_df |> 
    rename(COUNTY_NAME = County)


```


```{r}


test_df_2 <- test_df |> 
    left_join(ma_county_urban_test, by = "COUNTY_NAME")


```

```{r}

test_df_2 <- test_df_2 |> 
    select(-COUNTY_NAME)


```

```{r}

test_model_df <- initial_split(test_df_2, prop = 0.80)


```


```{r}





```


```{r}


train_data <- training(test_model_df)


```

```{r}

test_data <- testing(test_model_df)


```

```{r}


# Stratify by a categorical or numeric target column
data_split <- initial_split(your_dataframe, prop = 0.80, strata = target_variable)


```



```{r}



set.seed(1353)
car_split <- initial_validation_split(mtcars)
train_data <- training(car_split)
validation_data <- validation(car_split)
test_data <- testing(car_split)


```

```{r}



validation_split(data, prop = 3/4, strata = NULL, breaks = 4, pool = 0.1, ...)

validation_time_split(data, prop = 3/4, lag = 0, ...)

group_validation_split(data, group, prop = 3/4, ..., strata = NULL, pool = 0.1)


```

```{r}


print(test_model_df)


```

```{r}


rec <- recipe(PRICE ~ ., data = model_df) |> 
    step_center(all_numeric_predictors()) |> 
    step_scale(all_numeric_predictors())


```

```{r}


rec <- recipe(accident_duration ~ ., data = train_data) |> 
    step_center(all_numeric_predictors()) |> 
    step_scale(all_numeric_predictors())


```

```{r}



model_rec <- prep(rec, training= model_df)
scaled_df <- bake(model_rec, new_data = model_df)


```

```{r}

train_rec <- prep(rec, training= train_data)
scaled_train <- bake(train_rec, new_data = train_data)


```

```{r}


model_lme_1 <- lmer(log(accident_duration) ~ . + (1 | COUNTY), data = scaled_train)



```

```{r}



# Extract fixed-effect formula without random effects
ff <- nobars(formula(your_model)
X <- model.matrix(ff, data = your_data)
)
# Check for linear combinations
caret::findLinearCombos(X)


```

```{r}


library(caret)
```

```{r}

# Extract fixed-effect formula without random effects
ff <- nobars(formula(model_lme_1))
X <- model.matrix(ff, data = train_data) 


# Check for linear combinations
caret::findLinearCombos(X)



```

```{r}


summary(model_lme_1)



```

```{r}


icc(model_lme_1, nrep=50)

```


