---
title: "Predicting time of auto accidents"
author: Francis Lauder
date: 2026-09-10
format:
  html:
    toc: true
    toc-depth: 3
    toc_float: true
    number-sections: false
    toc-location: left
    embed-resources: true
---

# Introduction
<br><br>


# Data Dictionary

<br><br>


# Libraries

<br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"

library(dplyr)
library(ggplot2)
#library(plotly)
library(RColorBrewer)
library(forcats)
library(tidycensus)

library(sf)

library(jsonlite)

library(tigris)

library(stringr)

library(leaflet)
library(htmlwidgets)
library(htmltools)
library(gt)
library(DT)
library(GGally)
library(lme4)
library(lmerTest)
library(rstanarm)
library(brms)
library(broom)
library(lubridate)
library(tidymodels)
library(performance)
library(boot.pval)
library(bayesplot)
library(units)
library(tidybayes)
library(marginaleffects)
library(collapse)
library(ggfortify)
library(bayesplot)
library(tidyr)






```

<br><br>


# Data Inspection and Cleaning

## US Accident Data


```{r}
#| echo: false
#| include: false
#| eval: false


US_accidents <- read.csv("/Users/francislauder/Documents/Data Sets/US Acciidents/US_Accidents_March23.csv")


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(US_accidents, "US_accidents.rds")

```

```{r}
#| echo: false
#| include: false
#| eval: false


US_accidents <- readRDS("US_accidents.rds")


```



```{r}
#| echo: false
#| include: false
#| eval: false



glimpse(US_accidents)


```



```{r}
#| echo: false
#| include: false
#| eval: false


MA_accidents <- US_accidents |> 
    filter(State == "MA")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


MA_accidents <- MA_accidents %>% 
  mutate(Weather_Timestamp = as.Date(Weather_Timestamp))


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(MA_accidents, "MA_accidents.rds")



```

```{r}
#| echo: false


MA_accidents <- readRDS("MA_accidents.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



options(browser="false", scipen = 999, max.print = 20000)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



MA_accidents |> glimpse()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


summary(MA_accidents)


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"


dim(MA_accidents)


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"



MA_accidents |> 
    count(Severity) |> 
    mutate(perc = round(n / sum(n),3)) |> 
    arrange(desc(perc))


```



Some features are not in the correct data types for use in our analysis.  We will tranform them to the required data types. 



First, Sverity,  will be transformed to a factor from an integer data type. 

```{r}
#| code-fold: true
#| code-summary: "Show Code"




MA_accidents_2 <- MA_accidents |> 
    mutate(Severity = as.factor(Severity))


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"


class(MA_accidents_2$Severity)


```




We will now extract the year, month,day, and hourm of the day from the Start_Time feature. 





```{r}
#| code-fold: true
#| code-summary: "Show Code"



MA_accidents_2 <- MA_accidents_2 |> 
    mutate(Year = year(Start_Time)) |> 
    mutate(Month = month(Start_Time)) |> 
    mutate(day_of_week = wday(Start_Time)) |>
    mutate(day_type = if_else(wday(Start_Time) %in% c(1,7), "Weekend", "Weekday")) |> 
    mutate(hour_of_day = hour(Start_Time))


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"


MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day, day_type) |> 
    head()


```

Day of week starts with Sunday (1) and ends with Saturday (7)

Hour of day starts at 1:00 am (1) and ends at 12:00 am (24)




```{r}
#| code-fold: true
#| code-summary: "Show Code"


MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day) |> 
    glimpse()



```



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


MA_accidents_2 <- MA_accidents_2 |> 
    mutate(Year = as.factor(Year)) |> 
    mutate(Month = as.factor(Month)) |> 
    mutate(day_of_week = as.factor(day_of_week)) |> 
    mutate(day_type = as.factor(day_type))



```



```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(MA_accidents_2, "MA_accidents_2.rds")


```

```{r}
#| echo: false


MA_accidents_2 <- readRDS("MA_accidents_2.rds")



```

```{r}

MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day) |> 
    glimpse()


```

```{r}


MA_accidents_2 |> 
    count(Year)


```


Before continuing, we will filter our data into the years 2016-2021 for training and testing, and the year 2022 for predicting on unseen data.  Since the year 2023 appears to be incomplete, we will omit it from our analysis. 

<br>

Years 2016 through 2021


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_accidents_16_21 <- MA_accidents_2 |> 
    filter(Year %in% c("2016", "2017", "2018", "2019", "2020","2021"))


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_accidents_16_21, "ma_accidents_16_21.rds")

```

```{r}
#| echo: false


ma_accidents_16_21 <- readRDS("ma_accidents_16_21.rds")



```

```{r}


dim(ma_accidents_16_21)


```

Year 2022


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_accidents_22 <- MA_accidents_2 |> 
    filter(Year == "2022")


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_accidents_22, "ma_accidents_22.rds")


```

```{r}
#| echo: false


ma_accidents_22 <- readRDS("ma_accidents_22.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


dim(ma_accidents_22)


```


For us to be able to calculate the time duration of accidents we'll first need to convert the start and endtimes to a date time object that includes year, month, day, hours, minutes, and seconds. 


```{r}
#| code-fold: true
#| code-summary: "Show Code"

ma_accidents_16_21_2 <- ma_accidents_16_21 |> 
    mutate(Start_Time = ymd_hms(Start_Time)) |> 
    mutate(End_Time = ymd_hms(End_Time))
```

```{r}
#| code-fold: true
#| code-summary: "Show Code"

ma_accidents_16_21_2 |> 
    select(Start_Time, End_Time) |> 
    glimpse()


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_2 |> 
    select(Start_Time, End_Time) |> 
    head()


```

Next, we'll create a new feature accident_duration, which will calcualte the duration of an accident in minutes. 


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_2 <- ma_accidents_16_21_2 |> 
    mutate(accident_duration = round(difftime(End_Time,Start_Time, units = "mins"), 0))


```

 We also will create LATTITUDE and LONGITUDE features from Start_Lat and Start_Lng. These features will be useful for retieving US Census data and mapping. 




```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_accidents_16_21_2$LATITUDE <- ma_accidents_16_21_2$Start_Lat

ma_accidents_16_21_2$LONGITUDE <- ma_accidents_16_21_2$Start_Lng


```


```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_accidents_16_21_2, "ma_accidents_16_21_2.rds")


```


```{r}
#| echo: false


 ma_accidents_16_21_2 <- readRDS("ma_accidents_16_21_2.rds")


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_2 |> select(accident_duration) |>  head()


```

From the output, we see the accident duration is in minutes. 


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_2 |> select(LATITUDE, LONGITUDE) |> head()


```

Our output shows the two new featurres. 





## US Census Data


```{r}





long_lat_1 <- ma_accidents_16_21_3 |> 
    filter(id <=5000) |> 
    select(LATITUDE, LONGITUDE)



```


```{r}


long_lat_2 <- ma_accidents_16_21_3 |> 
    filter(id > 5000 & id <=15000) |> 
    select(LATITUDE, LONGITUDE)



```


```{r}




long_lat_3 <- ma_accidents_16_21_3 |> 
    filter(id > 15000 & id <=25000) |> 
    select(LATITUDE, LONGITUDE)

```


```{r}


long_lat_4 <- ma_accidents_16_21_3 |> 
    filter(id > 25000 & id <=35000) |> 
    select(LATITUDE, LONGITUDE)


```

```{r}


long_lat_5 <- ma_accidents_16_21_3 |> 
    filter(id > 35000 & id <=45000) |> 
    select(LATITUDE, LONGITUDE)



```

```{r}


long_lat_6 <- ma_accidents_16_21_3 |> 
    filter(id > 45000 & id <=55587) |> 
    select(LATITUDE, LONGITUDE)



```


```{r}
#| code-fold: true
#| code-summary: "Show Code"



Sys.getenv("CENSUS_API_KEY")


```


```{r}


long_lat_1$census_code <- apply(long_lat_1, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))




```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(long_lat_1 , "long_lat_1.rds")


```

```{r}


long_lat_1  <- readRDS("long_lat_1.rds")


```


```{r}

long_lat_2$census_code <- apply(long_lat_2, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))


```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(long_lat_2, "long_lat_2.rds")

```

```{r}


long_lat_2 <- readRDS("long_lat_2.rds")


```


```{r}



long_lat_3$census_code <- apply(long_lat_3, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))


```



```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(long_lat_3, "long_lat_3.rds")


```


```{r}


long_lat_3 <- readRDS("long_lat_3.rds")



```


```{r}


long_lat_4$census_code <- apply(long_lat_4, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))


```


```{r}
#| echo: false
#| include: false
#| eval: false



saveRDS(long_lat_4, "long_lat_4.rds")



```


```{r}


long_lat_4 <- readRDS("long_lat_4.rds")


```


```{r}

long_lat_5$census_code <- apply(long_lat_5, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(long_lat_5, "long_lat_5.rds")



```


```{r}



long_lat_5 <- readRDS("long_lat_5.rds")


```


```{r}


long_lat_6$census_code <- apply(long_lat_6, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))




```



```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(long_lat_6, "long_lat_6.rds")



```


```{r}

long_lat_6 <- readRDS("long_lat_6.rds")


```


```{r}




long_lat_7$census_code <- apply(long_lat_7, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))



```




```{r}



long_lat <- rbind(long_lat_1, long_lat_2, long_lat_3, long_lat_4, long_lat_5, long_lat_6)



```


```{r}



long_lat$GEOID <- substr(long_lat$census_code, 1,11)


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(long_lat, "long_lat.rds")

```


```{r}


long_lat <- readRDS("long_lat.rds")


```

```{r}


long_lat |>  head()


```


```{r}


census_geo <- long_lat |> 
    select(-LATITUDE, -LONGITUDE)


```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(census_geo, "census_geo.rds")


```


```{r}

census_geo <- readRDS("census_geo.rds")



```


```{r}




census_geo |> head()
```


```{r}



ma_accidents_16_21_3 <- cbind(ma_accidents_16_21_2, census_geo)



```

```{r}
#| echo: false
#| include: false
#| eval: false



saveRDS(ma_accidents_16_21_3, "ma_accidents_16_21_3.rds")


```


```{r}


ma_accidents_16_21_3 <- readRDS("ma_accidents_16_21_3.rds")


```




```{r}



ma_accidents_16_21_3 |> 
    summary()

```



There are five mssing values in the newly created GEOID.  These missing values (NAs) will be removed. 


```{r}


ma_accidents_16_21_4 <- ma_accidents_16_21_3 |> 
    filter(!is.na(GEOID))


```

```{r}


ma_accidents_16_21_4 |> 
    select(GEOID, census_code) |> 
    summary()


```


```{r}


ma_accidents_16_21_3 |> 
    select(GEOID,County, Wind_Speed.mph.) |> 
    filter(is.na(Wind_Speed.mph.)) |> 
    head(72)


```

```{r}


ma_accidents_16_21_4 |> 
    filter(!is.na(Wind_Speed.mph.)) |> 
    group_by(Severity) |> 
    summarise(median_wind_speed = median(Wind_Speed.mph.))


```


```{r}
#| eval: false


ma_accidents_16_21_4 <- ma_accidents_16_21_4 |> 
    filter(!is.na(Wind_Speed.mph.))


```

```{r}

ma_accidents_16_21_4 |> 
    summary()


```

```{r}


ma_accidents_16_21_4 |> 
    filter(!is.na(Visibility.mi.)) |> 
    group_by(Severity) |> 
    summarise(median_visibility = median(Visibility.mi.))


```

```{r}


ma_accidents_16_21_4|> 
    filter(!is.na(Precipitation.in.)) |> 
    group_by(Severity) |> 
    summarise(median_visibility = median(Precipitation.in.))


```

```{r}


ma_accidents_16_21_4 |> 
    select(id, Description) |> 
    filter(id == "51113") 


```

```{r}


ma_accidents_16_21_4  |> filter(accident_duration > 1000)  |> head()


```

```{r}

ma_accidents_16_21_4  |> filter(accident_duration > 1000)  |> tail()


```

```{r}


ma_accidents_16_21_4  |> filter(accident_duration <=1000) |> dim()


```


```{r}


ma_accidents_16_21_4 <- ma_accidents_16_21_4 |> 
    select(-Start_Lat, -End_Lat, -Start_Lng, -End_Lng, -State, -Zipcode, -Country, -Timezone, -Airport_Code, -Weather_Timestamp, -Wind_Chill.F., -Visibility.mi., -Precipitation.in., -Wind_Speed.mph.)


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_accidents_16_21_4, "ma_accidents_16_21_4.rds")

```

```{r}
#| echo: false


ma_accidents_16_21_4 <- readRDS("ma_accidents_16_21_4.rds")


```

```{r}


ma_accidents_16_21_4 |> 
    summary()


```

Remove NAs fron Temperature feature.

```{r}


ma_accidents_16_21_5 <- ma_accidents_16_21_4|>
    filter(!is.na(Temperature.F.)) 


```

Remove NAs from Humidity feature. 

```{r}


ma_accidents_16_21_5 <- ma_accidents_16_21_5 |> 
    filter(!is.na(Humidity...))


```


Transform accident_duration from difftime to numeric. 

```{r}



ma_accidents_16_21_5$accident_duration <- as.numeric(ma_accidents_16_21_5$accident_duration)


```


```{r}


ma_accidents_16_21_5 |> 
    count(County)




```


Remove Dukes and Nantucket levels from Counties. 


```{r}


ma_accidents_16_21_5 <- ma_accidents_16_21_5 |> 
    filter(County  %in% c("Barnstable", "Berkshire", "Bristol", "Essex", "Franklin", "Hampden", "Hampshire", "Middlesex", "Norfolk", "Plymouth", "Suffolk", "Worcester"))


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_accidents_16_21_5, "ma_accidents_16_21_5.rds")


```




```{r}
#| echo: false


ma_accidents_16_21_5  <- readRDS("ma_accidents_16_21_5.rds")


```


```{r}

ma_accidents_16_21_5 |> 
    count(County)


```


```{r}




ma_accidents_16_21_5 |> select(accident_duration, Severity, Start_Time,End_Time, City, LONGITUDE, LATITUDE, Street, Sunrise_Sunset, Civil_Twilight,Street)  |> 
  filter(accident_duration > 5000)  |> head()



```

```{r}



ma_accidents_16_21_5  |> select(accident_duration, Severity, Start_Time,End_Time, City, LONGITUDE, LATITUDE, Street, Sunrise_Sunset, Civil_Twilight,Street)  |> 
  filter(City == "Milton" & Street == "Adams St")  |> head()


```



```{r}



ma_accidents_16_21_5  |> select(accident_duration, Severity, Start_Time,End_Time, Year, Month, day_of_week, City, LONGITUDE, LATITUDE, Street, Sunrise_Sunset, Civil_Twilight,Street)  |> 
  filter(Year == "2021" & Month == "4" & day_of_week == "4")  |> tail()

```



```{r}




ma_accidents_16_21_5 |> 
    select_if(is.numeric) |> 
    glimpse()


```





 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_6 <- ma_accidents_16_21_5 |> 
    rename(temperature = Temperature.F.)


 ```

 ```{r}
 


ma_accidents_16_21_5 |> 
    count(Wind_Direction)



 ```


 ```{r}
 

ma_accidents_16_21_6$Wind_Direction[trimws(ma_accidents_16_21_6$Wind_Direction) == ""] <- "NR"


 ```

 ```{r}
 


ma_accidents_16_21_6 <- ma_accidents_16_21_6 |> 
    mutate(Wind_Direction = as.factor(Wind_Direction))


 ```

 ```{r}
 




ma_accidents_16_21_6 <- ma_accidents_16_21_6 |> 
    mutate(Wind_Direction = recode(Wind_Direction,  "CALM" = "Calm", "NR" = "Calm", "ENE" = "E", "ESE" = "E", "East" = "E", "NE" = "N", "NNE" = "N", "NNW" = "N", "NW" = "N", "North" = "N", "SE" = "S", 
                                                    "SSE" = "S", "SSW" ="S", "SW" = "S", "South" = "S", "VAR" = "Variable", "WNW" = "W", "WSW" = "W", "West" = "W"))



 ```



 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_6 <- ma_accidents_16_21_6 |> 
    select(-Source, -Start_Time, -End_Time, -Distance.mi., -Pressure.in.,-Humidity... )


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_6 <- ma_accidents_16_21_6 |> 
    mutate(Wind_Direction = as.factor(Wind_Direction),
            Weather_Condition = as.factor(Weather_Condition),
            Amenity = as.factor(Amenity),
            Bump = as.factor(Bump),
            Crossing = as.factor(Crossing),
            Give_Way = as.factor(Give_Way),
            Junction = as.factor(Junction),
            No_Exit = as.factor(No_Exit),
            Railway = as.factor(Railway),
            Roundabout = as.factor(Roundabout),
            Station = as.factor(Station),
            Stop = as.factor(Stop),
            Traffic_Calming = as.factor(Traffic_Calming),
            Traffic_Signal = as.factor(Traffic_Signal),
            Turning_Loop = as.factor(Turning_Loop),
            Sunrise_Sunset = as.factor(Sunrise_Sunset),
            Civil_Twilight = as.factor(Civil_Twilight),
            Nautical_Twilight = as.factor(Nautical_Twilight),
            Astronomical_Twilight = as.factor(Astronomical_Twilight))


 ```

 ```{r}
#| echo: false
#| include: false
#| eval: false
 

saveRDS(ma_accidents_16_21_6 , "ma_accidents_16_21_6.rds")


 ```

 ```{r}
 #| echo: false 
 

ma_accidents_16_21_6  <- readRDS("ma_accidents_16_21_6.rds")



 ```

 ```{r}
 

ma_accidents_16_21_6 |> 
    select_if(is.factor) |> 
    summary()


 ```


 ```{r}
 


ma_accidents_16_21_6 |> 
    count(Crossing) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
 

ma_accidents_16_21_6 |> 
    count(Wind_Direction) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 


 ```{r}
 

ma_accidents_16_21_6 |> 
    count(County, Bump)



 ```




 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 


ma_accidents_16_21_7 <- ma_accidents_16_21_6 |> 
    mutate(Street = recode(Street,  "Mass Pike" = "I-90"))


 ```

 

 

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 



# Force-create the category column directly onto your existing dataframe
# (Replace 'Street' below if your column is lowercase 'street')
col_to_check <- ma_accidents_16_21_7$Street

ma_accidents_16_21_7$road_category <- "Uncategorized"

# Apply regex flags directly
ma_accidents_16_21_7$road_category[grepl("\\b(I|Interstate)[-\\s]?\\d+", col_to_check, ignore.case = TRUE)] <- "Interstate"
ma_accidents_16_21_7$road_category[grepl("\\b(Route|Rte|Rt|State Route|SR|State Hwy|MA)[. -]*\\d+", col_to_check, ignore.case = TRUE)] <- "MA Highway"
ma_accidents_16_21_7$road_category[grepl("\\b(US)[-\\s]?\\d+", col_to_check, ignore.case = TRUE)] <- "US Route"
ma_accidents_16_21_7$road_category[grepl("\\b(Rd|Road)\\b", col_to_check, ignore.case = TRUE)] <- "Road"
ma_accidents_16_21_7$road_category[grepl("way(\\b|[ .]+(N|S|E|W|NE|NW|SE|SW|North|South|East|West)?$)", col_to_check, ignore.case = TRUE)] <- "Way"
ma_accidents_16_21_7$road_category[grepl("\\b(St|Street)\\b", col_to_check, ignore.case = TRUE)] <- "Street"
ma_accidents_16_21_7$road_category[grepl("\\b(Ave|Avenue)\\b", col_to_check, ignore.case = TRUE)] <- "Avenue"
ma_accidents_16_21_7$road_category[grepl("\\b(Blvd|Boulevard)\\b", col_to_check, ignore.case = TRUE)] <- "Boulevard"
ma_accidents_16_21_7$road_category[grepl("\\b(Dr|Drive)\\b", col_to_check, ignore.case = TRUE)] <- "Drive"
ma_accidents_16_21_7$road_category[grepl("\\b(Ln|Lane)\\b", col_to_check, ignore.case = TRUE)] <- "Lane"
ma_accidents_16_21_7$road_category[grepl("\\b(Hwy|Highway)\\b", col_to_check, ignore.case = TRUE)] <- "Highway"
ma_accidents_16_21_7$road_category[grepl("\\b(Pkwy|Parkway)\\b", col_to_check, ignore.case = TRUE)] <- "Parkway"
ma_accidents_16_21_7$road_category[grepl("\\b(Brg|Bridge)\\b", col_to_check, ignore.case = TRUE)] <- "Bridge"
ma_accidents_16_21_7$road_category[grepl("\\b(Cir|Circle)\\b", col_to_check, ignore.case = TRUE)] <- "Circle"
ma_accidents_16_21_7$road_category[grepl("\\b(Trl|Trail)\\b", col_to_check, ignore.case = TRUE)] <- "Trail"
ma_accidents_16_21_7$road_category[grepl("\\b(Ct|Cor|Court)\\b", col_to_check, ignore.case = TRUE)] <- "Court"
ma_accidents_16_21_7$road_category[grepl("\\b(Ter|Terrace)\\b", col_to_check, ignore.case = TRUE)] <- "Terrace"
ma_accidents_16_21_7$road_category[grepl("\\b(Pl|Place)\\b", col_to_check, ignore.case = TRUE)] <- "Place"
ma_accidents_16_21_7$road_category[grepl("\\b(Tpke|Turnpike)\\b", col_to_check, ignore.case = TRUE)] <- "Turnpike"
ma_accidents_16_21_7$road_category[grepl("\\b(Sq|Square)\\b", col_to_check, ignore.case = TRUE)] <- "Square"
ma_accidents_16_21_7$road_category[grepl("\\b(Pth|Path)\\b", col_to_check, ignore.case = TRUE)] <- "Path"
ma_accidents_16_21_7$road_category[grepl("\\b(Conn|Connector)\\b", col_to_check, ignore.case = TRUE)] <- "Connector"
ma_accidents_16_21_7$road_category[grepl("\\b(Ext|Exit)\\b", col_to_check, ignore.case = TRUE)] <- "Exit"
ma_accidents_16_21_7$road_category[grepl("\\b(Rty|Rotary)\\b", col_to_check, ignore.case = TRUE)] <- "Rotary"
ma_accidents_16_21_7$road_category[grepl("\\b(Expy|Expressway)\\b", col_to_check, ignore.case = TRUE)] <- "Expressway"
ma_accidents_16_21_7$road_category[grepl("\\b(Tunl|Tunnel)\\b", col_to_check, ignore.case = TRUE)] <- "Tunnel"


 ```

 

 ```{r}
 
ma_accidents_16_21_6 |> count(Weather_Condition)


 ```

 ```{r}
 

ma_accidents_16_21_6 |> select(Weather_Condition, City) |> 
    filter(Weather_Condition == "") |> 
    head(15)


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 
ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    mutate(Weather_Condition = as.character(Weather_Condition))


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_7$Weather_Condition[trimws(ma_accidents_16_21_7$Weather_Condition) == ""] <- "NR"


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    mutate(Weather_Condition = as.factor(Weather_Condition))


 ```

 ```{r}
 #| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    mutate(Weather_Condition = recode(Weather_Condition,  "NR" = "Fair Partley Cloudy", "Light Rain / Windy" = "Light Rain", "Light Rain Shower" = "Light Rain", "Light Rain with Thunder" = "Light Rain", "Light Drizzle" = "Light Rain","Rain / Windy" = "Rain", "Thunderstorms and Rain" = "Rain", "Thunderstorm" = "Rain", "T-Storm" = "Rain","Thunder" = "Overcast", "Cloudy" = "Overcast", "Thunder / Windy" = "Overcast", "Thunder in the Vicinity" = "Overcast","Cloudy / Windy" = "Overcast", "Mostly Cloudy" = "Overcast", "Mostly Cloudy / Windy" = "Overcast","Light Thunderstorms and Rain" = "Rain","Heavy Rain / Windy" = "Heavy Rain", "Heavy Thunderstorms and Rain" = "Heavy Rain","Heavy T-Storm / Windy" = "Heavy Rain","Heavy T-Storm" = "Heavy Rain", "Light Snow / Windy" = "Light Snow", "Light Snow and Sleet" = "Light Snow", "Clear" = "Fair Partley Cloudy","Blowing Snow" = "Heavy Snow", "Heavy Snow / Windy" = "Heavy Snow", "Light Drizzle / Windy" = "Light Rain", "Mist" = "Light Rain", "Drizzle and Fog" = "Light Rain", "Fair / Windy" = "Fair Partley Cloudy", "Partly Cloudy" = "Fair Partley Cloudy", "Partly Cloudy / Windy" = "Fair Partley Cloudy", "Fair" = "Fair Partley Cloudy", "Scattered Clouds" = "Fair Partley Cloudy",
   "Fog / Windy" = "Fog", "Patches of Fog" = "Fog","Shallow Fog" = "Fog","Wintry Mix / Windy" = "Wintry Mix", "Light Freezing Drizzle" = "Wintry Mix" , "Light Freezing Rain" = "Wintry Mix","Smoke" = "Haze", "Squalls" = "Snow",
   "Light Freezing Fog" = "Wintry Mix", "Ice Pellets" = "Wintry Mix","Small Hail" = "Wintry Mix",  "Snow / Windy" = "Snow", "N/A Precipitation" = "Fair Partley Cloudy") )


 ```




   Now we can perform our binning.  Essentiall we will place hour of day into eight ranges.




 ```{r}
 



ma_accidents_16_21_7 <- ma_accidents_16_21_7|> 
    mutate(hour_day_sections = case_when(
        between(hour_of_day,5,6) ~ "Dawn",
        between(hour_of_day,6,9) ~ "Early Morning",
        between(hour_of_day,9,12) ~ "Late Morning",
        between(hour_of_day,12,16) ~ "Early Afternoon",
        between(hour_of_day,16,18) ~ "Late Afternoon",
        between(hour_of_day,18,21) ~ "Evening",
        between(hour_of_day,21,24) ~ "Night",
        TRUE ~ "Middle of Night"

    ))



 ```

 ```{r}
 


ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    mutate(hour_day_sections = as.factor(hour_day_sections))


 ```

 ```{r}
 

ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    select(-hour_of_day)


 ```

 ```{r}
#| echo: false
#| include: false
#| eval: false
 

saveRDS(ma_accidents_16_21_7, "ma_accidents_16_21_7.rds")



 ```


 ```{r}
#| echo: false
 

ma_accidents_16_21_7 <- readRDS("ma_accidents_16_21_7.rds")


 ```



 ```{r}
 





 ```



 ```{r}
 

ma_accidents_16_21_7 |> 
    count(road_category)


 ```

 ```{r}
 

ma_accidents_16_21_7 |> 
    count(Weather_Condition)



 ```

 ```{r}
 

ma_accidents_16_21_7 |> 
    select_if(is.factor) |> 
    summary()

 ```


 ```{r}
 

ma_accidents_16_21_7 |> 
    count(No_Exit) |> 
    mutate(perc = round(n/sum(n),4)) |> 
    arrange(desc(perc))


 ```

 ```{r}
 
ma_accidents_16_21_7 |> 
    count(County, No_Exit)

 ```


```{r}


ma_accidents_16_21_7|> 
    group_by(No_Exit ) |> 
    summarise(median_acc_dur = median(accident_duration))

```

 ```{r}
 

ma_accidents_16_21_7 |> 
    count(Crossing) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
 
ma_accidents_16_21_7 |> 
    count(Railway) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))



 ```

 ```{r}
 


 ma_accidents_16_21_7 |> 
    count(County, Railway)

 ```

 ```{r}
 
ma_accidents_16_21_7 |> 
    count(Give_Way) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
 


ma_accidents_16_21_7 |> 
    count(County, Give_Way)


 ```

 ```{r}
 

ma_accidents_16_21_7 |> 
    count(Amenity) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
 
ma_accidents_16_21_7 |> 
    count(County, Amenity)


 ```

 ```{r}
 

ma_accidents_16_21_7 |> 
    count(Junction) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
 

ma_accidents_16_21_7 |> 
    count(County, Junction)


 ```

 ```{r}
 

ma_accidents_16_21_7 |> 
    count(Station) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))

 ```

 ```{r}
 


ma_accidents_16_21_7 |> 
    count(County, Station)

 ```

 ```{r}
 

ma_accidents_16_21_7 |> 
    count(Stop) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
 

 ma_accidents_16_21_7 |> 
    count(County, Stop)
 ```

 ```{r}
 

ma_accidents_16_21_7 |> 
    count(Traffic_Calming) |> 
    mutate(perc = round(n/sum(n),3)) |> 
    arrange(desc(perc))

 ```

 ```{r}
 

ma_accidents_16_21_7|> 
    group_by(Traffic_Calming) |> 
    summarise(median_acc_dur = median(accident_duration))



 ```

 ```{r}
 

ma_accidents_16_21_7 |> 
    count(Stop) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
 

ma_accidents_16_21_7 |> 
    count(Traffic_Signal) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
 

 ma_accidents_16_21_7 |> 
    count(County, Traffic_Signal)


 ```


 ```{r}
 


ma_accidents_16_21_7 |> 
    count(Wind_Direction)




 ```



 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_8 <- ma_accidents_16_21_7 |> 
    select(-No_Exit, -Traffic_Calming, -Nautical_Twilight, -Astronomical_Twilight, -Give_Way, -Railway, -Description, -Street, -Bump, -Roundabout, -Turning_Loop)


 ```

 ```{r}
#| echo: false
#| include: false
#| eval: false
 
saveRDS(ma_accidents_16_21_8, "ma_accidents_16_21_8.rds")


 ```

 ```{r}
 #| echo: false
 


ma_accidents_16_21_8 <- readRDS("ma_accidents_16_21_8.rds")

 ```

 ```{r}
 

ma_accidents_16_21_8 |> 
    glimpse()


 ```

 

 <br><br>

 

  <br><br>


 ```{r}
 
class(ma_accidents_16_21_9$hour_of_day)


 ```

  <br><br>

 



  <br><br>




Main Sections of the DayMorning (5:00 AM – 12:00 PM):Dawn / Sunrise: First light and when the sun comes up (approx. 5:00 AM – 6:00 AM).Early Morning: The start of typical morning routines (approx. 6:00 AM – 9:00 AM).Late Morning: Approaching midday (approx. 10:00 AM – 12:00 PM).Midday / Noon (12:00 PM): The exact middle of the day when the sun is highest.Afternoon (12:00 PM – 5:00 PM / 6:00 PM):Early Afternoon: Right after lunch (1:00 PM – 3:00 PM).Late Afternoon: Leading toward sunset (4:00 PM – 6:00 PM).Evening (5:00 PM / 6:00 PM – 9:00 PM):Sunset / Dusk: The sun goes down and twilight begins.Early/Late Evening: Transition time for relaxation and dinner.Night (9:00 PM – 4:00 AM / 5:00 AM):Late Night: Hours leading up to midnight (10:00 PM – 12:00 AM).Midnight: 12:00 AM, the middle of the night.Predawn / Middle of the Night: The quiet hours before the new day starts (1:00 AM – 4:00 AM).

 <br><br>




  <br><br>


  ```{r}
  
saveRDS(ma_accidents_16_21_9, "ma_accidents_16_21_9.rds")


  ```

  ```{r}
  

ma_accidents_16_21_9 <- readRDS("ma_accidents_16_21_9.rds")


  ```


  <br><br>



```{r}


ma_accidents_16_21_9 |> 
    count(hour_day_sections) |> 
    mutate(perc = round(n/sum(n),3)*100) |> 
    arrange(desc(perc))


```

  <br><br>


We now have eight categories of different parts of the day. 

  <br><br>




```{r}





```

```{r}



ma_accidents_16_21_3  <- ma_accidents_16_21_2  |> 
    mutate(id = row_number())


```





# Census

## County




```{r}


ma_county <- counties("MA", year = 2024)

```

```{r}

saveRDS(ma_county, "ma_county.rds")


```

```{r}


ma_county <- readRDS("ma_county.rds")


```

```{r}


ma_county |> 
    glimpse()


```

```{r}


ma_county |> 
    summary()


```

```{r}


print(ma_county |> select(NAME), n=14)


```


```{r}


ma_county_shape <- ma_county |> 
    filter(NAME != "Nantucket") |> 
    filter(NAME != "Dukes")


```

```{r}

saveRDS(ma_county_shape , "ma_county_shape.rds")


```


```{r}


ma_county_shape  <- readRDS("ma_county_shape.rds")



```

```{r}



print(ma_county_shape |> select(NAME), n=14) 


```

```{r}



ma_county_shape_2 <- ma_county_shape |> 
    select(COUNTYFP, GEOID, NAME, NAMELSAD, INTPTLAT, INTPTLON, geometry)


```

```{r}

ma_county_shape_2 <- ma_county_shape_2 |> 
    rename("GEOID_COUNTY" = "GEOID")


```

```{r}



saveRDS(ma_county_shape_2, "ma_county_shape_2.rds")

```

```{r}



ma_county_shape_2 <- readRDS("ma_county_shape_2.rds")


```

```{r}


ma_county_shape_2 |> glimpse()



```

```{r}





```


```{r}


census_county_urban <- readxl::read_xlsx("/Users/francislauder/Documents/census/2020_UA_COUNTY.xlsx")


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(census_county_urban, "census_county_urban.rds")


```

```{r}
#| echo: false

census_county_urban <- readRDS("census_county_urban.rds")


```


```{r}


census_county_urban |> glimpse()


```

```{r}


census_county_urban |> 
    count(STATE)


```


```{r}


ma_county_urban <- census_county_urban |> 
    filter(STATE_NAME == "Massachusetts")


```


```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_county_urban , "ma_county_urban.rds")


```

```{r}
#| echo: false


ma_county_urban  <- readRDS("ma_county_urban.rds")

```

```{r}


ma_county_urban |> summary()


```

```{r}


print(ma_county_urban |> select(COUNTY_NAME), n=14 )


```

```{r}


ma_county_urban_2 <- ma_county_urban |> 
    filter(COUNTY_NAME != "Dukes") |> 
    filter(COUNTY_NAME != "Nantucket")


```

```{r}


ma_county_urban_2 <- ma_county_urban_2 |> 
    select(COUNTY, POP_COU, HOU_COU, `ALAND_Mi²_COU`, POP_URB, POPPCT_URB, HOU_URB, HOUPCT_URB, `ALAND_Mi²_URB`, ALAND_PCT_URB, POP_RUR, POPPCT_RUR, HOU_URB, HOUPCT_RUR, `ALAND_Mi²_RUR`, ALAND_PCT_RUR)



```

```{r}


ma_county_urban_2 <- ma_county_urban_2 |> 
    rename("ALAND_MI_COU" = "ALAND_Mi²_COU", "ALAND_MI_URB" = "ALAND_Mi²_URB", "ALAND_MI_RUR" = "ALAND_Mi²_RUR", "COUNTYFP" = "COUNTY")

```

```{r}


saveRDS(ma_county_urban_2, "ma_county_urban_2.rds")



```

```{r}


ma_county_urban_2 <- readRDS("ma_county_urban_2.rds")


```

```{r}


ma_county_urban_2 |> glimpse()


```

```{r}



print(ma_county_urban_2 |> select(COUNTYFP), n=14)




```

```{r}



ma_county_urban_3 <- ma_county_urban_2 |> 
    mutate(county_pop_density = round(POP_COU / ALAND_MI_COU,0), county_house_density = round(HOU_COU / ALAND_MI_COU,0))


```

```{r}

saveRDS(ma_county_urban_3, "ma_county_urban_3.rds")


```

```{r}


ma_county_urban_3 <- readRDS("ma_county_urban_3.rds")


```

```{r}


print(ma_county_urban_3 |> select(county_house_density, county_pop_density),n=14)



```



```{r}


county_geo <- ma_county_shape_2 |> 
    full_join(ma_county_urban_3, by="COUNTYFP")



```

```{r}


saveRDS(county_geo , "county_geo.rds")


```

```{r}



county_geo  <- readRDS("county_geo.rds")



```

```{r}



county_geo |> glimpse()


```

```{r}


county_geo_2 <- county_geo |> 
    select(COUNTYFP, NAME, POP_COU, HOU_COU, POPPCT_URB, HOU_URB, county_pop_density) |> 
    st_drop_geometry()


```

```{r}

county_geo_2 <- county_geo_2 |> 
    rename("County" = "NAME", "Cou_POPPCT_URB" = "POPPCT_URB", "COU_HOU_URB" = "HOU_URB")


```

```{r}


county_geo_2 |> glimpse()

```

```{r}


saveRDS(county_geo_2, "county_geo_2.rds")

```

```{r}


county_geo_2 <- readRDS("county_geo_2.rds")


```



```{r}


ma_county_urban |> 
    select(COUNTY_NAME, POPDEN_COU, POPPCT_URB, HOUDEN_COU, HOUDEN_URB,HOUDEN_RUR, HOUPCT_URB)


```

## Tracts

```{r}


ma_tracts <- tracts("MA", year = 2024)


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tracts, "ma_tracts.rds")


```

```{r}

ma_tracts <- readRDS("ma_tracts.rds")


```

```{r}

ma_tracts


```


Define the census tract variables we want to retrieve from the acs. 

```{r}


variables_to_get <- c(
    total_tract_housing_units = "B25001_001",
    total_vehicles_tract = "B25046_001",
    total_households_tract = "B25044_001",
    total_population_tract = "B01003_001"


)


```



```{r}


ma_tract_data <- get_acs(
    geography = "tract",
    variables = variables_to_get,
    state = "MA",
    geometry = TRUE,
    output = "wide",
    year = 2021
)



```

```{r}


ma_tract_data |> 
    glimpse()


```

```{r}


ma_tract_data |> summary()


```



```{r}


ma_tract_geometry <- ma_tract_data |> 
    select(GEOID, geometry)


```


```{r}


saveRDS(ma_tract_geometry, "ma_tract_geometry.rds")

```

```{r}


ma_tract_geometry <- readRDS("ma_tract_geometry.rds")


```

```{r}



ma_tract_cls <- ma_tract_geometry |> 
    st_transform(5070) 


```

```{r}

saveRDS(ma_tract_cls, "ma_tract_cls.rds")


```

```{r}


ma_tract_cls <- readRDS("ma_tract_cls.rds")


```

```{r}


ma_tract_data |> 
    select(NAME) |> 
    head(10)



```



The output of the NAME column shows it contains Census Tract, Coumty, and State. Will will use the separate function to split the NAME column into three new columns.



```{r}

ma_tract_data_2 <- ma_tract_data |> 
    separate(NAME, into=c("census_tract", "County", "state"), sep=", ") |> 
    st_drop_geometry()



```

Remove State

```{r}


ma_tract_data_2 <- ma_tract_data_2 |> 
    select(-state)


```


```{r}


saveRDS(ma_tract_data_2 , "ma_tract_data_2.rds")

```


```{r}

ma_tract_data_2  <- readRDS("ma_tract_data_2.rds")


```

```{r}


ma_tract_data_3 <- ma_tract_geometry |> 
    full_join(ma_tract_data_2, by = "GEOID")


```

```{r}
saveRDS(ma_tract_data_3 , "ma_tract_data_3.rds")

```

```{r}


ma_tract_data_3  <- readRDS("ma_tract_data_3.rds")


```





```{r}



ma_tract_data_3 |> summary()




```


```{r}


ma_tract_data_4 <- ma_tract_data_3 |> 
    filter(GEOID != "25001990000") |> 
    filter(GEOID != "25023990003")



```


```{r}


ma_tract_data_4 |> 
    filter(GEOID == "25001990000" | GEOID == "25023990003") |> 
    head()


```

```{r}

ma_tract_data_4 |> 
    filter(GEOID == "25001990000") |> 
    head()


```

```{r}


ma_tract_data_3 |> 
    select(GEOID, total_vehicles_tractE) |> 
    filter(is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summary()



```


```{r}




ma_tract_data_3 |> 
    select(GEOID, census_tract,County, total_vehicles_tractE) |> 
    filter(is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    head(72)

```


```{r}





ma_tract_data_3 |> 
    filter(County == "Suffolk County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))




```

```{r}


ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Suffolk County" & is.na(total_vehicles_tractE), 1310,total_vehicles_tractE ))




```



```{r}



ma_tract_data_3 |> 
    filter(County == "Bristol County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```

```{r}



ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Bristol County" & is.na(total_vehicles_tractE), 2785,total_vehicles_tractE ))



```

```{r}



ma_tract_data_3 |> 
    filter(County == "Hampshire County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))



```

```{r}



ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Hampshire County" & is.na(total_vehicles_tractE), 3340,total_vehicles_tractE ))


```

```{r}


ma_tract_data_3 |> 
    filter(County == "Essex County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```

```{r}



ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Essex County" & is.na(total_vehicles_tractE), 2989,total_vehicles_tractE ))


```

```{r}


ma_tract_data_3 |> 
    filter(County == "Middlesex County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```

```{r}



ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Middlesex County" & is.na(total_vehicles_tractE), 2828,total_vehicles_tractE))
```


```{r}

ma_tract_data_3 |> 
    filter(County == "Plymouth County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))

```

```{r}



ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Plymouth County" & is.na(total_vehicles_tractE), 3471,total_vehicles_tractE))



```

```{r}



ma_tract_data_3 |> 
    filter(County == "Worcester County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```

```{r}


ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Worcester County" & is.na(total_vehicles_tractE), 2898,total_vehicles_tractE))



```

```{r}


ma_tract_data_3 |> 
    filter(County == "Barnstable County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))

```

```{r}


ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Barnstable County" & is.na(total_vehicles_tractE), 3166,total_vehicles_tractE))


```

```{r}

ma_tract_data_3 |> 
    filter(County == "Hampden County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```

```{r}


ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Hampden County" & is.na(total_vehicles_tractE), 2736,total_vehicles_tractE))


```

```{r}


ma_tract_data_3 |> 
    filter(County == "Berkshire County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```

```{r}


ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Berkshire County" & is.na(total_vehicles_tractE), 2248,total_vehicles_tractE))


```

```{r}

ma_tract_data_3 |> 
    filter(County == "Norfolk County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```





```{r}


ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Norfolk County" & is.na(total_vehicles_tractE), 3080,total_vehicles_tractE))



```



```{r}





ma_tract_data_3 |> 
    select(GEOID, census_tract,County, total_vehicles_tractE) |> 
    filter(is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    head(72)


```

```{r}


saveRDS(ma_tract_data_4, "ma_tract_data_4.rds")

```

```{r}


ma_tract_data_4 <- readRDS("ma_tract_data_4.rds")



```

```{r}


ma_tract_data_4 |> 
    summary()


```


```{r}


ma_tract_data_5 <- ma_tract_data_4 |> 
    filter(total_population_tractE != 0)


```


```{r}


ma_tract_data_5 <- ma_tract_data_5 |> 
    filter(County != "Nantucket County") |> 
    filter(County != "Dukes County")


```

```{r}


ma_tract_data_5 <- ma_tract_data_5 |> 
    filter(total_households_tractE != 0)



```



```{r}


saveRDS(ma_tract_data_5, "ma_tract_data_5.rds")


```

```{r}


ma_tract_data_5 <- readRDS("ma_tract_data_5.rds")



```

```{r}

ma_tract_data_5 |> count(County)


```

```{r}

ma_tract_data_5 |> 
    filter(is.na(total_vehicles_tractE)) |> 
    count(County)


```



```{r}


ma_tract_data_5 |> 
    summary()


```




### Feature Engineering



From our retrieved census tract variables, we will create new variables for housing density of census tract, population density of census tract, and average cars per houshold of census tract. 







```{r}
ma_tract_data_6 <- ma_tract_data_5 |> 
    mutate(
        tract_house_density_sqmi = round(as.numeric(set_units(total_tract_housing_unitsE/ st_area(geometry),"1/mi2")),2),
        tract_pop_density_sqmi = round(as.numeric(set_units(total_population_tractE/ st_area(geometry),"1/mi2")),2),
        avg_cars_per_household = round(total_vehicles_tractE / total_households_tractE, 2)  
    #select(!ends_with("M"))
    #rename_with(.fn = ~str_remove(.x, "E$"))


    )



```



```{r}



print(ma_tract_data_5 |> 
    filter(total_households_tractE == 0), n=70)


```






```{r}


ma_tract_data_6 <- ma_tract_data_6 |> 
    select(-total_tract_housing_unitsM, -total_vehicles_tractM, -total_households_tractM, -total_population_tractM)


```

```{r}


ma_tract_data_6 <- ma_tract_data_6 |> 
    select(-County)


```


```{r}

saveRDS(ma_tract_data_6 , "ma_tract_data_6.rds")


```

```{r}


ma_tract_data_6 <- readRDS("ma_tract_data_6.rds")


```



```{r}




ma_tract_data_6|> glimpse()



```

```{r}


ma_tract_data_6 |> summary()


```

```{r}


ma_tract_data_7 <- ma_tract_data_6 |> 
    select(-census_tract, -County) |> 
    st_drop_geometry()




```

```{r}


saveRDS(ma_tract_data_7, "ma_tract_data_7.rds")


```

```{r}

ma_tract_data_7 <- readRDS("ma_tract_data_7.rds")


```


```{r}


ma_tract_data_7|> 
    glimpse()


```






## Joining


# Exploratory Data Analysis


## Features


```{r}


ma_accidents_16_21_6 |> 
    ggplot(aes(x=accident_duration, y=Temperature.F.)) +
    geom_point()



```

```{r}



ma_accidents_16_21_6 |> 
    ggplot(aes(x=log(accident_duration), y=Temperature.F.)) +
    geom_point()


```

## Maps

### County


```{r}


county_map <- ggplot()+
    geom_sf(data=ma_county, fill=NA)+
    ggtitle("Counties of Massachusetts") +
    theme(legend.position = "none",plot.title = element_text(color="blue", size=14, face="bold.italic", hjust=0.5)) +
    theme_bw()


```


```{r}


ma_county_geo_4326 <- county_geo |> 
    st_transform(crs=4326)


```

```{r}


ma_county_stcrs <- st_crs(ma_county_geo_4326)


```


```{r}


ma_county_stcrs$input



```









```{r}


bins<-c(0,71029,129026,162308,228996,465825,530819,579200,725981,797936,809829,862111,Inf)


```

```{r}



pal<-colorBin(palette="RdYlBu",
            domain=ma_county_geo_4326$POP_COU,bins=bins)
```

```{r}



ma_county_formated <- ma_county_geo_4326 |> 
    mutate(county_pop_density = format(county_pop_density, big.mark = ",", big.interval = 3, digits=12) ) |> 
    mutate(POPPCT_URB = round(POPPCT_URB * 100,2))


```

```{r}




mytext <- paste(
  "County: ", ma_county_formated$NAME,"<br/>",
  "Population Density: ", ma_county_formated$county_pop_density, "<br/>",
  "County Population Percent Urban: ", ma_county_formated$POPPCT_URB,"<br/>", 
    sep="") %>%
  lapply(htmltools::HTML)


```



```{r}


map_county_pop_density_plt<- ma_county_geo_4326 |> leaflet(options = leafletOptions(minZoom = 9.4, maxZoom = 18)) |> 
  addProviderTiles(providers$Esri.WorldStreetMap) |> 
  setView(lng = -071.9402820, lat = 42.3116934, zoom = 9.4) |> 
  addPolygons(
 fillColor=~pal(POP_COU),
 weight=2,
 opacity=0.3,
 color= "white",
 label=~mytext,
 dashArray = "3",
 fillOpacity=0.3,
 highlightOptions = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.5,
    bringToFront = FALSE)) 
    
  




```



```{r}

county_map_pop_density<-map_county_pop_density_plt %>% 
  addLegend(pal = pal, values = ~county_pop_density, opacity = 0.6, title = "Massachusetts County Population Density",
  position = "bottomright")


```

```{r}

map_county_pop_density_plt



```



### Census Tracts


```{r}



ma_tracts_plot <- ggplot(ma_tracts_2) +
    geom_sf()+
    ggtitle("Census Tracts of Massachusetts") +
    theme(legend.position = "none",plot.title = element_text(color="blue", size=14, face="bold.italic", hjust=0.5)) +
    theme_void()



```

```{r}

ma_tracts_plot


```

```{r}

tract_accident_count <- ma_accidents_16_21_8 |> 
    count(GEOID, name = "tract_accident_count")


```

```{r}

saveRDS(tract_accident_count, "tract_accident_count.rds")

```

```{r}


tract_accident_count <- readRDS("tract_accident_count.rds")



```

```{r}



tract_accident_count |> 
    slice_max(tract_accident_count, n=20) 

```

```{r}

ma_accidents_16_21_8 |> 
    filter(GEOID == "25017334100")



```

```{r}


ma_accidents_16_21_8 |> 
    filter(accident_duration >=30 & accident_duration <=60)

```


```{r}


ma_accidents_16_21_10 <- ma_accidents_16_21_8 |> 
    full_join(tract_accident_count, by = "GEOID")



```

```{r}


saveRDS(ma_accidents_16_21_10, "ma_accidents_16_21_10.rds")


```

```{r}



ma_accidents_16_21_10 <- readRDS("ma_accidents_16_21_10.rds")


```

 ```{r}
 
ma_accidents_16_21_10 |> 
 count(GEOID) |> 
 arrange(desc(n))


 ```

 ```{r}
 



 ```

 ```{r}
 

ma_accidents_16_21_10 |> 
    summary()



 ```

```{r}



ma_tract_accident <- ma_tract_data_6 |> 
    full_join(ma_accidents_16_21_10, by = "GEOID")


```

```{r}



saveRDS(ma_tract_accident, "ma_tract_accident.rds")


```

```{r}


ma_tract_accident <- readRDS("ma_tract_accident.rds")


```

```{r}



ma_tract_accident |> 
    summary()



```

```{r}


ma_tract_accident |> 
    filter(is.na(total_households_tractE)) |> 
    summary()


```

```{r}




ma_tract_accident |> 
    filter(is.na(total_households_tractE)) |> 
    count(GEOID)


```

```{r}

ma_tract_accident |> 
    filter(is.na(total_households_tractE)) |> 
    head(1000)



```

```{r}


ma_tract_accident_2 <- ma_tract_accident |> 
    mutate(total_households_tractE = replace_na(total_households_tractE, 0), 
           total_population_tractE = replace_na(total_population_tractE, 0),
           tract_house_density_sqmi = replace_na(tract_house_density_sqmi,0),
           tract_pop_density_sqmi = replace_na(tract_pop_density_sqmi,0),
           total_vehicles_tractE = replace_na(total_vehicles_tractE,0),
           total_tract_housing_unitsE = replace_na(total_tract_housing_unitsE,0) )


```

```{r}

saveRDS(ma_tract_accident_2, "ma_tract_accident_2.rds")

```

```{r}


ma_tract_accident_2 <- readRDS("ma_tract_accident_2.rds")

```

```{r}

ma_tract_accident_2 |> 
    summary()

```

```{r}




ma_tract_accident_3 <- ma_tract_accident_2 |> 
    filter(!is.na(day_of_week))
```

```{r}

saveRDS(ma_tract_accident_3, "ma_tract_accident_3.rds")

```

```{r}

ma_tract_accident_3 <- readRDS("ma_tract_accident_3.rds")


```

```{r}

ma_tract_accident_3 |> 
    summary()


```

```{r}

ma_tract_cls <- st_crs(ma_tract_accident_3)


```

```{r}




```


```{r}

ma_census4326 <- ma_tract_accident_3 |> 
    st_transform(crs=4326)


```

```{r}


saveRDS(ma_census4326, "ma_census4326.rds")


```

```{r}



ma_census4326 <- readRDS("ma_census4326.rds")


```

```{r}


ma_tract_cls_2 <- st_crs(ma_census4326)


```

```{r}

ma_tract_cls_2$input


```

```{r}

bins<-c(0,500,2000,3000,4000,5000,6500,8500,9500,10500,Inf)


```

```{r}


pal<-colorBin(palette="RdYlBu",
            domain=ma_census4326$total_population_tractE,bins=bins)



```

```{r}

       
       
ma_tract_formated <- ma_tract_accident_3 |> 
    mutate(total_tract_housing_unitsE = format(total_tract_housing_unitsE, big.mark = ",", big.interval = 3, digits=12) ) |> 
    mutate(total_vehicles_tractE = format(total_vehicles_tractE, big.mark = ",", big.interval = 3, digits=12) ) 
   


```

```{r}


mytext <- paste(
  "County: ", ma_tract_formated$County,"<br/>",
  "Census Tract: ", ma_tract_formated$GEOID, "<br/>",
  "Total Vehicles Census Tract: ", ma_tract_formated$total_vehicles_tractE, "<br/>",
  "Total Housing Census Tract: ", ma_tract_formated$total_tract_housing_unitsE,"<br/>", 
  "Census Tract Accident County: ", ma_tract_formated$tract_accident_count,"<br/>",
    sep="") %>%
  lapply(htmltools::HTML)


```

```{r}



map_tract_population_plt<- ma_census4326 |> leaflet(options = leafletOptions(minZoom = 9.4, maxZoom = 18)) |> 
  addProviderTiles(providers$Esri.WorldStreetMap) |> 
  setView(lng = -071.9402820, lat = 42.3116934, zoom = 9.4) |> 
  addPolygons(
 fillColor=~pal(total_population_tractE),
 weight=2,
 opacity=0.8,
 color= "white",
 label=~mytext,
 dashArray = "3",
 fillOpacity=0.3,
 highlightOptions = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.5,
    bringToFront = FALSE)) 


```

```{r}



tract_map_population<-map_tract_population_plt %>% 
  addLegend(pal = pal, values = ~total_population_tractE, opacity = 0.6, title = "Massachusetts Census Tract Population",
  position = "bottomright")

```

```{r}


tract_map_population
  

```


```{r}


ma_census4326 |> 
    filter(GEOID == "25017368102")


```

```{r}


ma_census_middle <- ma_census4326 |> 
    filter(County == "Middlesex")



```

```{r}


saveRDS(ma_census_middle, "ma_census_middle.rds")

```

```{r}


ma_census_middle <- readRDS("ma_census_middle.rds")



```

```{r}


pal_2<-colorBin(palette="RdYlBu",
            domain=ma_census_middle$total_population_tractE,bins=bins)


```

```{r}


ma_tract_middle_formated <- ma_census_middle |> 
    mutate(total_tract_housing_unitsE = format(total_tract_housing_unitsE, big.mark = ",", big.interval = 3, digits=12) ) |> 
    mutate(total_vehicles_tractE = format(total_vehicles_tractE, big.mark = ",", big.interval = 3, digits=12) ) 


```

```{r}

saveRDS(ma_tract_middle_formated, "ma_tract_middle_formated.rds")

```

```{r}

ma_tract_middle_formated <- readRDS("ma_tract_middle_formated.rds")


```

```{r}



mytext_2 <- paste(
  "County: ", ma_tract_middle_formated$County,"<br/>",
  "Census Tract: ", ma_tract_middle_formated$GEOID, "<br/>",
  "Total Vehicles Census Tract: ", ma_tract_middle_formated$total_vehicles_tractE, "<br/>",
  "Total Housing Census Tract: ", ma_tract_middle_formated$total_tract_housing_unitsE,"<br/>", 
  "Census Tract Accident County: ", ma_tract_middle_formated$tract_accident_count,"<br/>",
    sep="") %>%
  lapply(htmltools::HTML)


```

```{r}



map_middle_tract_population_plt<- ma_census_middle |> leaflet(options = leafletOptions(minZoom = 11, maxZoom = 18)) |> 
  addProviderTiles(providers$Esri.WorldStreetMap) |> 
  setView(lng = -071.23928, lat = 42.38980, zoom = 11.5) |> 
  addPolygons(
 fillColor=~pal_2(total_population_tractE),
 weight=2,
 opacity=0.8,
 color= "white",
 label=~mytext_2,
 dashArray = "3",
 fillOpacity=0.3,
 highlightOptions = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.5,
    bringToFront = FALSE)) 




```

```{r}



tract_middle_map_population<-map_middle_tract_population_plt %>% 
  addLegend(pal = pal_2, values = ~total_population_tractE, opacity = 0.6, title = "Middlesex County  Census Tract Population",
  position = "bottomright")


```

```{r}


tract_middle_map_population


```



# Preprocessing


```{r}



```


#rows / columns 
10:1 rule 
You should have at least 10 rows of data for every 1 column (feature) in your dataset. 

`

```{r}


census_tract_df <- ma_accidents_16_21_8 |> 
    left_join(ma_tract_data_7, by = "GEOID")

```

```{r}



census_tract_df |> 
   summary()


```

```{r}

census_tract_df  |> filter(is.na(total_population_tractE))  |> count(GEOID)


```

```{r}

census_tract_df |> filter(total_tract_housing_unitsE == 0) |> head(10)


```

```{r}


census_tract_county_df <- census_tract_df |> 
    left_join(county_geo_2, by = "County")


```

```{r}


census_tract_df_2 <- census_tract_df |> 
    select(-ID, -Severity, -City, -County, -Year, -LATITUDE, -LONGITUDE, -census_code, -tract_house_density_sqmi, -Stop,-Amenity, -Crossing,-Junction ,-Station,-Traffic_Signal,-Sunrise_Sunset,-Civil_Twilight, -day_of_week, -total_tract_housing_unitsE, -total_vehicles_tractE, -total_population_tractE, -avg_cars_per_household, -total_households_tractE, -tract_pop_density_sqmi)


```




```{r}


census_tract_df_2 |> summary()

```

```{r}

census_tract_df_2 |> count(Weather_Condition)


```

```{r}


census_tract_df_2 |> 
    glimpse()


```

```{r}



census_tract_df |> select(GEOID, County, total_households_tractE, total_population_tractE, total_vehicles_tractE) |> filter(is.na(total_households_tractE)) |> head()

```

```{r}

census_tract_df_2 |> 
    summary()


```

```{r}


census_tract_df_2 <- census_tract_df_2 |> 
    filter(!is.na(total_vehicles_tractE))


```

```{r}

census_tract_df_2 <- census_tract_df_2 |> 
    filter(total_tract_housing_unitsE != 0)


```


```{r}


census_tract_county_df_2 <- census_tract_county_df |> 
    select(-ID, -Severity, -City, -County, -Year, -LATITUDE, -LONGITUDE, -census_code, -tract_house_density_sqmi, -Stop,-Amenity, -Crossing,-Junction ,-Station,-Traffic_Signal,-Sunrise_Sunset,-Civil_Twilight, -day_of_week, -total_tract_housing_unitsE, -total_vehicles_tractE, -total_population_tractE, -avg_cars_per_household, -total_households_tractE, -tract_pop_density_sqmi, -County, 
    -POP_COU, -HOU_COU, -COU_HOU_URB)



```

```{r}


census_tract_county_df_2 |> summary()


```



```{r}



census_tract_df_2 |> 
    ggplot(aes(temperature)) +
    geom_histogram(bins=25)


```

```{r}


census_tract_df_2 |> 
    ggplot(aes(log(wind_speed))) +
    geom_histogram(bins=25)


```


```{r}


census_tract_df_2 |> 
    ggplot(aes(log(accident_duration),log(wind_speed)))) +
    geom_point()


```

```{r}



census_tract_df_2 |> 
    mutate(GEOID = as.factor(GEOID)) |> 
    count(GEOID) |>
    arrange(desc(n)) 



```


```{r}

census_tract_df_2 |> 
    summary()


```










```{r}

test_df <- test_df |> 
    rename(COUNTY_NAME = County)


```




```{r}

acci_census_df <- initial_split(census_tract_df_2, prop = 0.80)


```

```{r}


acci_count_tract_df <- initial_split(census_tract_county_df_2, prop = 0.80)


```


```{r}


print(acci_count_tract_df)


```


```{r}


train_census_data <- training(acci_census_df)


```

```{r}

test_census_data <- testing(acci_census_df)


```

```{r}


train_tract_cou <- training(acci_count_tract_df)

```

```{r}



test_tract_coun <- testing(acci_count_tract_df)


```





```{r}


census_rec <- recipe(accident_duration ~ ., data = train_census_data) |> 
    step_center(all_numeric_predictors()) |> 
    step_scale(all_numeric_predictors())


```

```{r}



census_cou_rec <- recipe(accident_duration ~ ., data = train_tract_cou) |> 
    step_center(all_numeric_predictors()) |> 
    step_scale(all_numeric_predictors())

```




```{r}

train_census_rec <- prep(census_rec, training= train_census_data)
scaled_census_train <- bake(train_census_rec, new_data = train_census_data)


```


```{r}


train_tract_cou_rec <- prep(census_cou_rec, training=train_tract_cou)
scaled_tract_cou_train <- bake(train_tract_cou_rec, new_data = train_tract_cou)


```

```{r}


model_lme_1 <- lmer(log(accident_duration) ~ . -GEOID + (1 | GEOID), data = scaled_census_train)



```

```{r}

model_lme_2 <- lmer(log(accident_duration) ~ . -GEOID -COUNTYFP + (1 | COUNTYFP/GEOID), data = scaled_tract_cou_train)


```

```{r}


summary(model_lme_2)

```

```{r}


summary(model_lme_1)


```





```{r}


stan_model_1 <- stan_lmer(
    log(accident_duration) ~ . -GEOID + (temperature | GEOID),
  data = scaled_census_train,
  chains = 4,
  iter = 8000,
  cores = 30, 
  seed = 42
)


```



```{r}



brms_model <- brm(
  log(accident_duration) ~ . -GEOID + (temperature | GEOID),
  data = scaled_census_train,
  chains = 4,
  iter = 10000,
  family = student(),
  cores = 30, 
  seed = 42
)



```


```{r}


summary(brms_model)

```

```{r}

saveRDS(brms_model, "brms_model.rds")


```


```{r}




brms_model_2 <- brm(
  log(accident_duration) ~ . -GEOID -COUNTYFP + (temperature | COUNTYFP),
  data = scaled_tract_cou_train,
  chains = 4,
  iter = 8000,
  family = student(),
  cores = 30, 
  seed = 42
)


```

```{r}


summary(brms_model_2)


```

```{r}



# Extract fixed-effect formula without random effects
ff <- nobars(formula(your_model)
X <- model.matrix(ff, data = your_data)
)
# Check for linear combinations
caret::findLinearCombos(X)


```

```{r}


library(caret)
```

```{r}

# Extract fixed-effect formula without random effects
ff <- nobars(formula(model_lme_1))
X <- model.matrix(ff, data = train_data) 


# Check for linear combinations
caret::findLinearCombos(X)



```

```{r}


summary(model_lme_1)



```

```{r}


icc(model_lme_1, nrep=50)

```


