---
title: "Predicting time of auto accidents"
author: Francis Lauder
date: 2026-09-10
format:
  html:
    toc: true
    toc-depth: 3
    toc_float: true
    number-sections: false
    toc-location: left
    embed-resources: true
---

# Introduction
<br><br>


# Data Dictionary

<br><br>


# Libraries

<br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-1"

library(dplyr)
library(ggplot2)
#library(plotly)
library(RColorBrewer)
library(forcats)
library(tidycensus)

library(sf)

library(jsonlite)

library(tigris)

library(stringr)

library(leaflet)
library(htmlwidgets)
library(htmltools)
library(gt)
library(DT)
library(GGally)
library(lme4)
library(lmerTest)
library(rstanarm)
library(brms)
library(broom)
library(lubridate)
library(tidymodels)
library(performance)
library(boot.pval)
library(bayesplot)
library(units)
library(tidybayes)
library(marginaleffects)
library(collapse)
library(ggfortify)
library(bayesplot)
library(tidyr)
library(tibble)
library(spdep)
library(matrix)






```

<br><br>


# Data Inspection and Cleaning

## US Accident Data


```{r}
#| echo: false
#| include: false
#| eval: false


US_accidents <- read.csv("/Users/francislauder/Documents/Data Sets/US Acciidents/US_Accidents_March23.csv")


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(US_accidents, "US_accidents.rds")

```

```{r}
#| echo: false
#| include: false
#| eval: false


US_accidents <- readRDS("US_accidents.rds")


```



```{r}
#| echo: false
#| include: false
#| eval: false



glimpse(US_accidents)


```



```{r}
#| echo: false
#| include: false
#| eval: false


MA_accidents <- US_accidents |> 
    filter(State == "MA")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-2"


MA_accidents <- MA_accidents %>% 
  mutate(Weather_Timestamp = as.Date(Weather_Timestamp))


```


```{r}
#| echo: false
#| include: false
#| eval: false
#| label: "ck-3"


saveRDS(MA_accidents, "MA_accidents.rds")



```

```{r}
#| include: false
#| label: "ck-4"


MA_accidents <- readRDS("MA_accidents.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-5"



options(browser="false", scipen = 999, max.print = 20000)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-6"


dim(MA_accidents)


```

<br><br>

Our dataset has 61,996 rows and 46 columns. 

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-7"



MA_accidents |> glimpse()


```

<br><br>

```{r}
#| echo: false
#| include: false
#| eval: false
#| label: "ck-8"


summary(MA_accidents)


```

<br><br>





```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-9"



MA_accidents |> 
    count(Severity) |> 
    mutate(perc = round(n / sum(n),3)) |> 
    arrange(desc(perc))


```

<br><br>

Glimpsing our dataset we find some features do not have the correct data types for use in our analysis.  We will transform them to the required data types. 



<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-10"


class(MA_accidents$Severity)


```
<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-11"




MA_accidents_2 <- MA_accidents |> 
    mutate(Severity = as.factor(Severity))


```


<br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-12"


class(MA_accidents_2$Severity)


```

<br><br>


 Our analysis we will require year, month, day, and hour of an accident.  Since our dataset does not have these features, we will create them from the Start_Time feature. 



<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-13"



MA_accidents_2 <- MA_accidents_2 |> 
    mutate(Year = year(Start_Time)) |> 
    mutate(Month = month(Start_Time)) |> 
    mutate(day_of_week = wday(Start_Time)) |>
    mutate(day_type = if_else(wday(Start_Time) %in% c(1,7), "Weekend", "Weekday")) |> 
    mutate(hour_of_day = hour(Start_Time))


```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-14"


MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day, day_type) |> 
    head()


```

Day of week starts with Sunday (1) and ends with Saturday (7)

Hour of day starts at 1:00 am (1) and ends at 12:00 am (24)


<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-15"


MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day, day_type) |> 
    glimpse()



```

<br><br>

The new features Year, Month, day_of_week, and day_type have a  data types of dbl or chr.  These will need to be transformed to a factor data type. 

<br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-16"


MA_accidents_2 <- MA_accidents_2 |> 
    mutate(Year = as.factor(Year)) |> 
    mutate(Month = as.factor(Month)) |> 
    mutate(day_of_week = as.factor(day_of_week)) |> 
    mutate(day_type = as.factor(day_type))



```

<br><br>






```{r}
#| echo: false
#| include: false
#| eval: false
#| label: "ck-17"


saveRDS(MA_accidents_2, "MA_accidents_2.rds")


```

```{r}
#| include: false
#| label: "ck-18"


MA_accidents_2 <- readRDS("MA_accidents_2.rds")



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-19"

MA_accidents_2 |> 
    select(Year, Month, day_of_week, day_type) |> 
    glimpse()


```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-20"


MA_accidents_2 |> 
    count(Year)


```


<br><br>


Before continuing, we will filter our data into two datasets,  the years 2016-2021 will be for training and testing, and the year 2022 for predicting on unseen data.  Since the year 2023 appears to be incomplete, we will omit it from our analysis

<br>

Years 2016 through 2021


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-21"


ma_accidents_16_21 <- MA_accidents_2 |> 
    filter(Year %in% c("2016", "2017", "2018", "2019", "2020","2021"))


```

```{r}
#| echo: false
#| include: false
#| eval: false
#| label: "ck-22"


saveRDS(ma_accidents_16_21, "ma_accidents_16_21.rds")

```

```{r}
#| include: false
#| label: "ck-23"


ma_accidents_16_21 <- readRDS("ma_accidents_16_21.rds")



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-24"


dim(ma_accidents_16_21)


```

<br><br>

Year 2022

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-25"



ma_accidents_22 <- MA_accidents_2 |> 
    filter(Year == "2022")


```

<br><br>

```{r}
#| echo: false
#| include: false
#| eval: false
#| label: "ck-26"


saveRDS(ma_accidents_22, "ma_accidents_22.rds")


```

```{r}
#| include: false
#| label: "ck-27"


ma_accidents_22 <- readRDS("ma_accidents_22.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-28"


dim(ma_accidents_22)


```


<br><br>


Our dataset has the feature Severity which is the target for prediction.  We were unable to access the definition of this feature and how it was determined, as such we will create a new feature for our target.  The time duration will be calculated from the Start_Time and End_Time features.  

<br>


Before we can  calculate the time duration of accidents, we'll first need to convert the start and end times to a date time object that includes year, month, day, hours, minutes, and seconds. 

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-29"

ma_accidents_16_21_2 <- ma_accidents_16_21 |> 
    mutate(Start_Time = ymd_hms(Start_Time)) |> 
    mutate(End_Time = ymd_hms(End_Time))
```

Before we can calculate the time duration of accidents, we'll first need to convert the start and end times to a date time object that includes year, month, day, hours, minutes, and seconds. 

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-30"

ma_accidents_16_21_2 |> 
    select(Start_Time, End_Time) |> 
    glimpse()


```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-31"


ma_accidents_16_21_2 |> 
    select(Start_Time, End_Time) |> 
    head()


```

<br><br>


Next, we'll create the new feature accident_duration, which will calcualte the duration of an accident in minutes. 

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-32"


ma_accidents_16_21_2 <- ma_accidents_16_21_2 |> 
    mutate(accident_duration = round(difftime(End_Time,Start_Time, units = "mins"), 0))


```

<br><br>

We will also create LATTITUDE and LONGITUDE features from Start_Lat and Start_Lng. These features will be useful for retrieving US Census data and mapping.

<br><br>





```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-33"



ma_accidents_16_21_2$LATITUDE <- ma_accidents_16_21_2$Start_Lat

ma_accidents_16_21_2$LONGITUDE <- ma_accidents_16_21_2$Start_Lng


```


```{r}
#| echo: false
#| include: false
#| eval: false
#| label: "ck-34"

saveRDS(ma_accidents_16_21_2, "ma_accidents_16_21_2.rds")


```


```{r}
#| include: false
#| label: "ck-35"


 ma_accidents_16_21_2 <- readRDS("ma_accidents_16_21_2.rds")


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-36"


ma_accidents_16_21_2 |> select(accident_duration) |>  head()


```

<br><br>


From the output, we observe the accident duration is in minutes. 

<br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-37"


ma_accidents_16_21_2 |> select(LATITUDE, LONGITUDE) |> head()


```

<br><br>

Our output shows the two new featurres. 


<br><br>


## Census Tract GEOID





<br><br>





```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| include: false
#| label: "ck-38"



Sys.getenv("CENSUS_API_KEY")


```





```{r}
#| echo: false
#| include: false
#| eval: false
#| label: "ck-39"


saveRDS(long_lat_1 , "long_lat_1.rds")


```

```{r}
#| echo: false
#| include: false
#| eval: false


long_lat_1  <- readRDS("long_lat_1.rds")


```



```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(long_lat_2, "long_lat_2.rds")

```

```{r}
#| echo: false
#| include: false
#| eval: false


long_lat_2 <- readRDS("long_lat_2.rds")


```






```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(long_lat_3, "long_lat_3.rds")


```


```{r}
#| echo: false
#| include: false
#| eval: false


long_lat_3 <- readRDS("long_lat_3.rds")



```




```{r}
#| echo: false
#| include: false
#| eval: false



saveRDS(long_lat_4, "long_lat_4.rds")



```


```{r}
#| echo: false
#| include: false
#| eval: false


long_lat_4 <- readRDS("long_lat_4.rds")


```



```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(long_lat_5, "long_lat_5.rds")



```


```{r}
#| echo: false
#| include: false
#| eval: false



long_lat_5 <- readRDS("long_lat_5.rds")


```






```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(long_lat_6, "long_lat_6.rds")



```


```{r}
#| echo: false
#| include: false
#| eval: false

long_lat_6 <- readRDS("long_lat_6.rds")


```




<br><br>


For our analysis we will require the GEOID for each census tract.  GEOID is a unique code for each US census tract.

<br>

The GEOID can be retrieved from the US Census API by submitting a data frame of the longitude and latitude for each observation of our dataset.  
<br>

This will be acomplished by using the call_geolocator_latlon function from the tigris library. 

<br><br>





```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-40"



long_lat$census_code <- apply(long_lat, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))


```

<br>


The GEOID returned includes the block extension.  Using the substr() function from the stringr library, we will extract the first 11 digits, which is our census trac GEOID.  

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-41"



long_lat$GEOID <- substr(long_lat$census_code, 1,11)


```


```{r}
#| echo: false
#| include: false
#| eval: false
#| label: "ck-42"


saveRDS(long_lat, "long_lat.rds")

```


```{r}
#| include: false
#| label: "ck-43"


long_lat <- readRDS("long_lat.rds")


```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-44"


long_lat |>  head()


```

<br><br>

We observe the GEOID for each associated Latitude and longitude corrdinates.  

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-45"


census_geo <- long_lat |> 
    select(-LATITUDE, -LONGITUDE)


```

```{r}
#| echo: false
#| include: false
#| eval: false
#| label: "ck-46"

saveRDS(census_geo, "census_geo.rds")


```


```{r}
#| include: false
#| label: "ck-47"

census_geo <- readRDS("census_geo.rds")



```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-48"




census_geo |> head()
```

<br><br>

We will combine the ma_accidents_10_21_2 and census_geo datasets. 

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-49"



ma_accidents_16_21_3 <- cbind(ma_accidents_16_21_2, census_geo)



```

```{r}
#| echo: false
#| include: false
#| eval: false
#| label: "ck-50"



saveRDS(ma_accidents_16_21_3, "ma_accidents_16_21_3.rds")


```


```{r}
#| include: false
#| label: "ck-51"


ma_accidents_16_21_3 <- readRDS("ma_accidents_16_21_3.rds")


```


<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-52"



ma_accidents_16_21_3 |> 
    summary()

```



There are five mssing values in the newly created GEOID.  These missing values (NAs) will be removed. 

<br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-53"


ma_accidents_16_21_4 <- ma_accidents_16_21_3 |> 
    filter(!is.na(GEOID))


```

<br><br>







The Wind_Speed.mph featues has 3,324 missinf values.  We will drop this feature from our dataset. 




<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-54"


ma_accidents_16_21_4 <- ma_accidents_16_21_4 |> 
    select(-Start_Lat, -End_Lat, -Start_Lng, -End_Lng, -State, -Zipcode, -Country, -Timezone, -Airport_Code, -Weather_Timestamp, -Wind_Chill.F., -Visibility.mi., -Precipitation.in., -Wind_Speed.mph., -Humidity...)


```


```{r}
#| echo: false
#| include: false
#| eval: false
#| label: "ck-55"


saveRDS(ma_accidents_16_21_4, "ma_accidents_16_21_4.rds")

```

```{r}
#| include: false
#| label: "ck-56"


ma_accidents_16_21_4 <- readRDS("ma_accidents_16_21_4.rds")


```


<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-57"


ma_accidents_16_21_4 |> 
    summary()


```


<br><br>

Remove NAs fron Temperature feature.

<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-58"


ma_accidents_16_21_5 <- ma_accidents_16_21_4|>
    filter(!is.na(Temperature.F.)) 


```


<br><br>

The accident_duration feature has mins at the end.  We will drop this by transforming  accident_duration from a difftime  data type to a numeric data type. 

<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-59"



ma_accidents_16_21_5$accident_duration <- as.numeric(ma_accidents_16_21_5$accident_duration)


```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-60"


ma_accidents_16_21_5 |> 
    count(County)




```

<br><br>


The Dukes and Nantucket counties have single digits counts, thus we will remove them from the dataset. 



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-61"



ma_accidents_16_21_5 <- ma_accidents_16_21_5 |> 
    filter(County  %in% c("Barnstable", "Berkshire", "Bristol", "Essex", "Franklin", "Hampden", "Hampshire", "Middlesex", "Norfolk", "Plymouth", "Suffolk", "Worcester"))


```

```{r}
#| echo: false
#| include: false
#| eval: false
#| label: "ck-62"


saveRDS(ma_accidents_16_21_5, "ma_accidents_16_21_5.rds")


```




```{r}
#| include: false
#| label: "ck-63"


ma_accidents_16_21_5  <- readRDS("ma_accidents_16_21_5.rds")


```



<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-64"

ma_accidents_16_21_5 |> 
    count(County)


```

<br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-65"




ma_accidents_16_21_5 |> select(accident_duration, Severity, Start_Time,End_Time, City, LONGITUDE, LATITUDE, Street, Sunrise_Sunset, Civil_Twilight,Street)  |> 
  filter(accident_duration > 5000)  |> head()



```

<br><br>


```{r}
#| echo: false
#| include: false
#| eval: false
#| label: "ck-66"



ma_accidents_16_21_5  |> select(accident_duration, Severity, Start_Time,End_Time, City, LONGITUDE, LATITUDE, Street, Sunrise_Sunset, Civil_Twilight,Street)  |> 
  filter(City == "Milton" & Street == "Adams St")  |> head()


```



```{r}
#| echo: false
#| include: false
#| eval: false
#| label: "ck-67"



ma_accidents_16_21_5  |> select(accident_duration, Severity, Start_Time,End_Time, Year, Month, day_of_week, City, LONGITUDE, LATITUDE, Street, Sunrise_Sunset, Civil_Twilight,Street)  |> 
  filter(Year == "2021" & Month == "4" & day_of_week == "4")  |> tail()

```




```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-68"




ma_accidents_16_21_5 |> 
    select_if(is.numeric) |> 
    glimpse()


```


<br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-69"
 

ma_accidents_16_21_6 <- ma_accidents_16_21_5 |> 
    rename(temperature = Temperature.F.)


 ```

 <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-70"
 


ma_accidents_16_21_5 |> 
    count(Wind_Direction)



 ```

 <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-71"

 

ma_accidents_16_21_6$Wind_Direction[trimws(ma_accidents_16_21_6$Wind_Direction) == ""] <- "NR"


 ```

 <br><br>

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-72"

 


ma_accidents_16_21_6 <- ma_accidents_16_21_6 |> 
    mutate(Wind_Direction = as.factor(Wind_Direction))


 ```

 <br><br>

 Wind_Direction has 25 categories.  We will consoladte these categories into 6. 

 <br><br>

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-73"
 




ma_accidents_16_21_6 <- ma_accidents_16_21_6 |> 
    mutate(Wind_Direction = recode(Wind_Direction,  "CALM" = "Calm", "NR" = "Calm", "ENE" = "E", "ESE" = "E", "East" = "E", "NE" = "N", "NNE" = "N", "NNW" = "N", "NW" = "N", "North" = "N", "SE" = "S", 
                                                    "SSE" = "S", "SSW" ="S", "SW" = "S", "South" = "S", "VAR" = "Variable", "WNW" = "W", "WSW" = "W", "West" = "W"))



 ```

 <br><br>



 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-74"
 

ma_accidents_16_21_6 <- ma_accidents_16_21_6 |> 
    select(-Source, -Start_Time, -End_Time, -Distance.mi., -Pressure.in.)


 ```

 <br><br>

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-75"
 

ma_accidents_16_21_6 <- ma_accidents_16_21_6 |> 
    mutate(Wind_Direction = as.factor(Wind_Direction),
            Weather_Condition = as.factor(Weather_Condition),
            Amenity = as.factor(Amenity),
            Bump = as.factor(Bump),
            Crossing = as.factor(Crossing),
            Give_Way = as.factor(Give_Way),
            Junction = as.factor(Junction),
            No_Exit = as.factor(No_Exit),
            Railway = as.factor(Railway),
            Roundabout = as.factor(Roundabout),
            Station = as.factor(Station),
            Stop = as.factor(Stop),
            Traffic_Calming = as.factor(Traffic_Calming),
            Traffic_Signal = as.factor(Traffic_Signal),
            Turning_Loop = as.factor(Turning_Loop),
            Sunrise_Sunset = as.factor(Sunrise_Sunset),
            Civil_Twilight = as.factor(Civil_Twilight),
            Nautical_Twilight = as.factor(Nautical_Twilight),
            Astronomical_Twilight = as.factor(Astronomical_Twilight))


 ```

 ```{r}
#| echo: false
#| include: false
#| eval: false
#| label: "ck-76"
 

saveRDS(ma_accidents_16_21_6 , "ma_accidents_16_21_6.rds")


 ```

 ```{r}
#| include: false 
#| label: "ck-77"
 

ma_accidents_16_21_6  <- readRDS("ma_accidents_16_21_6.rds")



 ```

 <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-78"

 

ma_accidents_16_21_6 |> 
    select_if(is.factor) |> 
    summary()


 ```


<br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-79"
 


ma_accidents_16_21_6 |> 
    count(Crossing) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-80"
 

ma_accidents_16_21_6 |> 
    count(Wind_Direction) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-81"
 

 ma_accidents_16_21_6 |> 
    group_by(Wind_Direction) |> 
    summarise(median_accident_duration_by_wind_direction = median(accident_duration)) 
   


 ```

 
 <br><br>


 


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-82"
 

ma_accidents_16_21_6 |> 
    count(County, Bump)



 ```




 <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-83"
 


ma_accidents_16_21_7 <- ma_accidents_16_21_6 |> 
    mutate(Street = recode(Street,  "Mass Pike" = "I-90"))


 ```

  <br><br>

 
We will create a new feature which will include the types of road categories, such as interstate, Route, etc. To accomplish this, we will use the grep function and regular expressions to extract roads and highways and then consolidate them into categories.  An example is extracting I-90 and I-95 and consolidating them into an interstate category. 
  <br><br>


 

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-84"
 

col_to_check <- ma_accidents_16_21_7$Street

ma_accidents_16_21_7$road_category <- "Uncategorized"


ma_accidents_16_21_7$road_category[grepl("\\b(I|Interstate)[-\\s]?\\d+", col_to_check, ignore.case = TRUE)] <- "Interstate"
ma_accidents_16_21_7$road_category[grepl("\\b(Route|Rte|Rt|State Route|SR|State Hwy|MA)[. -]*\\d+", col_to_check, ignore.case = TRUE)] <- "MA Highway"
ma_accidents_16_21_7$road_category[grepl("\\b(US)[-\\s]?\\d+", col_to_check, ignore.case = TRUE)] <- "US Route"
ma_accidents_16_21_7$road_category[grepl("\\b(Rd|Road)\\b", col_to_check, ignore.case = TRUE)] <- "Road"
ma_accidents_16_21_7$road_category[grepl("way(\\b|[ .]+(N|S|E|W|NE|NW|SE|SW|North|South|East|West)?$)", col_to_check, ignore.case = TRUE)] <- "Way"
ma_accidents_16_21_7$road_category[grepl("\\b(St|Street)\\b", col_to_check, ignore.case = TRUE)] <- "Street"
ma_accidents_16_21_7$road_category[grepl("\\b(Ave|Avenue)\\b", col_to_check, ignore.case = TRUE)] <- "Avenue"
ma_accidents_16_21_7$road_category[grepl("\\b(Blvd|Boulevard)\\b", col_to_check, ignore.case = TRUE)] <- "Boulevard"
ma_accidents_16_21_7$road_category[grepl("\\b(Dr|Drive)\\b", col_to_check, ignore.case = TRUE)] <- "Drive"
ma_accidents_16_21_7$road_category[grepl("\\b(Ln|Lane)\\b", col_to_check, ignore.case = TRUE)] <- "Lane"
ma_accidents_16_21_7$road_category[grepl("\\b(Hwy|Highway)\\b", col_to_check, ignore.case = TRUE)] <- "Highway"
ma_accidents_16_21_7$road_category[grepl("\\b(Pkwy|Parkway)\\b", col_to_check, ignore.case = TRUE)] <- "Parkway"
ma_accidents_16_21_7$road_category[grepl("\\b(Brg|Bridge)\\b", col_to_check, ignore.case = TRUE)] <- "Bridge"
ma_accidents_16_21_7$road_category[grepl("\\b(Cir|Circle)\\b", col_to_check, ignore.case = TRUE)] <- "Circle"
ma_accidents_16_21_7$road_category[grepl("\\b(Trl|Trail)\\b", col_to_check, ignore.case = TRUE)] <- "Trail"
ma_accidents_16_21_7$road_category[grepl("\\b(Ct|Cor|Court)\\b", col_to_check, ignore.case = TRUE)] <- "Court"
ma_accidents_16_21_7$road_category[grepl("\\b(Ter|Terrace)\\b", col_to_check, ignore.case = TRUE)] <- "Terrace"
ma_accidents_16_21_7$road_category[grepl("\\b(Pl|Place)\\b", col_to_check, ignore.case = TRUE)] <- "Place"
ma_accidents_16_21_7$road_category[grepl("\\b(Tpke|Turnpike)\\b", col_to_check, ignore.case = TRUE)] <- "Turnpike"
ma_accidents_16_21_7$road_category[grepl("\\b(Sq|Square)\\b", col_to_check, ignore.case = TRUE)] <- "Square"
ma_accidents_16_21_7$road_category[grepl("\\b(Pth|Path)\\b", col_to_check, ignore.case = TRUE)] <- "Path"
ma_accidents_16_21_7$road_category[grepl("\\b(Conn|Connector)\\b", col_to_check, ignore.case = TRUE)] <- "Connector"
ma_accidents_16_21_7$road_category[grepl("\\b(Ext|Exit)\\b", col_to_check, ignore.case = TRUE)] <- "Exit"
ma_accidents_16_21_7$road_category[grepl("\\b(Rty|Rotary)\\b", col_to_check, ignore.case = TRUE)] <- "Rotary"
ma_accidents_16_21_7$road_category[grepl("\\b(Expy|Expressway)\\b", col_to_check, ignore.case = TRUE)] <- "Expressway"
ma_accidents_16_21_7$road_category[grepl("\\b(Tunl|Tunnel)\\b", col_to_check, ignore.case = TRUE)] <- "Tunnel"


 ```

   <br><br>



 

 ```{r}
#| code-fold: true
#| code-summary: "Show Code" 
#| label: "ck-85"
 
ma_accidents_16_21_6 |> count(Weather_Condition)


 ```

   <br><br>


   Next, we will reudce the numbre of Weawther_Condition categories through consolidation. 

     <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code" 
#| label: "ck-86"
 

ma_accidents_16_21_6 |> select(Weather_Condition, City) |> 
    filter(Weather_Condition == "") |> 
    head(15)


 ```

   <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-87"
 
ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    mutate(Weather_Condition = as.character(Weather_Condition))


 ```

   <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-88"
 

ma_accidents_16_21_7$Weather_Condition[trimws(ma_accidents_16_21_7$Weather_Condition) == ""] <- "NR"


 ```

   <br><br>



 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-89"
 

ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    mutate(Weather_Condition = as.factor(Weather_Condition))


 ```

   <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-90"
 

ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    mutate(Weather_Condition = recode(Weather_Condition,  "NR" = "Fair Partley Cloudy", "Light Rain / Windy" = "Light Rain", "Light Rain Shower" = "Light Rain", "Light Rain with Thunder" = "Light Rain", "Light Drizzle" = "Light Rain","Rain / Windy" = "Rain", "Thunderstorms and Rain" = "Rain", "Thunderstorm" = "Rain", "T-Storm" = "Rain","Thunder" = "Overcast", "Cloudy" = "Overcast", "Thunder / Windy" = "Overcast", "Thunder in the Vicinity" = "Overcast","Cloudy / Windy" = "Overcast", "Mostly Cloudy" = "Overcast", "Mostly Cloudy / Windy" = "Overcast","Light Thunderstorms and Rain" = "Rain","Heavy Rain / Windy" = "Heavy Rain", "Heavy Thunderstorms and Rain" = "Heavy Rain","Heavy T-Storm / Windy" = "Heavy Rain","Heavy T-Storm" = "Heavy Rain", "Light Snow / Windy" = "Light Snow", "Light Snow and Sleet" = "Light Snow", "Clear" = "Fair Partley Cloudy","Blowing Snow" = "Heavy Snow", "Heavy Snow / Windy" = "Heavy Snow", "Light Drizzle / Windy" = "Light Rain", "Mist" = "Light Rain", "Drizzle and Fog" = "Light Rain", "Fair / Windy" = "Fair Partley Cloudy", "Partly Cloudy" = "Fair Partley Cloudy", "Partly Cloudy / Windy" = "Fair Partley Cloudy", "Fair" = "Fair Partley Cloudy", "Scattered Clouds" = "Fair Partley Cloudy",
   "Fog / Windy" = "Fog", "Patches of Fog" = "Fog","Shallow Fog" = "Fog","Wintry Mix / Windy" = "Wintry Mix", "Light Freezing Drizzle" = "Wintry Mix" , "Light Freezing Rain" = "Wintry Mix","Smoke" = "Haze", "Squalls" = "Snow",
   "Light Freezing Fog" = "Wintry Mix", "Ice Pellets" = "Wintry Mix","Small Hail" = "Wintry Mix",  "Snow / Windy" = "Snow", "N/A Precipitation" = "Fair Partley Cloudy") )


 ```


  <br><br>

Next, we create a new feature hour_day_sextion.  This will be done by binning an hour of day into specific periods of the day such as afternoon, night. 


  <br><br>

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-91"
 



ma_accidents_16_21_7 <- ma_accidents_16_21_7|> 
    mutate(hour_day_sections = case_when(
        between(hour_of_day,5,6) ~ "Dawn",
        between(hour_of_day,6,9) ~ "Early Morning",
        between(hour_of_day,9,12) ~ "Late Morning",
        between(hour_of_day,12,16) ~ "Early Afternoon",
        between(hour_of_day,16,18) ~ "Late Afternoon",
        between(hour_of_day,18,21) ~ "Evening",
        between(hour_of_day,21,24) ~ "Night",
        TRUE ~ "Middle of Night"

    ))



 ```

   <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-92"
 


ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    mutate(hour_day_sections = as.factor(hour_day_sections))


 ```

   <br><br>


We can remove the hour_of_day feature now that hour_day_sections feature has been created. 

<br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-93"
 

ma_accidents_16_21_7 <- ma_accidents_16_21_7 |> 
    select(-hour_of_day)


 ```

   <br><br>


 ```{r}
#| echo: false
#| include: false
#| eval: false
#| label: "ck-94"
 

saveRDS(ma_accidents_16_21_7, "ma_accidents_16_21_7.rds")



 ```


 ```{r}
#| include: false
#| label: "ck-95"
 

ma_accidents_16_21_7 <- readRDS("ma_accidents_16_21_7.rds")


 ```

   <br><br>




 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-96"
 

ma_accidents_16_21_7 |> 
    count(hour_day_sections) |> 
    mutate(perc = round(n/sum(n),3)*100) |> 
    arrange(desc(perc))



 ```

 <br><br>


 We now have eight categories of different parts of the day. 

<br><br>



 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-97"
 

ma_accidents_16_21_7 |> 
    count(road_category)


 ```

 27 road categories. 

 <br><br>

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-98"
 

print(ma_accidents_16_21_7 |> 
    group_by(road_category) |> 
    summarise(median_acc_duration_road = median(accident_duration)) |> 
    arrange(desc(median_acc_duration_road)), n=27)


 ```

 <br><br>

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-99"
 

ma_accidents_16_21_7 |> 
    count(Weather_Condition)



 ```

  <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-100"

 

print(ma_accidents_16_21_7 |> 
    group_by(Weather_Condition) |> 
    summarise(median_acc_duration_weather = median(accident_duration)) |> 
    arrange(desc(median_acc_duration_weather)), n=11)


 ```


  <br><br>



 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
#| label: "ck-101"
 

ma_accidents_16_21_7 |> 
    select_if(is.factor) |> 
    summary()

 ```

  <br><br>



 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-102"
 

ma_accidents_16_21_7 |> 
    count(No_Exit) |> 
    mutate(perc = round(n/sum(n),4)) |> 
    arrange(desc(perc))


 ```

  <br><br>

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-103"
 
ma_accidents_16_21_7 |> 
    count(County, No_Exit)

 ```

  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-104"


ma_accidents_16_21_7|> 
    group_by(No_Exit ) |> 
    summarise(median_acc_dur_exit = median(accident_duration)) |> 
    arrange(desc(median_acc_dur_exit))

```

 <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-105"
 

ma_accidents_16_21_7 |> 
    count(Crossing) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

  <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-106"
 
ma_accidents_16_21_7 |> 
    count(Railway) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))



 ```

  <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-107"

 


 ma_accidents_16_21_7 |> 
    count(County, Railway)

 ```

  <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-108"
 
ma_accidents_16_21_7 |> 
    count(Give_Way) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

  <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-109"
 


ma_accidents_16_21_7 |> 
    count(County, Give_Way)


 ```

  <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-110"
 

ma_accidents_16_21_7 |> 
    count(Amenity) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

  <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-111"
 
ma_accidents_16_21_7 |> 
    count(County, Amenity)


 ```

  <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-112"
 

ma_accidents_16_21_7 |> 
    count(Junction) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

  <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-113"
 

ma_accidents_16_21_7 |> 
    count(County, Junction)


 ```

  <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-114"
 

ma_accidents_16_21_7 |> 
    count(Station) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))

 ```

  <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-115"
 


ma_accidents_16_21_7 |> 
    count(County, Station)

 ```

  <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| label: "ck-110"
 

ma_accidents_16_21_7 |> 
    count(Stop) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```
  <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

 ma_accidents_16_21_7 |> 
    count(County, Stop)
 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_7 |> 
    count(Traffic_Calming) |> 
    mutate(perc = round(n/sum(n),3)) |> 
    arrange(desc(perc))

 ```

  <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_7|> 
    group_by(Traffic_Calming) |> 
    summarise(median_acc_dur = median(accident_duration))



 ```

  <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_7 |> 
    count(Stop) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

  <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_7 |> 
    count(Traffic_Signal) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))


 ```

  <br><br>


 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

 ma_accidents_16_21_7 |> 
    count(County, Traffic_Signal)


 ```

  <br><br>






 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false
 

ma_accidents_16_21_8 <- ma_accidents_16_21_7 |> 
    select(-No_Exit, -Traffic_Calming, -Nautical_Twilight, -Astronomical_Twilight, -Give_Way, -Railway, -Description, -Street, -Bump, -Roundabout, -Turning_Loop)


 ```

 ```{r}
#| echo: false
#| include: false
#| eval: false
 
saveRDS(ma_accidents_16_21_8, "ma_accidents_16_21_8.rds")


 ```

 ```{r}
 #| include: false
 


ma_accidents_16_21_8 <- readRDS("ma_accidents_16_21_8.rds")

 ```

 ```{r}
#| code-fold: true
#| code-summary: "Show Code"
 

ma_accidents_16_21_8 |> 
    glimpse()


 ```

 

 <br><br>

 

  <br><br>





  ```{r}
  #| echo: false
#| include: false
#| eval: false
  
saveRDS(ma_accidents_16_21_9, "ma_accidents_16_21_9.rds")


  ```

  ```{r}
#| echo: false
#| include: false
#| eval: false
  

ma_accidents_16_21_9 <- readRDS("ma_accidents_16_21_9.rds")


  ```


  <br><br>




  <br><br>






```{r}





```

```{r}
#| echo: false
#| include: false
#| eval: false



ma_accidents_16_21_3  <- ma_accidents_16_21_2  |> 
    mutate(id = row_number())


```





# Census

## County




```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_county <- counties("MA", year = 2024)

```

  <br><br>

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_county, "ma_county.rds")


```

```{r}
#| include: false


ma_county <- readRDS("ma_county.rds")


```

  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_county |> 
    glimpse()


```

  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_county |> 
    summary()


```
  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"


print(ma_county |> select(NAME), n=14)


```


  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_county_shape <- ma_county |> 
    filter(NAME != "Nantucket") |> 
    filter(NAME != "Dukes")


```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_county_shape , "ma_county_shape.rds")


```


```{r}
#| include: false


ma_county_shape  <- readRDS("ma_county_shape.rds")



```

  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"



print(ma_county_shape |> select(NAME), n=14) 


```

  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_county_shape_2 <- ma_county_shape |> 
    select(COUNTYFP, GEOID, NAME, NAMELSAD, INTPTLAT, INTPTLON, geometry)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

ma_county_shape_2 <- ma_county_shape_2 |> 
    rename("GEOID_COUNTY" = "GEOID")


```

```{r}
#| echo: false
#| include: false
#| eval: false



saveRDS(ma_county_shape_2, "ma_county_shape_2.rds")

```

```{r}
#| include: false



ma_county_shape_2 <- readRDS("ma_county_shape_2.rds")


```

  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_county_shape_2 |> glimpse()



```



```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(census_county_urban, "census_county_urban.rds")


```

```{r}
#| include: false

census_county_urban <- readRDS("census_county_urban.rds")


```

  <br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"


census_county_urban |> glimpse()


```

  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"


census_county_urban |> 
    count(STATE)


```

  <br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_county_urban <- census_county_urban |> 
    filter(STATE_NAME == "Massachusetts")


```


```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_county_urban , "ma_county_urban.rds")


```

```{r}
#| include: false


ma_county_urban  <- readRDS("ma_county_urban.rds")

```

  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_county_urban |> summary()


```

  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"



print(ma_county_urban |> select(COUNTY_NAME), n=14 )


```


  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_county_urban_2 <- ma_county_urban |> 
    filter(COUNTY_NAME != "Dukes") |> 
    filter(COUNTY_NAME != "Nantucket")


```

  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_county_urban_2 <- ma_county_urban_2 |> 
    select(COUNTY, POP_COU, HOU_COU, `ALAND_Mi²_COU`, POP_URB, POPPCT_URB, HOU_URB, HOUPCT_URB, `ALAND_Mi²_URB`, ALAND_PCT_URB, POP_RUR, POPPCT_RUR, HOU_URB, HOUPCT_RUR, `ALAND_Mi²_RUR`, ALAND_PCT_RUR)



```

  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_county_urban_2 <- ma_county_urban_2 |> 
    rename("ALAND_MI_COU" = "ALAND_Mi²_COU", "ALAND_MI_URB" = "ALAND_Mi²_URB", "ALAND_MI_RUR" = "ALAND_Mi²_RUR", "COUNTYFP" = "COUNTY")

```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_county_urban_2, "ma_county_urban_2.rds")



```

```{r}
#| include: false


ma_county_urban_2 <- readRDS("ma_county_urban_2.rds")


```

  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_county_urban_2 |> glimpse()


```

  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"



print(ma_county_urban_2 |> select(COUNTYFP), n=14)




```

  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_county_urban_3 <- ma_county_urban_2 |> 
    mutate(county_pop_density = round(POP_COU / ALAND_MI_COU,0), county_house_density = round(HOU_COU / ALAND_MI_COU,0))


```


  <br><br>

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_county_urban_3, "ma_county_urban_3.rds")


```

```{r}
#| include: false


ma_county_urban_3 <- readRDS("ma_county_urban_3.rds")


```

  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"


print(ma_county_urban_3 |> select(county_house_density, county_pop_density),n=14)



```

  <br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


county_geo <- ma_county_shape_2 |> 
    full_join(ma_county_urban_3, by="COUNTYFP")



```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(county_geo , "county_geo.rds")


```

```{r}
#| include: false



county_geo  <- readRDS("county_geo.rds")



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



county_geo |> glimpse()


```

  <br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


county_geo_2 <- county_geo |> 
    select(COUNTYFP, NAME, POP_COU, HOU_COU, POPPCT_URB, HOU_URB, county_pop_density) |> 
    st_drop_geometry()


```

  <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

county_geo_2 <- county_geo_2 |> 
    rename("County" = "NAME", "Cou_POPPCT_URB" = "POPPCT_URB", "COU_HOU_URB" = "HOU_URB")




```



  <br><br>


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(county_geo_2, "county_geo_2.rds")

```

```{r}
#| include: false


county_geo_2 <- readRDS("county_geo_2.rds")


```


  <br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"


county_geo_2 |> glimpse()


```


  <br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_county_urban |> 
    select(COUNTY_NAME, POPDEN_COU, POPPCT_URB, HOUDEN_COU, HOUDEN_URB,HOUDEN_RUR, HOUPCT_URB)


```


  <br><br>

## Tracts

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tracts <- tracts("MA", year = 2024)


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tracts, "ma_tracts.rds")


```

```{r}
#| include: false

ma_tracts <- readRDS("ma_tracts.rds")


```


  <br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tracts


```

  <br><br>


We will define the census tract variables we want to retrieve from the US Census American Communities Survey (acs).
 <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


variables_to_get <- c(
    total_tract_housing_units = "B25001_001",
    total_vehicles_tract = "B25046_001",
    total_households_tract = "B25044_001",
    total_population_tract = "B01003_001"


)


```

 <br><br>




```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data <- get_acs(
    geography = "tract",
    variables = variables_to_get,
    state = "MA",
    geometry = TRUE,
    output = "wide",
    year = 2021
)



```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tract_data, "ma_tract_data.rds")


```


```{r}
#| include: false

ma_tract_data <- readRDS("ma_tract_data.rds")



```




 <br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data |> 
    glimpse()


```

<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data |> summary()


```


<br><br>




```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_geometry <- ma_tract_data |> 
    select(GEOID, geometry)


```

<br><br>



```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tract_geometry, "ma_tract_geometry.rds")

```

```{r}
#| include: false


ma_tract_geometry <- readRDS("ma_tract_geometry.rds")


```


<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_tract_cls <- ma_tract_geometry |> 
    st_transform(5070) 


```

<br><br>


```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_tract_cls, "ma_tract_cls.rds")


```

```{r}
#| include: false


ma_tract_cls <- readRDS("ma_tract_cls.rds")


```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_data |> 
    select(NAME) |> 
    head(10)



```

<br><br>




The output of the NAME column shows it contains Census Tract, Coumty, and State. Will will use the separate() function to split the NAME column into three new columns.


<br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

ma_tract_data_2 <- ma_tract_data |> 
    separate(NAME, into=c("census_tract", "County", "state"), sep=", ") |> 
    st_drop_geometry()



```

<br><br>


Remove State

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_2 <- ma_tract_data_2 |> 
    select(-state)


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tract_data_2 , "ma_tract_data_2.rds")

```


```{r}
#| include: false

ma_tract_data_2  <- readRDS("ma_tract_data_2.rds")


```

<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_3 <- ma_tract_geometry |> 
    full_join(ma_tract_data_2, by = "GEOID")


```



```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tract_data_3 , "ma_tract_data_3.rds")

```

```{r}
#| include: false


ma_tract_data_3  <- readRDS("ma_tract_data_3.rds")


```


<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data_3 |> summary()




```

<br><br>



```{r}
#| echo: false
#| include: false
#| eval: false


ma_tract_data_4 <- ma_tract_data_3 |> 
    filter(GEOID != "25001990000") |> 
    filter(GEOID != "25023990003")



```



<br><br>






```{r}
#| echo: false
#| include: false
#| eval: false


ma_tract_data_4 |> 
    filter(GEOID == "25001990000") |> 
    head()


```


<br><br>

### Missing Values

<br><br>
<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_data_3 |> 
    select(GEOID, total_vehicles_tractE) |> 
    filter(is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summary()



```


<br><br>


```{r}
#| echo: false
#| include: false
#| eval: false




ma_tract_data_3 |> 
    select(GEOID, census_tract,County, total_vehicles_tractE) |> 
    filter(is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    head(72)

```



<br><br>

Certain tract variables we retrieved from ACS have missing values.  We will replace these values with the median of the county the census tract is located in. 


<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"





ma_tract_data_3 |> 
    filter(County == "Suffolk County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))




```

<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Suffolk County" & is.na(total_vehicles_tractE), 1310,total_vehicles_tractE ))




```

<br><br>


<br><br>






```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data_3 |> 
    filter(County == "Bristol County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Bristol County" & is.na(total_vehicles_tractE), 2785,total_vehicles_tractE ))



```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"




ma_tract_data_3 |> 
    filter(County == "Hampshire County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))



```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Hampshire County" & is.na(total_vehicles_tractE), 3340,total_vehicles_tractE ))


```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data_3 |> 
    filter(County == "Essex County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Essex County" & is.na(total_vehicles_tractE), 2989,total_vehicles_tractE ))


```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data_3 |> 
    filter(County == "Middlesex County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Middlesex County" & is.na(total_vehicles_tractE), 2828,total_vehicles_tractE))
```

<br><br>




```{r}
#| code-fold: true
#| code-summary: "Show Code"




ma_tract_data_3 |> 
    filter(County == "Plymouth County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))

```

<br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Plymouth County" & is.na(total_vehicles_tractE), 3471,total_vehicles_tractE))



```

<br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"




ma_tract_data_3 |> 
    filter(County == "Worcester County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```

<br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Worcester County" & is.na(total_vehicles_tractE), 2898,total_vehicles_tractE))



```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data_3 |> 
    filter(County == "Barnstable County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))

```

<br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Barnstable County" & is.na(total_vehicles_tractE), 3166,total_vehicles_tractE))


```

<br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_data_3 |> 
    filter(County == "Hampden County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```





```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Hampden County" & is.na(total_vehicles_tractE), 2736,total_vehicles_tractE))


```

<br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data_3 |> 
    filter(County == "Berkshire County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```

<br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Berkshire County" & is.na(total_vehicles_tractE), 2248,total_vehicles_tractE))


```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_data_3 |> 
    filter(County == "Norfolk County") |> 
    filter(!is.na(total_vehicles_tractE)) |> 
    st_drop_geometry() |> 
    summarise(median_tract_vehicles = median(total_vehicles_tractE))


```


<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_4 <- ma_tract_data_4 |> 
    mutate(total_vehicles_tractE = ifelse(County == "Norfolk County" & is.na(total_vehicles_tractE), 3080,total_vehicles_tractE))



```



<br><br>





```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tract_data_4, "ma_tract_data_4.rds")

```

```{r}
#| include: false


ma_tract_data_4 <- readRDS("ma_tract_data_4.rds")



```

<br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_data_4 |> 
    summary()


```


```{r}
#| echo: false
#| include: false
#| eval: false


ma_tract_data_5 <- ma_tract_data_4 |> 
    filter(total_population_tractE != 0)


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_5 <- ma_tract_data_4 |> 
    filter(County != "Nantucket County") |> 
    filter(County != "Dukes County")


```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_5 <- ma_tract_data_5 |> 
    filter(!is.na(total_households_tractE))



```

<br><br>



```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tract_data_5, "ma_tract_data_5.rds")


```

```{r}
#| include: false


ma_tract_data_5 <- readRDS("ma_tract_data_5.rds")



```

<br><br>



```{r}

ma_tract_data_5 |> count(County)


```

<br><br>




```{r}
#| code-fold: true
#| code-summary: "Show Code"

ma_tract_data_5 |> 
    filter(is.na(total_vehicles_tractE)) |> 
    count(County)


```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_data_5 |> 
    summary()


```

<br><br>

<br><br>






### Feature Engineering
<br><br>
<br><br>



From our retrieved census tract variables, we will create new variables for housing density , population density , and average cars per houshold. 



<br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_6 <- ma_tract_data_5 |> 
    mutate(
        tract_house_density_sqmi = round(as.numeric(set_units(total_tract_housing_unitsE/ st_area(geometry),"1/mi2")),2),
        tract_pop_density_sqmi = round(as.numeric(set_units(total_population_tractE/ st_area(geometry),"1/mi2")),2),
        avg_cars_per_household = round(total_vehicles_tractE / total_households_tractE, 2)  
    #select(!ends_with("M"))
    #rename_with(.fn = ~str_remove(.x, "E$"))


    )



```

<br><br>




```{r}
#| echo: false
#| include: false
#| eval: false




print(ma_tract_data_5 |> 
    filter(total_households_tractE == 0), n=70)


```






```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_6 <- ma_tract_data_6 |> 
    select(-total_tract_housing_unitsM, -total_vehicles_tractM, -total_households_tractM, -total_population_tractM)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_6 <- ma_tract_data_6 |> 
    select(-County)


```

<br><br>



```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tract_data_6 , "ma_tract_data_6.rds")


```

```{r}
#| include: false


ma_tract_data_6 <- readRDS("ma_tract_data_6.rds")


```

<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"





ma_tract_data_6|> glimpse()



```

<br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data_6 |> summary()


```

<br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_data_7 <- ma_tract_data_6 |> 
    select(-census_tract) |> 
    st_drop_geometry()




```

<br><br>


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_tract_data_7, "ma_tract_data_7.rds")


```

```{r}
#| include: false

ma_tract_data_7 <- readRDS("ma_tract_data_7.rds")


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_data_7|> 
    glimpse()


```






## Joining


# Exploratory Data Analysis

<br><br>




## Features


<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_accidents_16_21_6 |> 
    ggplot(aes(x=accident_duration, y=temperature)) +
    geom_point()



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"




ma_accidents_16_21_6 |> 
    ggplot(aes(x=log(accident_duration), y=temperature)) +
    geom_point()


```

<br><br>



## Maps

<br><br>



### County

<br><br>





```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_county_geo_4326 <- county_geo |> 
    st_transform(crs=4326)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_county_stcrs <- st_crs(ma_county_geo_4326)


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_county_stcrs$input



```









```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



bins<-c(0,71029,150000,400000,600000,800000,1000000,1300000,Inf)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false




pal<-colorBin(palette="RdYlBu",
            domain=ma_county_geo_4326$POP_COU,bins=bins)
```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false




ma_county_formated <- ma_county_geo_4326 |> 
    mutate(county_pop_density = format(county_pop_density, big.mark = ",", big.interval = 3, digits=12) ) |> 
    mutate(POPPCT_URB = round(POPPCT_URB * 100,2))


```

ma_county_formated 



```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_county_formated , "ma_county_formated.rds")


```

```{r}
#| include: false


ma_county_formated  <- readRDS("ma_county_formated.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false





mytext <- paste(
  "County: ", ma_county_formated$NAME,"<br/>",
  "Population Density: ", ma_county_formated$county_pop_density, "<br/>",
  "County Population Percent Urban: ", ma_county_formated$POPPCT_URB,"<br/>", 
    sep="") %>%
  lapply(htmltools::HTML)


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



map_county_pop_density_plt<- ma_county_geo_4326 |> leaflet(options = leafletOptions(minZoom = 9.4, maxZoom = 18)) |> 
  addProviderTiles(providers$Esri.WorldStreetMap) |> 
  setView(lng = -071.9402820, lat = 42.3116934, zoom = 9.4) |> 
  addPolygons(
 fillColor=~pal(POP_COU),
 weight=2,
 opacity=0.3,
 color= "white",
 label=~mytext,
 dashArray = "3",
 fillOpacity=0.3,
 highlightOptions = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.5,
    bringToFront = FALSE)) 
    
  




```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(map_county_pop_density_plt, "map_county_pop_density_plt.rds")

```

```{r}
#| include: false




map_county_pop_density_plt <- readRDS("map_county_pop_density_plt.rds")

```



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


county_map_pop_density<-map_county_pop_density_plt %>% 
  addLegend(pal = pal, values = ~county_pop_density, opacity = 0.6, title = "Massachusetts County Population Density",
  position = "bottomright")


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(county_map_pop_density, "county_map_pop_density.rds")

```

```{r}
#| include: false


county_map_pop_density <- readRDS("county_map_pop_density.rds")


```

```{r}

county_map_pop_density



```



### Census Tracts




```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

tract_accident_count <- ma_accidents_16_21_8 |> 
    count(GEOID, name = "tract_accident_count")


```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(tract_accident_count, "tract_accident_count.rds")

```

```{r}
#| include: false


tract_accident_count <- readRDS("tract_accident_count.rds")



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"





tract_accident_count |> 
    slice_max(tract_accident_count, n=20) 

```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_8 |> 
    filter(GEOID == "25017334100")



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_accidents_16_21_8 |> 
    filter(accident_duration >=30 & accident_duration <=60)

```


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_accidents_16_21_10 <- ma_accidents_16_21_8 |> 
    full_join(tract_accident_count, by = "GEOID")



```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_accidents_16_21_10, "ma_accidents_16_21_10.rds")


```

```{r}
#| include: false



ma_accidents_16_21_10 <- readRDS("ma_accidents_16_21_10.rds")


```

 ```{r}
#| echo: false
#| include: false
#| eval: false

 
ma_accidents_16_21_10 |> 
 count(GEOID) |> 
 arrange(desc(n))


 ```

 ```{r}
 



 ```

 ```{r}
 #| code-fold: true
#| code-summary: "Show Code"

 

ma_accidents_16_21_10 |> 
    summary()



 ```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_tract_accident <- ma_tract_data_6 |> 
    full_join(ma_accidents_16_21_10, by = "GEOID")


```

```{r}
#| echo: false
#| include: false
#| eval: false



saveRDS(ma_tract_accident, "ma_tract_accident.rds")


```

```{r}
#| include: false


ma_tract_accident <- readRDS("ma_tract_accident.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"




ma_tract_accident |> 
    summary()



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_accident |> 
    filter(is.na(total_households_tractE)) |> 
    summary()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"





ma_tract_accident |> 
    filter(is.na(total_households_tractE)) |> 
    count(GEOID)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_accident |> 
    filter(is.na(total_households_tractE)) |> 
    head(1000)



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_tract_accident_2 <- ma_tract_accident |> 
    mutate(total_households_tractE = replace_na(total_households_tractE, 0), 
           total_population_tractE = replace_na(total_population_tractE, 0),
           tract_house_density_sqmi = replace_na(tract_house_density_sqmi,0),
           tract_pop_density_sqmi = replace_na(tract_pop_density_sqmi,0),
           total_vehicles_tractE = replace_na(total_vehicles_tractE,0),
           total_tract_housing_unitsE = replace_na(total_tract_housing_unitsE,0) )


```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_tract_accident_2, "ma_tract_accident_2.rds")

```

```{r}
#| include: false


ma_tract_accident_2 <- readRDS("ma_tract_accident_2.rds")

```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_accident_2 |> 
    summary()

```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false




ma_tract_accident_3 <- ma_tract_accident_2 |> 
    filter(!is.na(day_of_week))
```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_tract_accident_3, "ma_tract_accident_3.rds")

```

```{r}
#| include: false

ma_tract_accident_3 <- readRDS("ma_tract_accident_3.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

ma_tract_accident_3 |> 
    summary()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_cls <- st_crs(ma_tract_accident_3)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

ma_tract_cls


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

ma_census4326 <- ma_tract_accident_3 |> 
    st_transform(crs=4326)


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_census4326, "ma_census4326.rds")


```

```{r}
#| include: false



ma_census4326 <- readRDS("ma_census4326.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_tract_cls_2 <- st_crs(ma_census4326)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_tract_cls_2$input


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


bins<-c(0,500,2000,3000,4000,5000,6500,8500,9500,10500,Inf)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



pal<-colorBin(palette="RdYlBu",
            domain=ma_census4326$total_population_tractE,bins=bins)



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

       
       
ma_tract_formated <- ma_tract_accident_3 |> 
    mutate(total_tract_housing_unitsE = format(total_tract_housing_unitsE, big.mark = ",", big.interval = 3, digits=12) ) |> 
    mutate(total_vehicles_tractE = format(total_vehicles_tractE, big.mark = ",", big.interval = 3, digits=12) ) 
   


```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_tract_formated , "ma_tract_formated.rds")


```

```{r}
#| include: false

ma_tract_formated  <- readRDS("ma_tract_formated.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



mytext <- paste(
  "County: ", ma_tract_formated$County,"<br/>",
  "Census Tract: ", ma_tract_formated$GEOID, "<br/>",
  "Total Vehicles Census Tract: ", ma_tract_formated$total_vehicles_tractE, "<br/>",
  "Total Housing Census Tract: ", ma_tract_formated$total_tract_housing_unitsE,"<br/>", 
  "Census Tract Accident Count: ", ma_tract_formated$tract_accident_count,"<br/>",
    sep="") %>%
  lapply(htmltools::HTML)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false




map_tract_population_plt<- ma_census4326 |> leaflet(options = leafletOptions(minZoom = 9.4, maxZoom = 18)) |> 
  addProviderTiles(providers$Esri.WorldStreetMap) |> 
  setView(lng = -071.9402820, lat = 42.3116934, zoom = 9.4) |> 
  addPolygons(
 fillColor=~pal(total_population_tractE),
 weight=2,
 opacity=0.5,
 color= "white",
 label=~mytext,
 dashArray = "3",
 fillOpacity=0.05,
 highlightOptions = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.5,
    bringToFront = FALSE)) 


```



```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(map_tract_population_plt, "map_tract_population_plt.rds")

```

```{r}
#| include: false

map_tract_population_plt <- readRDS("map_tract_population_plt.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false




tract_map_population<-map_tract_population_plt %>% 
  addLegend(pal = pal, values = ~total_population_tractE, opacity = 0.6, title = "Massachusetts Census Tract Population",
  position = "bottomright")

```

```{r}
#| echo: false
#| include: false
#| eval: false



saveRDS(tract_map_population, "tract_map_population.rds")

```

```{r}
#| include: false


tract_map_population <- readRDS("tract_map_population.rds")


```

```{r}


tract_map_population
  

```


```{r}
#| code-fold: true
#| code-summary: "Show Code"



ma_census4326 |> 
    filter(GEOID == "25017368102")


```

```{r}






```
















# Preprocessing


```{r}



```


#rows / columns 
10:1 rule 
You should have at least 10 rows of data for every 1 column (feature) in your dataset. 



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


census_tract_df <- ma_accidents_16_21_8 |> 
    left_join(ma_tract_data_7, by = "GEOID")

```

```{r}
#| echo: false
#| include: false
#| eval: false



saveRDS(census_tract_df, "census_tract_df.rds")


```


```{r}
#| include: false


census_tract_df <- readRDS("census_tract_df.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



census_tract_df |> 
   summary()


```

```{r}

census_tract_df  |> filter(is.na(total_population_tractE))  |> count(GEOID)


```

```{r}

census_tract_df |> filter(total_tract_housing_unitsE == 0) |> head(10)


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


census_tract_df_2 <- census_tract_df |> 
    select(-ID, -Severity, -City, -County, -Year, -LATITUDE, -LONGITUDE, -census_code, -tract_house_density_sqmi, -Stop,-Amenity, -Crossing,-Junction ,-Station,-Traffic_Signal,-Sunrise_Sunset,-Civil_Twilight, -day_of_week, -total_tract_housing_unitsE, -total_vehicles_tractE, -total_population_tractE, -avg_cars_per_household, -total_households_tractE, -tract_pop_density_sqmi)


```

```{r}
#| echo: false
#| include: false
#| eval: false



saveRDS(census_tract_df_2, "census_tract_df_2.rds")


```

```{r}
#| include: false


census_tract_df_2 <- readRDS("census_tract_df_2.rds")



```

## Data Possion Regression

<br><br>



Our next data preparation will be for a poisson regression model.  Unlike our linear regression models, which predicted a continuous variable, the poisson regression model will predict count data.  In our case, we will predict the number of accidents that occurred within a census tract.  Due to this difference, we will prepare our data differently. 

<br><br>


```{r}


census_tract_df_3 <- census_tract_df |> 
    select(-ID, -Severity, -City, -County, -Year, -LATITUDE, -LONGITUDE, -census_code, -tract_house_density_sqmi, -Stop,-Amenity, -Crossing,-Junction ,-Station,-Traffic_Signal,-Sunrise_Sunset,-Civil_Twilight, -day_of_week, -total_tract_housing_unitsE, -total_vehicles_tractE, -avg_cars_per_household, -total_households_tractE, -tract_pop_density_sqmi)



```

```{r}

saveRDS(census_tract_df_3 , "census_tract_df_3.rds")


```

```{r}

census_tract_df_3  <- readRDS("census_tract_df_3.rds")



```


<br><br>


```{r}



census_tract_df_3 |> summary()

```


<br><br>

The total_population_tractE feature has 787 missing values (NAs).  Our model does not accept missing values; thus, we will remove them from the data. 



<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


census_tract_df_4 <- census_tract_df_3 |> 
    filter(!is.na(total_population_tractE))


```

```{r}

saveRDS(census_tract_df_4, "census_tract_df_4.rds")


```

```{r}

census_tract_df_4 <- readRDS("census_tract_df_4.rds")



```

<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"


census_tract_df_4 |> summary()


```

<br><br>

Now that missing values have been removed from total population, we must ensure that there are no duplicates for the GEOID, population combination.  We will use the base R function any() for this. 

<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"



tract_pop <- census_tract_df_4 |> 
    select(GEOID, total_population_tractE )


```

<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"



any(duplicated(tract_pop$GEOID))


```

<br><br>

The response TRUE, means that there are duplicates of the GEOID, Population combination.   We will use the distinct() function from the dplyr library to only include non-duplicates in the tract_pop data. 

<br><br>



```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


tract_pop <- tract_pop |> 
    distinct(GEOID, .keep_all = TRUE)


```

<br><br>

```{r}

saveRDS(tract_pop , "tract_pop.rds")


```

```{r}


tract_pop <- readRDS("tract_pop.rds")


```

<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"




any(duplicated(tract_pop$GEOID))


```

<br><br>

We must now group and aggregate our categorical features (GEOID, Month, day_type) and calculate the mean of our numeric feature (temperature). This will create an accident count grouped by the categorical features.   The summarize() function from dplyr library will be used to accomplish this

<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



agg_observed <- census_tract_df_4 |> 
    group_by(GEOID, Month, day_type) |> 
    summarize(
        accident_count = n(),
        mean_temp = mean(temperature, na.rm =TRUE),
        .groups = "drop"
    )


```

<br><br>

```{r}

saveRDS(agg_observed, "agg_observed.rds")


```

```{r}



agg_observed <- readRDS("agg_observed.rds")

```

<br><br>


Next, we want to ensure that all rows of the combination GEOID, Month, and day_type, for accident_count, have no NAs but an actual value.  The complete() function from tidyr will be used for this task.  We will call complete() with the names of our combination of features and fill in the missing values of accident_count. 

Additionally, we will also ensure that the mean_temperature feature has not missing values.  We will then scale the mean_temperature using the scale() function from base R. 



<br><br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



census_poi_df <- agg_observed |> 
    complete(
        GEOID, Month, day_type, 
        fill = list(accident_count = 0)
    ) |> 
    group_by(Month) |> 
    mutate(mean_temp = if_else(is.na(mean_temp), mean(mean_temp, na.rm = TRUE), mean_temp)) |> 
    ungroup() |> 
    mutate(temp_z = as.numeric(scale(mean_temp)))


```


<br><br>

```{r}

saveRDS(census_poi_df, "census_poi_df.rds")


```

```{r}



census_poi_df <- readRDS("census_poi_df.rds")

```

<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"




census_poi_df |> summary()



```


<br><br>

The census_poi_df and tract_pop will now be joined.


<br><br>


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


census_poi_df_2 <- census_poi_df |> 
    left_join(tract_pop, by = "GEOID", relationship = "many-to-one") |> 
    filter(!is.na(total_population_tractE) & total_population_tractE > 0)



```

```{r}


saveRDS(census_poi_df_2, "census_poi_df_2.rds")

```

```{r}


census_poi_df_2 <- readRDS("census_poi_df_2.rds")


```









```{r}
#| echo: false
#| include: false
#| eval: false



census_tract_df_2 |> 
    ggplot(aes(temperature)) +
    geom_histogram(bins=25)


```

```{r}
#| echo: false
#| include: false
#| eval: false


census_tract_df_2 |> 
    ggplot(aes(log(wind_speed))) +
    geom_histogram(bins=25)


```


```{r}
#| echo: false
#| include: false
#| eval: false


census_tract_df_2 |> 
    filter(accident_duration < 5000) |> 
    ggplot(aes(accident_duration,log(temperature))) +
    geom_point()


```

```{r}
#| echo: false
#| include: false
#| eval: false



census_tract_df_2 |> 
    mutate(GEOID = as.factor(GEOID)) |> 
    count(GEOID) |>
    arrange(desc(n)) 



```





## Train/Test Split

### Linear and Linear Mized Effect Split

```{r}
#| code-fold: true
#| code-summary: "Show Code"


acci_census_df <- initial_split(census_tract_df_2, prop = 0.80)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"

print(acci_census_df)



```

### Poisson Split

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


acci_census_df_2 <- initial_split(census_poi_df_2, prop = 0.80)


```

```{r}

saveRDS(acci_census_df_2, "acci_census_df_2.rds")


```

```{r}


acci_census_df_2 <- readRDS("acci_census_df_2.rds")


```






### Linear train/test

```{r}
#| code-fold: true
#| code-summary: "Show Code"



train_census_data <- training(acci_census_df)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"

dim(train_census_data)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


test_census_data <- testing(acci_census_df)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"

dim(test_census_data)


```



### Poisson train/test

```{r}
#| code-fold: true
#| code-summary: "Show Code"

train_census_data_2 <- training(acci_census_df_2)


```


```{r}

saveRDS(train_census_data_2, "train_census_data_2.rds")

```

```{r}

train_census_data_2 <- readRDS("train_census_data_2.rds")



```

```{r}


test_census_data_2 <- testing(acci_census_df_2)


```

```{r}

saveRDS(test_census_data_2, "test_census_data_2.rds")

```

```{r}


test_census_data_2 <- readRDS("test_census_data_2.rds")


```






### Linear scale and center
```{r}
#| code-fold: true
#| code-summary: "Show Code"



census_rec <- recipe(accident_duration ~ ., data = train_census_data) |> 
    step_center(all_numeric_predictors()) |> 
    step_scale(all_numeric_predictors())


```






### Linear

```{r}
#| code-fold: true
#| code-summary: "Show Code"


train_census_rec <- prep(census_rec, training= train_census_data)
scaled_census_train <- bake(train_census_rec, new_data = train_census_data)


```






# Linear Fixed/Mixed Effects Models

## Linear Regression

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



lr <- lm(log(accident_duration) ~ . -GEOID, data = train_census_data)


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(lr, "lr.rds")

```

```{r}
#| include: false


lr <- readRDS("lr.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



summary(lr)



```


```{r}


lr_resid <- data.frame(residuals = residuals(lr))

```



```{r}


lr_qq <- ggplot(lr_resid, aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(
    title = "LR Model Normal Q-Q Plot of Residuals",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  )
```

```{r}


lr_qq

```

```{r}



lr_resid_data <- data.frame(
    Fitted = fitted(lr),
    Residuals = resid(lr)
)


```

```{r}


lr_residual_plot <- ggplot(lr_resid_data, aes(x = Fitted, y = Residuals)) +
    geom_point(color = "steelblue", size =2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    labs(title = "lr Model Residual Plot", x = "Fitted Values", y = "Residuals" ) +
    theme_minimal()

```

```{r}


lr_residual_plot

```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


lr_2 <- lm(log(accident_duration) ~ . -GEOID, data = train_census_data_2)


```



```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(lr_2, "lr_2.rds")



```

```{r}
#| include: false


lr_2 <- readRDS("lr_2.rds")

```

```{r}

summary(lr_2)

```

```{r}


lr_2_resid <- data.frame(residuals = residuals(lr_2))


```

```{r}

lr_2_qq <- ggplot(lr_2_resid, aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(
    title = "LR_2 Model Normal Q-Q Plot of Residuals",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  )

```

```{r}

lr_2_qq

```

```{r}

lr_2_resid_data <- data.frame(
    Fitted = fitted(lr_2),
    Residuals = resid(lr_2)
)

```

```{r}

lr_2_residual_plot <- ggplot(lr_2_resid_data, aes(x = Fitted, y = Residuals)) +
    geom_point(color = "steelblue", size =2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    labs(title = "lr Model Residual Plot", x = "Fitted Values", y = "Residuals" ) +
    theme_minimal()

```

```{r}

lr_2_residual_plot 

```






## Bayesian Linear Mixed Effects


<br><br>




```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



brms_model <- brm(
  log(accident_duration) ~ . -GEOID + (temperature | GEOID),
  data = scaled_census_train,
  chains = 4,
  iter = 12000,
  family = student(),
  cores = 40, 
  seed = 42
)



```


```{r}




```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(brms_model, "brms_model.rds")


```


```{r}
#| include: false


brms_model <- readRDS("brms_model.rds")


```





```{r}
#| code-fold: true
#| code-summary: "Show Code"


summary(brms_model)



```

```{r}
#| code-fold: true
#| code-summary: "Show Code"




performance::check_singularity(brms_model)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false

performance::check_collinearity(brms_model)


```



```{r}


icc(brms_model, nrep=50)


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


brms_model_2 <- brm(
  log(accident_duration) ~ . -GEOID -Wind_Direction -Weather_Condition + (temperature | GEOID),
  data = scaled_census_train,
  chains = 4,
  iter = 8000,
  family = student(),
  cores = 40, 
  seed = 42
)



```

```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(brms_model_2, "brms_model_2.rds")


```

```{r}
#| include: false


brms_model_2 <- readRDS("brms_model_2.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



summary(brms_model_2)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



performance::check_collinearity(brms_model_2)


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



brms_model_3 <- brm(
  log(accident_duration) ~ . -GEOID -Wind_Direction -Weather_Condition + temperature * Month + (temperature | GEOID),
  data = scaled_census_train,
  chains = 4,
  iter = 8000,
  family = student(),
  cores = 16, 
  seed = 42
)


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(brms_model_3, "brms_model_3.rds")


```

```{r}
#| include: false


brms_model_3 <- readRDS("brms_model_3.rds")



```

```{r}


summary(brms_model_3)


```











```{r}
#| echo: false
#| include: false
#| eval: false


brms_model_3 <- brm(
  log(accident_duration) ~ . -GEOID -Wind_Direction -Weather_Condition + temperature * Month + hour_day_sections * day_type + (temperature | GEOID),
  data = scaled_census_train,
  chains = 4,
  iter = 8000,
  family = student(),
  cores = 16, 
  seed = 42
)


```

```{r}
#| echo: false
#| include: false
#| eval: false


summary(brms_model_3)


```



```{r}



brms_model_cat_plot_top_60 <- brms_model_2 %>%
  spread_draws(r_GEOID[GEOID, Intercept]) %>%
  median_qi(.width = 0.95) %>%
  slice_max(order_by = abs(r_GEOID), n = 60) %>%
  ggplot(aes(x = r_GEOID, y = reorder(GEOID, r_GEOID), xmin = .lower, xmax = .upper)) +
  geom_pointinterval(color = "#1f78b4") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title = "Top 60 Most Extreme Census Tract Effects (brms_model)",
    x = "Log Accident Duration Adjustment",
    y = "Census Tract (GEOID)"
  )


```



```{r}


brms_model_cat_plot_top_60_2 <- brms_model_2 %>%
  spread_draws(r_GEOID[GEOID, Intercept]) %>%
  median_qi(.width = 0.95) %>%
  slice_max(order_by = abs(r_GEOID), n = 100) %>%
  ggplot(aes(x = r_GEOID, y = reorder(GEOID, r_GEOID), xmin = .lower, xmax = .upper)) +
  geom_pointinterval(color = "#1f78b4") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title = "Top 60 Most Extreme Census Tract Effects (brms_model)",
    x = "Log Accident Duration Adjustment",
    y = "Census Tract (GEOID)"
  )



```

```{r}



brms_model_cat_plot_top_60


```

```{r}


brms_model_cat_plot_top_60_2


```

```{r}


ma_census4326 |> 
    filter(GEOID %in% c("25021418004",  "25001011802","25005644200", "25017321500", "25021416101", "25021421200","25023507101", "25017312000", "25005630202","25017315401", "25021421100", "25017373902",
                         "25025130402", "25017358700", "25009211201","25025101001", "25025100800", "25025140300","25023520100", "25009208401", "25025981100", "25023502101")) |> 
                            count(County) |> 
                            arrange(desc(n)) |> 
                            st_drop_geometry()


```

```{r}


ma_census4326 |> 
    filter(GEOID %in% c("25021418004",  "25001011802","25005644200", "25017321500", "25021416101", "25021421200","25023507101", "25017312000", "25005630202","25017315401", "25021421100", "25017373902",
                         "25025130402", "25017358700", "25009211201","25025101001", "25025100800", "25025140300","25023520100", "25009208401", "25025981100", "25023502101")) |> 
                            count(County,GEOID) |> 
                            arrange(desc(n)) |> 
                            st_drop_geometry()



```

```{r}


ma_census4326 |> 
    filter(GEOID %in% c( "25017320105",  "25021419700", "25021419800", "25009254401", "25025010600", "25027744103", "25017368400","25013802101", "25025170301", "25025170400", "25017350105", "25013810100",
                        "25017320106", "25021411200", "25025081001", "25027732901","25017358600","25017374800", "25005614102", "25017337300", "25027710200", "25021403100", "25017342501", "25023501204",
                          "25017383700", "25017384003", "25017374200", "25025050901", "25021410100","25021414100")) |> 
                            count(County) |> 
                            arrange(desc(n)) |> 
                            st_drop_geometry()


```

```{r}


ma_census4326 |> 
    filter(GEOID %in% c( "25017320105",  "25021419700", "25021419800", "25009254401", "25025010600", "25027744103", "25017368400","25013802101", "25025170301", "25025170400", "25017350105", "25013810100",
                        "25017320106", "25021411200", "25025081001", "25027732901","25017358600","25017374800", "25005614102", "25017337300", "25027710200", "25021403100", "25017342501", "25023501204",
                          "25017383700", "25017384003", "25017374200", "25025050901", "25021410100","25021414100")) |> 
                            count(County, GEOID) |> 
                            arrange(desc(n)) |> 
                            st_drop_geometry()

```


```{r}

ma_census_middle <- ma_census4326 |> 
    filter(County == "Middlesex")


```

```{r}


saveRDS(ma_census_middle, "ma_census_middle.rds")


```

```{r}
#| include: false


ma_census_middle <- readRDS("ma_census_middle.rds")


```

```{r}

ma_extreme_middle_higher <- ma_census_middle |> 
    select(GEOID, ID, accident_duration, County, City, hour_day_sections, day_type, Month,total_vehicles_tractE,road_category, LONGITUDE, LATITUDE) |> 
    filter(GEOID %in% c("25017321500","25017358700", "25017315401", "25017373902", "25017312000" )) |> 
        st_drop_geometry()

```

```{r}


saveRDS(ma_extreme_middle_higher, "ma_extreme_middle_higher.rds")


```

```{r}
#| include: false

ma_extreme_middle_higher <- readRDS("ma_extreme_middle_higher.rds")


```

```{r}

bins<-c(0,200,4000,5400,6500,7500,8500,Inf)

```





```{r}

pal<-colorBin(palette="RdYlBu",
            domain=ma_census_middle$total_population_tractE,bins=bins)



```












```{r}


ma_tract_middle_extreme_formatted <- ma_extreme_middle_higher |> 
    mutate(accident_duration = format(accident_duration, big.mark = ",", big.interval = 3, digits=12) ) 
     


```

```{r}

saveRDS(ma_tract_middle_extreme_formatted, "ma_tract_middle_extreme_formatted.rds")

```

```{r}

ma_tract_middle_extreme_formatted <- readRDS("ma_tract_middle_extreme_formatted.rds")


```

```{r}



mytext <- paste(
  "Census Tract: ", ma_census_middle$GEOID, "<br/>",
  "Total Vehicles Census Tract: ", ma_census_middle$total_vehicles_tractE, "<br/>",
  "Total Housing Census Tract: ", ma_census_middle$total_tract_housing_unitsE,"<br/>", 
  "Census Tract Accident County: ", ma_census_middle$tract_accident_count,"<br/>",
    sep="") %>%
  lapply(htmltools::HTML)


```

```{r}




map_tract_extreme_middle_plt <- ma_census_middle |> leaflet(options = leafletOptions(minZoom = 11, maxZoom = 18)) |> 
  addProviderTiles(providers$Esri.WorldStreetMap) |> 
  setView(lng = -071.23928, lat = 42.38980, zoom = 11.5) |> 
  addPolygons(
 fillColor=~pal(total_population_tractE),
 weight=2,
 opacity=0.8,
 color= "white",
 label=~mytext,
 dashArray = "3",
 fillOpacity=0.05,
 highlightOptions = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.2,
    bringToFront = FALSE)) |> 
  addMarkers(
    data = ma_tract_middle_extreme_formatted,
    lng = ~LONGITUDE,
    lat = ~LATITUDE,
    popup = ~paste("Accident ID: ", ID,
                   "<br>Accident Duration: ", accident_duration, 
                   "<br>City: ", City, 
                   "<br>Weekday/Weekend: ", day_type,
                   "<br>Time of Day: ", hour_day_sections,
                   "<br>Road Type: ", road_category) )     

```

```{r}


tract_map_extreme_middle<-map_tract_extreme_middle_plt %>% 
  addLegend(pal = pal, values = ~total_population_tractE, opacity = 0.6, title = "Massachusetts Extreme Census Tract Population",
  position = "bottomright")


```

```{r}

tract_map_extreme_middle


```

```{r}


bins_2<-c(0,200,4000,5400,6500,7500,8500,Inf)


```

```{r}


ma_census_norfolk <- ma_census4326 |> 
    filter(County == "Norfolk")


```

```{r}

saveRDS(ma_census_norfolk, "ma_census_norfolk.rds")

```

```{r}
#| include: false


ma_census_norfolk <- readRDS("ma_census_norfolk.rds")


```

```{r}



ma_extreme_norfolk_higher <- ma_census_norfolk |> 
    select(GEOID, ID, accident_duration, County, City, hour_day_sections, day_type, Month,total_vehicles_tractE,road_category, LONGITUDE, LATITUDE) |> 
    filter(GEOID %in% c("25021421200","25021421100","25021416101", "25021418004")) |> 
        st_drop_geometry()


```

```{r}

saveRDS(ma_extreme_norfolk_higher, "ma_extreme_norfolk_higher.rds")

```

```{r}
#| include: false

ma_extreme_norfolk_higher <- readRDS("ma_extreme_norfolk_higher.rds")


```

```{r}

pal_2 <- colorBin(palette="RdYlBu",
            domain=ma_census_norfolk$total_population_tractE,bins=bins_2)


```

```{r}


ma_tract_norfolk_extreme_formatted <- ma_extreme_norfolk_higher |> 
    mutate(accident_duration = format(accident_duration, big.mark = ",", big.interval = 3, digits=12) )


```

```{r}

saveRDS(ma_tract_norfolk_extreme_formatted, "ma_tract_norfolk_extreme_formatted.rds")


```

```{r}
#| include: false


ma_tract_norfolk_extreme_formatted <- readRDS("ma_tract_norfolk_extreme_formatted.rds")


```

```{r}


mytext_2 <- paste(
  "Census Tract: ", ma_census_norfolk$GEOID, "<br/>",
  "Total Vehicles Census Tract: ", ma_census_norfolk$total_vehicles_tractE, "<br/>",
  "Total Housing Census Tract: ", ma_census_norfolk$total_tract_housing_unitsE,"<br/>", 
  "Census Tract Accident County: ", ma_census_norfolk$tract_accident_count,"<br/>",
    sep="") %>%
  lapply(htmltools::HTML)


```

```{r}



map_tract_extreme_norfolk_plt <- ma_census_norfolk |> leaflet(options = leafletOptions(minZoom = 9, maxZoom = 18)) |> 
  addProviderTiles(providers$Esri.WorldStreetMap) |> 
  setView(lng = -71.17127  , lat = 42.17051 , zoom = 11.5) |> 
  addPolygons(
 fillColor=~pal_2(total_population_tractE),
 weight=2,
 opacity=0.5,
 color= "white",
 label=~mytext_2,
 dashArray = "3",
 fillOpacity=0.05,
 highlightOptions = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.2,
    bringToFront = FALSE)) |> 
  addMarkers(
    data = ma_tract_norfolk_extreme_formatted,
    lng = ~LONGITUDE,
    lat = ~LATITUDE,
    popup = ~paste("Accident ID: ", ID,
                   "<br>Accident Duration: ", accident_duration, 
                   "<br>City: ", City, 
                   "<br>Weekday/Weekend: ", day_type,
                   "<br>Time of Day: ", hour_day_sections,
                   "<br>Road Type: ", road_category) ) 


```

```{r}


tract_map_extreme_norfolk<-map_tract_extreme_norfolk_plt %>% 
  addLegend(pal = pal_2, values = ~total_population_tractE, opacity = 0.6, title = "Norfolk Extreme Census Tract Population",
  position = "bottomright")


```

```{r}

tract_map_extreme_norfolk


```

```{r}



print(ma_census_norfolk  |> select(GEOID, County), n=1500)


```


```{r}




brms_extreme_df_2 <- brms_model_2 %>%
  spread_draws(r_GEOID[GEOID, Intercept]) %>%
  median_qi(.width = 0.95) %>%
  slice_max(order_by = abs(r_GEOID), n = 100)


```

```{r}


print(brms_extreme_df_2, n=100)


```







# Bayesian Possion Mixed Effect Regression (Count Data) Models




In our model, the prior argument includes:
prior(normal(0,0.5), class = "b" - this  regularizes prior for fixed effects.

prior(exponential(1), class = "sd") - This is the prior for census tract varying intercept. 


prior(gamma(0.01, 0.01), class = "shape") - The negative binomial distribution estimates an additional parameter (shape) that allows the conditioinal variance to scale quadratically. 





```{r}



brms_poisson_2 <- brm(
    formula = accident_count ~ offset(log(total_population_tractE)) + temp_z + Month + day_type + (1|GEOID),
    data = train_census_data_2,
    family = negbinomial(link = "log"),
    prior = c(
        prior(normal(0, 0.5), class = "b"),
        prior(exponential(1), class = "sd"),
        prior(gamma(0.01, 0.01), class = "shape")
    ),
    chains = 4,
    cores = 12,
    iter = 12000

)


```



```{r}


saveRDS(brms_poisson_2, "brms_poisson_2.rds")


```

```{r}


brms_poisson_2 <- readRDS("brms_poisson_2.rds")


```

```{r}


ppcheck_poi_2 <- pp_check(brms_poisson_2, type = "dens_overlay", ndraws = 50)


```

```{r}


ppcheck_poi_2


```

```{r}



summary(brms_poisson_2)


```



```{r}

tract_ranef <- ranef(brms_poisson_2)$GEOID[, , "Intercept"] |> 
    as.data.frame() |> 
    rownames_to_column(var = "GEOID") |> 
    rename(
        ranef_log = Estimate,
        se_log = Est.Error,
        lower_95 = Q2.5,
        upper_95 = Q97.5
    ) |> 
    mutate(
        rate_ratio = exp(ranef_log),
        rate_ratio_lower = exp(lower_95),
        rate_ratio_upper = exp(upper_95)
    )   





```

```{r}



head(tract_ranef)

```


```{r}

boundaries = ma_tract_data_6 |> 
    select(GEOID, census_tract, geometry)

```

```{r}

saveRDS(boundaries, "boundaries.rds")

```

```{r}

boundaries <- readRDS("boundaries.rds")


```

```{r}

boundaries_clean <- boundaries %>%
  mutate(GEOID = as.character(GEOID)) %>%
  st_make_valid()

```

```{r}


saveRDS(boundaries_clean, "boundaries_clean.rds")

```

```{r}


boundaries_clean <- readRDS("boundaries_clean.rds")


```


Joining spatial polygons with random effect rate ratios. 

```{r}



map_sf <- boundaries_clean |> 
    left_join(tract_ranef, by = "GEOID")


``` 



```{r}




```

```{r}






```

```{r}


map_sf



```



We'kll filterthe map_sf dataset to tracts that have estimated random effects. 

```{r}



map_analysis <- map_sf |> 
    filter(!is.na(ranef_log))


```

```{r}




```

```{r}





```

```{r}


map_analysis



```



We'll use the poly2nb function from the spdep library to build a adjacency list (Queen contiguity: sharing a border or corner point)

```{r}


neighbors <- poly2nb(map_analysis, queen = TRUE)


```

```{r}






```

```{r}





```

```{r}


neighbors


```



We have observations with no neighbors. To handle these islands/isolated tracts (tracts with 0 neighbors). We will use the nb2listw function from the spdep library to allow tracts without neighbors to remain without crashing. 


```{r}

weights_list <- nb2listw(neighbors, style = "W", zero.policy = TRUE)



```

```{r}





```

```{r}





```

```{r}


weights_list


```


```{r}



moran_analytical <- moran.test(
    map_analysis$ranef_log,
    weights_list,
    zero.policy = TRUE
)


```

```{r}






```

```{r}




```

```{r}




moran_analytical

```

Analyzing the Moran's I test on log-scale random effects:

The p-valueis less than 0.05 and I is greater than 0: This means there is a statisrically signficant spatital splillover between neighboring census tracts. An independent random  intercept treates all tracts as exhangeable (assuming trac A shares no more similarity with its immediate neighbour tha with a tract 50 miles away). 








## Conditional Autoregressive Model (CAR)



The Moran's I test detected residual spatial clustering in our first Bayesian mixed effects poisson model.  As such, we will build a conditional autoregressive (CAR) Our first bayesian mixed effect model


### CAR data preparation



The st_make_valid function is from the sf library. 

```{r}



map_clean <- map_analysis |> 
    st_make_valid() |> 
    mutate(GEOID == as.character(GEOID))


```

Quen continguity neighbors list

```{r}


nb <- poly2nb(map_analysis, queen = TRUE)


```



Creating a standardized row weights (zero.ploicy handles islands like peninsulas etc..)

```{r}


lw <- nb2listw(nb, style = "W", zero.policy = TRUE)


```





Check for osland polygons (0 neighbors)i
card() function is fron the spdep library


```{r}


island_indices <- which(card(nb) == 0)


```

Next, we will identify and remove any islands. 

```{r}


if(length(island_indices) > 0) {
    cat("Dropping", length(island_indices), "islands with zero neighbors. \n")
    map_clean <- map_clean[-island_indices, ]
    nb <- poly2nb(map_clean, queen = TRUE)
}



```

```{r}


W_bin <- nb2mat(nb_statewide, style = "B", zero.policy = TRUE)
rownames(W_bin)


```

Our output show that one island with zero neighbors was dropped. 

<br><br>

Next, we'll create a strictly binary matrix (style = "B") using the nb2mat function from spdep. 


style = "B" is used for binary 1/0 adjacency, which is required by brms CAR models. 


```{r}

W_bin <- nb2mat(nb, style = "B", zero.policy = TRUE)



```
```{r}

rownames(W_bin) <- map_clean$GEOID
colnames(W_bin) <- map_clean$GEOID


```

```{r}

rownames(W_bin) <- map_statewide$GEOID

colnames(W_bin) <- map_statewide$GEOID


```




Verify symmetry and diagonal is 0. 

stopifnot() and all() are functions from base R
isSymmetric() and diag() are functions from the matrix library. 

```{r}


stopifnot(isSymmetric(W_bin))
stopifnot(all(diag(W_bin) == 0))


```


```{r}


common_geoids <- sort(intersect(unique(boundaries_clean$GEOID), unique(train_census_df_3$GEOID)))

```

```{r}


map_statewide <- boundaries_clean |> 
    filter(GEOID %in% common_geoids) |> 
    arrange(GEOID)


```

```{r}


nb_statewide <- poly2nb(map_statewide, queen = TRUE)


```

```{r}


main_component_id <- which.max(table(comp$comp.id))
mainland_indices <- which(comp$comp.id == main_component_id)

map_mainland <- map_statewide[mainland_indices, ]
nb_mainland  <- poly2nb(map_mainland, queen = TRUE)
W_mainland   <- nb2mat(nb_mainland, style = "B")


```

```{r}


islands <- which(card(nb_statewide) == 0)

```

```{r}




if (length(islands) > 0) {
    island_geoids <- map_statewide$GEOID[islands]
    cat("Dropping", length(islands), "island tract(s) with 0 neighbors:\n", paste)island_geoids, collapse = ", "), "\n")

    map_statewide <- map_statewide[-islands, ]

    nb_statewide <- poly2nb(map_statewide, queen = TRUE)


```

```{r}


train_census_df_3 <- train_census_data_2 |> 
    filter(GEOID %in% rownames(W_bin)) |> 
    mutate(GEOID = factor(GEOID, levels = rownames(W_bin)))

```

```{r}


saveRDS(train_census_df_3, "train_census_df_3.rds")

```

```{r}


train_census_df_3 <- readRDS("train_census_df_3.rds")


```

```{r}







```



we'll 


```{r}


brms_poisson_3 <- brm(
    formula = accident_count ~ offset(log(total_population_tractE)) + temp_z + Month + day_type + car(W_bin, gr = GEOID, type = "icar"),
    data = train_census_df_3,
    data2 = list(W_bin = W_bin),
    family = negbinomial(link = "log"),
    prior = c(
        prior(normal(0, 0.5), class = "b"),
        prior(exponential(1), class = "sdcar")
    ),
    chains = 4,
    cores = 16,
    iter = 10000,
    warmup = 5000,
    control = list(
      adapt_delta = 0.98,
      max_treedepth = 12)

)


```

```{r}


saveRDS(brms_poisson_3 , "brms_poisson_3.rds")


```

```{r}


brms_poisson_3  <- readRDS("brms_poisson_3.rds")


```

```{r}




summary(brms_poisson_3)

```

```{r}


pp_checl(brms_poisson_3)


```


```{r}



loo_compare(loo(brms_poisson_2), loo(brms_poisson_3)



```


 Extract ICAR tract random effects

```{r}


spatial_effects <- summary(statewide_car_fit)$random$GEOID 

```

If pulling directly via posterior draws:

```{r}




library(posterior)

icar_draws <- as_draws_df(brms_poisson_3) %>%
  subset_draws(variable = "^rcar\\[", regex = TRUE)


```

```{r}

saveRDS(icar_draws, "icar_draws.rds")

```

```{r}


icar_draws <- readRDS("icar_draws.rds")


```




```{r}




library(readr)

icar_summary <- summarise_draws(
  icar_draws,
  "mean",
  ~quantile(.x, probs = c(0.025, 0.975))
) %>%
  mutate(
    # Extracts the integer directly from "rcar[123]"
    idx = parse_number(variable),
    GEOID = levels(train_census_df_3$GEOID)[idx],
    spatial_RR = exp(mean),
    spatial_RR_lower = exp(`2.5%`),
    spatial_RR_upper = exp(`97.5%`)
  ) %>%
  select(GEOID, spatial_log = mean, spatial_RR, spatial_RR_lower, spatial_RR_upper)




```

```{r}


saveRDS(icar_summary, "icar_summary.rds")


```



```{r}

icar_summary <- readRDS("icar_summary.rds")



```

```{r}



head(icar_summary)

```

```{r}


plot_df <- icar_summary %>%
  arrange(spatial_RR) %>%
  mutate(
    rank = row_number(),
    # Categorize whether tract is significantly higher, lower, or average
    significance = case_when(
      spatial_RR_lower > 1.0 ~ "Significantly Higher Risk",
      spatial_RR_upper < 1.0 ~ "Significantly Lower Risk",
      TRUE                   ~ "Average Risk"
    ),
    significance = factor(significance, levels = c(
      "Significantly Lower Risk", "Average Risk", "Significantly Higher Risk"
    ))
  )




```

```{r}

saveRDS(plot_df, "plot_df.rds")


```

```{r}


plot_df <- readRDS("plot_df.rds")


```



```{r}




ggplot(plot_df, aes(x = rank, y = spatial_RR, color = significance)) +
  # Null reference line at RR = 1.0
  geom_hline(yintercept = 1, linetype = "dashed", color = "black", linewidth = 0.6) +
  # 95% Credible interval error bars
  geom_linerange(aes(ymin = spatial_RR_lower, ymax = spatial_RR_upper), 
                 alpha = 0.35, linewidth = 0.3) +
  # Posterior median/mean point estimate
  geom_point(size = 0.5, alpha = 0.7) +
  scale_y_log10(
    breaks = c(0.1, 0.25, 0.5, 1.0, 2.0, 4.0, 8.0),
    labels = c("0.1x", "0.25x", "0.5x", "1.0x", "2.0x", "4.0x", "8.0x")
  ) +
  scale_color_manual(
    values = c(
      "Significantly Lower Risk"  = "#2166ac", # Cool Blue
      "Average Risk"              = "grey60",  # Neutral Gray
      "Significantly Higher Risk" = "#b2182b"  # Crimson Red
    ),
    name = "Tract Classification"
  ) +
  labs(
    title = "Statewide Census Tract Spatial Risk (ICAR Model)",
    subtitle = "Rank-ordered tract rate ratios (RR) with 95% Credible Intervals",
    x = "Census Tract Rank (Lowest to Highest Risk)",
    y = "Spatial Rate Multiplier (vs Statewide Mean, log scale)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey40", size = 10)
  )



```

```{r}


extremes_df <- plot_df %>%
  filter(rank <= 40 | rank > (max(rank) - 40)) %>%
  mutate(GEOID = factor(GEOID, levels = GEOID))



```

```{r}


saveRDS(extremes_df, "extremes_df.rds")

```

```{r}

extremes_df <- readRDS("extremes_df.rds")



```



```{r}


ggplot(extremes_df, aes(x = GEOID, y = spatial_RR, color = significance)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "black", linewidth = 0.5) +
  geom_pointrange(aes(ymin = spatial_RR_lower, ymax = spatial_RR_upper), size = 0.4) +
  scale_y_log10() +
  scale_color_manual(values = c(
    "Significantly Lower Risk"  = "#2166ac",
    "Significantly Higher Risk" = "#b2182b"
  )) +
  coord_flip() +
  labs(
    title = "Top 40 Highest & Lowest Risk Census Tracts",
    x = "Tract GEOID",
    y = "Spatial Rate Multiplier (log scale)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")


```

```{r}



head(icar_summary)


```




```{r}


common_geoids <- sort(intersect(unique(boundaries_clean$GEOID), unique(train_census_df_3$GEOID)))


```

```{r}


saveRDS(common_geoids, "common_geoids.rds")

```

```{r}


common_geoids <- saveRDS("common_geoids.rds")


```

```{r}



map_statewide <- boundaries_clean |> 
    filter(GEOID %in% common_geoids) |> 
    arrange(GEOID)


```

Subset spartial polygons strictly to common GEOIDs



```{r}


nb_statewide <- poly2nb(map_statewide, queen = TRUE)


```

```{r}

saveRDS(nb_statewide, "nb_statewide.rds")


```

```{r}



main_component_id <- which.max(table(comp$comp.id))
mainland_indices <- which(comp$comp.id == main_component_id)

map_mainland <- map_statewide[mainland_indices, ]
nb_mainland  <- poly2nb(map_mainland, queen = TRUE)
W_mainland   <- nb2mat(nb_mainland, style = "B")


```

```{r}


islands <- which(card(nb_statewide) == 0)


```


First, remove islands forn the spartial ploygons.

Next, we recaluculate neighbors withoout islands

```{r}



if (length(islands) > 0) {
    island_geoids <- map_statewide$GEOID[islands]
    cat("Dropping", length(islands), "island tract(s) with 0 neighbors:\n", paste)island_geoids, collapse = ", "), "\n")

    map_statewide <- map_statewide[-islands, ]

    nb_statewide <- poly2nb(map_statewide, queen = TRUE)

```

```{r}

saveRDS( map_statewide, " map_statewide.rds")

```

```{r}


saveRDS(nb_statewide,"nb_statewide.rds")


```

```{r}



W_bin <- nb2mat(nb_statewide, style = "B", zero.policy = TRUE)
rownames(W_bin)

```


```{r}


rownames(W_bin) <- map_statewide$GEOID

colnames(W_bin) <- map_statewide$GEOID



```


```{r}

saveRDS(W_bin,"W_bin.rds")


```

```{r}

train_census_df_3 <- train_census_data_2 |> 
    filter(GEOID %in% rownames(W_bin)) |> 
    mutate(GEOID = factor(GEOID, levels = rownames(W_bin)))




```



```{r}


saveRDS(train_census_df_3, "train_census_df_3.rds")


```

```{r}



train_census_df_3 <- readRDS("train_census_df_3.rds")
```

```{r}



map_icar <- map_mainland %>%
  left_join(icar_summary, by = "GEOID")


```

```{r}


saveRDS(map_icar, "map_icar.rds")

```

```{r}


map_icar <- readRDS("map_icar.rds")


```







```{r}





map_leaflet_icar <- st_transform(map_icar, 4326)


```


```{r}


saveRDS(map_leaflet_icar, "map_leaflet_icar.rds")


```

```{r}


map_leaflet_icar <- reawdRDS("map_leaflet_icar.rds")


```

```{r}



max_val <- max(abs(log(map_icar$spatial_RR)), na.rm = TRUE)



```

```{r}

pal_car <- colorNumeric(
  palette = "RdYlBu",
  domain  = c(-max_val, max_val),
  reverse = TRUE  # Red = high risk, Blue = low risk
)



```

```{r}


map_icar_sf <- st_transform(map_icar, 4326)

```

```{r}


saveRDS(map_icar_sf , "map_icar_sf.rds")


```

```{r}


map_icar_sf  <- readRDS("map_icar_sf.rds")


```

```{r}


labels_car <- sprintf(
  "<strong>Tract GEOID:</strong> %s<br/>
   <strong>Spatial Rate Multiplier:</strong> %.2fx<br/>
   <strong>95%% CrI:</strong> [%.2f, %.2f]",
  map_icar_sf$GEOID,
  map_icar_sf$spatial_RR,
  map_icar_sf$spatial_RR_lower,
  map_icar_sf$spatial_RR_upper
) %>% lapply(htmltools::HTML)


```

```{r}


leaflet(map_icar_sf) %>%
  addProviderTiles(providers$Esri.WorldStreetMap) %>%
  addPolygons(
    fillColor = ~pal_car(log(spatial_RR)),
    weight = 0.5,
    opacity = 1,
    color = "white",
    dashArray = "1",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#444",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = labels_car,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "13px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal_car,
    values = c(-max_val, max_val),
    opacity = 0.7,
    title = "Log Spatial RR<br/>(0 = Statewide Avg)",
    position = "bottomright"
  )



```

```{r}




```

```{r}






```


<br><br>

To help with the interpretation of our map, we’ll review the terms used and how they apply to our analysis.  

<br>

Baseline Score

First, there’s the baseline score.  Every census tract starts at that baseline after we factor in its resident population, the month of year, and whether it’s a weekday or weekend.  An analogy would be to think of the baseline as statewide average temperature as 70 degrees. The Spatial Rate Multiplier tells us whether a census tract runs hotter or colder for accidents than expected, while sharing common conditions with its immediate neighbors. 

Baseline = 1.0x:

This census tract has an average accident rate.  It has exactly as many crashes as expected for its population size. 
Multiplier = 1.50x (or 1.5):

This census tract has 50% more crashes than an average Massachusetts census tract with similar population and conditions.

Multiplier = 2.0x (or 2.0):
This census tract has double the baseline crash rate. 

Multiplier = 0.70x (or 0.7):
This census tract has 30% fewer crashes than expected (a lower risk area. 

You’ll find the Multiplier Score when you hove over any census tract. 

Spatial

We call it spatial because traffic doesn’t stop at invisible This model smooths out the noisy one-off spikes by looking at a census tract and its surrounding neighbors together, giving us a true underlying risk map instead of random flukes. 

<br>

Spatial

We call it spatial because traffic doesn’t stop at invisible This model smooths out the noisy one-off spikes by looking at a census tract and its surrounding neighbors together, giving us a true underlying risk map instead of random flukes. 

<br>
Log Spatial RR

On our map you’ll find a legend scale for Log Spatial RR.  This is the shades a census tract base on its log spatial RR.  We use the Log Rate Ratio to make colors look fair on a map. 

 0.0 is the neutral center: It means a census tract is right at the statewide average (Multiplier = 1.0).


Positive numbers (+): Represent areas with above-average risk (warm colors / Red)
Negative numbers (-): Represent areas with below-average risk (cool colors / Blue).

<br>

How to read the values: 

1.0(0.0 on the log scale): Exactly at statewide expected levels. 
> 1.0(Red): Higher-risk accident hotspots (e.g., 1.4x = 40% higher risk. 
< 1.0(Blue): Lower-risk areas (e.g., 0.7 = 30% lower risk).



```{r}






```










```{r}



comp <- n.comp.nb(nb_statewide)

cat("Number of disjoint graph components:", comp$nc, "\n")


```








```{r}



stopifnot(identical(levels(train_census_df_3$GEOID), rownames(W_bin)))

cat("Total connected tracts:", length(rownames(W_bin)), "\n")

cat("Total modeling rows: ", nrow(train_census_df_3), "\n")




```







```{r}


fixef_summary <- fixef(brms_poisson_2)


```

```{r}

irr_table <- exp(fixef_summary[, c("Estimate", "Q2.5", "Q97.5")])



```




