---
title: "Predicting time of auto accidents"
author: Francis Lauder
date: 2026-09-10
format:
  html:
    toc: true
    toc-depth: 3
    toc_float: true
    number-sections: false
    toc-location: left
    embed-resources: true
---

# Introduction
<br><br>


# Data Dictionary

<br><br>


# Libraries

<br>

```{r}
#| code-fold: true
#| code-summary: "Show Code"

library(dplyr)
library(ggplot2)
library(plotly)
library(RColorBrewer)
library(forcats)
library(tidycensus)

library(sf)

library(jsonlite)

library(tigris)

library(stringr)

library(leaflet)
library(htmlwidgets)
library(htmltools)
library(gt)
library(DT)
library(GGally)
library(lme4)
library(lmerTest)
library(rstanarm)
library(brms)
library(broom)
library(lubridate)
library(tidymodels)





```

<br><br>


# Data Inspection and Cleaning

## US Accident Data


```{r}
#| echo: false
#| include: false
#| eval: false


US_accidents <- read.csv("/Users/francislauder/Documents/Data Sets/US Acciidents/US_Accidents_March23.csv")


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(US_accidents, "US_accidents.rds")

```

```{r}
#| echo: false
#| include: false
#| eval: false


US_accidents <- readRDS("US_accidents.rds")


```



```{r}
#| echo: false
#| include: false
#| eval: false



glimpse(US_accidents)


```



```{r}
#| echo: false
#| include: false
#| eval: false


MA_accidents <- US_accidents |> 
    filter(State == "MA")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


MA_accidents <- MA_accidents %>% 
  mutate(Weather_Timestamp = as.Date(Weather_Timestamp))


```


```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(MA_accidents, "MA_accidents.rds")



```

```{r}
#| echo: false


MA_accidents <- readRDS("MA_accidents.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"



MA_accidents |> glimpse()


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


summary(MA_accidents)


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"


dim(MA_accidents)


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"



MA_accidents |> 
    count(Severity) |> 
    mutate(perc = round(n / sum(n),3)) |> 
    arrange(desc(perc))


```



Some features are in the correct data types for use in our analysis.  We will tranform them to the required data types. 



First, Sverity, which is our outcome feature, will be transformed to a factor from an integer data type. 

```{r}
#| code-fold: true
#| code-summary: "Show Code"




MA_accidents_2 <- MA_accidents |> 
    mutate(Severity = as.factor(Severity))


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"


class(MA_accidents_2$Severity)


```




We will now extract the year, month,day, and hourm of the day from the Start_Time feature. 





```{r}
#| code-fold: true
#| code-summary: "Show Code"



MA_accidents_2 <- MA_accidents_2 |> 
    mutate(Year = year(Start_Time)) |> 
    mutate(Month = month(Start_Time)) |> 
    mutate(day_of_week = wday(Start_Time)) |> 
    mutate(hour_of_day = hour(Start_Time))


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"


MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day) |> 
    head()


```

Day of week starts with Sunday (1) and ends with Saturday (7)

Hour of day starts at 1:00 am (1) and ends at 12:00 am (24)




```{r}
#| code-fold: true
#| code-summary: "Show Code"


MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day) |> 
    glimpse()



```



```{r}
#| code-fold: true
#| code-summary: "Show Code"


MA_accidents_2 <- MA_accidents_2 |> 
    mutate(Year = as.factor(Year)) |> 
    mutate(Month = as.factor(Month)) |> 
    mutate(day_of_week = as.factor(day_of_week))



```



```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(MA_accidents_2, "MA_accidents_2.rds")


```

```{r}
#| echo: false


MA_accidents_2 <- readRDS("MA_accidents_2.rds")



```

```{r}

MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day) |> 
    glimpse()


```

```{r}


MA_accidents_2 |> 
    count(Year)


```


Before continuing, we will filter our data into the years 2016-2021 for training and testing, and the year 2022 for predicting on unseen data.  Since the year 2023 appears to be incomplete, we will omit it from our analysis. 

<br>

Years 2016 through 2021


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false


ma_accidents_16_21 <- MA_accidents_2 |> 
    filter(Year %in% c("2016", "2017", "2018", "2019", "2020","2021"))


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_accidents_16_21, "ma_accidents_16_21.rds")

```

```{r}
#| echo: false


ma_accidents_16_21 <- readRDS("ma_accidents_16_21.rds")



```

```{r}


dim(ma_accidents_16_21)


```

Year 2022


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_accidents_22 <- MA_accidents_2 |> 
    filter(Year == "2022")


```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_accidents_22, "ma_accidents_22.rds")


```

```{r}
#| echo: false


ma_accidents_22 <- readRDS("ma_accidents_22.rds")


```

```{r}
#| code-fold: true
#| code-summary: "Show Code"


dim(ma_accidents_22)


```


For us to be able to calculate the time duration of accidents we'll first need to convert the start and endtimes to a date time object that includes year, month, day, hours, minutes, and seconds. 


```{r}
#| code-fold: true
#| code-summary: "Show Code"

ma_accidents_16_21_2 <- ma_accidents_16_21 |> 
    mutate(Start_Time = ymd_hms(Start_Time)) |> 
    mutate(End_Time = ymd_hms(End_Time))
```

```{r}
#| code-fold: true
#| code-summary: "Show Code"

ma_accidents_16_21_2 |> 
    select(Start_Time, End_Time) |> 
    glimpse()


```


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_2 |> 
    select(Start_Time, End_Time) |> 
    head()


```

Next, we'll create a new feature accident_duration, which will calcualte the duration of an accident in minutes. 


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_2 <- ma_accidents_16_21_2 |> 
    mutate(accident_duration = round(difftime(End_Time,Start_Time, units = "mins"), 0))


```

 We also will create LATTITUDE and LONGITUDE features from Start_Lat and Start_Lng. These features will be useful for retieving US Census data and mapping. 




```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_accidents_16_21_2$LATITUDE <- ma_accidents_16_21_2$Start_Lat

ma_accidents_16_21_2$LONGITUDE <- ma_accidents_16_21_2$Start_Lng


```


```{r}
#| echo: false
#| include: false
#| eval: false

saveRDS(ma_accidents_16_21_2, "ma_accidents_16_21_2.rds")


```


```{r}
#| echo: false


 ma_accidents_16_21_2 <- readRDS("ma_accidents_16_21_2.rds")


```



```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_2 |> select(accident_duration) |>  head()


```

From the output, we see the accident duration is in minutes. 


```{r}
#| code-fold: true
#| code-summary: "Show Code"


ma_accidents_16_21_2 |> select(LATITUDE, LONGITUDE) |> head()


```

Our output shows the two new featurres. 



Next, we will create artificial row numbers.  These will be useful for analyzing and filtering visualizations  along with removing row(s).


```{r}
#| code-fold: true
#| code-summary: "Show Code"
#| eval: false



ma_accidents_16_21_3  <- ma_accidents_16_21_2  |> 
    mutate(id = row_number())




```

```{r}
#| echo: false
#| include: false
#| eval: false


saveRDS(ma_accidents_16_21_3 , "ma_accidents_16_21_3.rds")

```

```{r}
#| echo: false


ma_accidents_16_21_3 <- readRDS("ma_accidents_16_21_3.rds")


```

```{r}

ma_accidents_16_21_3 |> 
    select(id) |> 
    head()



```

## US Census Data


```{r}





long_lat_1 <- ma_accidents_16_21_3 |> 
    filter(id <=5000) |> 
    select(LATITUDE, LONGITUDE)



```


```{r}


long_lat_2 <- ma_accidents_16_21_3 |> 
    filter(id > 5000 & id <=15000) |> 
    select(LATITUDE, LONGITUDE)



```


```{r}




long_lat_3 <- ma_accidents_16_21_3 |> 
    filter(id > 15000 & id <=25000) |> 
    select(LATITUDE, LONGITUDE)

```


```{r}


long_lat_4 <- ma_accidents_16_21_3 |> 
    filter(id > 25000 & id <=35000) |> 
    select(LATITUDE, LONGITUDE)


```

```{r}


long_lat_5 <- ma_accidents_16_21_3 |> 
    filter(id > 35000 & id <=45000) |> 
    select(LATITUDE, LONGITUDE)



```

```{r}


long_lat_6 <- ma_accidents_16_21_3 |> 
    filter(id > 45000 & id <=55587) |> 
    select(LATITUDE, LONGITUDE)



```


```{r}


Sys.getenv("CENSUS_API_KEY")


```


```{r}


long_lat_1$census_code <- apply(long_lat_1, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))




```


```{r}


saveRDS(long_lat_1 , "long_lat_1.rds")


```

```{r}


long_lat_1  <- readRDS("long_lat_1.rds")


```


```{r}

long_lat_2$census_code <- apply(long_lat_2, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))


```

```{r}

saveRDS(long_lat_2, "long_lat_2.rds")

```

```{r}


long_lat_2 <- readRDS("long_lat_2.rds")


```


```{r}



long_lat_3$census_code <- apply(long_lat_3, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))


```



```{r}


saveRDS(long_lat_3, "long_lat_3.rds")


```


```{r}


long_lat_3 <- readRDS("long_lat_3.rds")



```


```{r}


long_lat_4$census_code <- apply(long_lat_4, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))


```


```{r}



saveRDS(long_lat_4, "long_lat_4.rds")



```


```{r}


long_lat_4 <- readRDS("long_lat_4.rds")


```


```{r}

long_lat_5$census_code <- apply(long_lat_5, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))


```


```{r}


saveRDS(long_lat_5, "long_lat_5.rds")



```


```{r}



long_lat_5 <- readRDS("long_lat_5.rds")


```


```{r}


long_lat_6$census_code <- apply(long_lat_6, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))




```



```{r}

saveRDS(long_lat_6, "long_lat_6.rds")



```


```{r}

long_lat_6 <- readRDS("long_lat_6.rds")


```


```{r}




long_lat_7$census_code <- apply(long_lat_7, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))



```


```{r}

saveRDS(long_lat_7, "long_lat_7.rds")


```


```{r}


long_lat_7 <- readRDS("long_lat_7.rds")

```


```{r}



long_lat <- rbind(long_lat_1, long_lat_2, long_lat_3, long_lat_4, long_lat_5, long_lat_6)



```


```{r}



long_lat$GEOID <- substr(long_lat$census_code, 1,11)


```


```{r}


saveRDS(long_lat, "long_lat.rds")

```


```{r}


long_lat <- readRDS("long_lat.rds")


```

```{r}


long_lat |>  head()


```


```{r}


census_geo <- long_lat |> 
    select(-LATITUDE, -LONGITUDE)


```

```{r}

saveRDS(census_geo, "census_geo.rds")


```


```{r}

census_geo <- readRDS("census_geo.rds")



```


```{r}




census_geo |> head()
```


```{r}



ma_accidents_16_21_4 <- cbind(ma_accidents_16_21_3, census_geo)



```

```{r}



saveRDS(ma_accidents_16_21_4, "ma_accidents_16_21_4.rds")


```


```{r}


ma_accidents_16_21_4 <- readRDS("ma_accidents_16_21_4.rds")


```




```{r}



ma_accidents_16_21_4 |> 
    summary()

```

There are five mssing values in the newly created GEOID.  These missing values (NAs) will be removed. 


```{r}


ma_accidents_16_21_5 <- ma_accidents_16_21_4 |> 
    filter(!is.na(GEOID))


```

```{r}


ma_accidents_16_21_5 |> 
    select(GEOID, census_code) |> 
    summary()


```

```{r}


ma_accidents_16_21_5 |> 
    filter(!is.na(Wind_Speed.mph.)) |> 
    group_by(Severity) |> 
    summarise(median_wind_speed = median(Wind_Speed.mph.))


```


```{r}


ma_accidents_16_21_5 <- ma_accidents_16_21_5 |> 
    filter(!is.na(Wind_Speed.mph.))


```

```{r}

ma_accidents_16_21_5 |> 
    summary()


```

```{r}


ma_accidents_16_21_5 |> 
    filter(!is.na(Visibility.mi.)) |> 
    group_by(Severity) |> 
    summarise(median_visibility = median(Visibility.mi.))


```

```{r}


ma_accidents_16_21_5 |> 
    filter(!is.na(Precipitation.in.)) |> 
    group_by(Severity) |> 
    summarise(median_visibility = median(Precipitation.in.))


```

```{r}


ma_accidents_16_21_5 |> 
    select(id, Description) |> 
    filter(id == "51113") 


```

```{r}


ma_accidents_16_21_5  |> filter(accident_duration > 1000)  |> head()


```

```{r}

ma_accidents_16_21_5  |> filter(accident_duration > 1000)  |> tail()


```

```{r}


ma_accidents_16_21_5  |> filter(accident_duration <=1000) |> dim()


```


```{r}


ma_accidents_16_21_5 <- ma_accidents_16_21_5 |> 
    select(-Start_Lat, -End_Lat, -Start_Lng, -End_Lng, -State, -Zipcode, -Country, -Timezone, -Airport_Code, -Weather_Timestamp, -Wind_Chill.F., -Visibility.mi., -Precipitation.in.)


```


```{r}


saveRDS(ma_accidents_16_21_5, "ma_accidents_16_21_5.rds")

```

```{r}
#| echo: false


ma_accidents_16_21_5 <- readRDS("ma_accidents_16_21_5.rds")


```






```{r}


ma_accidents_16_21_5 |> 
    count(County)


```




```{r}


ma_county <- counties("MA", year = 2024)

```

```{r}

saveRDS(ma_county, "ma_county.rds")


```

```{r}


ma_county <- readRDS("ma_county.rds")


```

```{r}


ma_county


```

```{r}


ma_tracts <- tracts("MA", year = 2024)


```

```{r}


saveRDS(ma_tracts, "ma_tracts.rds")


```

```{r}

ma_tracts <- readRDS("ma_tracts.rds")


```

```{r}

ma_tracts


```



