---
title: "Untitled"
format: html
---

# Introduction



# Libraries

```{r}

library(dplyr)
library(ggplot2)
library(plotly)
library(RColorBrewer)
library(forcats)
library(tidycensus)

library(sf)

library(jsonlite)

library(tigris)

library(stringr)

library(leaflet)
library(htmlwidgets)
library(htmltools)
library(gt)
library(DT)
library(GGally)
library(lme4)
library(lmerTest)
library(rstanarm)
library(brms)
library(broom)
library(lubridate)
library(tidymodels)





```


# Data Inspection and Cleaning

## US Accident Data


```{r}


US_accidents <- read.csv("/Users/francislauder/Documents/Data Sets/US Acciidents/US_Accidents_March23.csv")


```


```{r}


saveRDS(US_accidents, "US_accidents.rds")

```

```{r}


US_accidents <- readRDS("US_accidents.rds")


```



```{r}


glimpse(US_accidents)


```



```{r}


MA_accidents <- US_accidents |> 
    filter(State == "MA")


```


```{r}


saveRDS(MA_accidents, "MA_accidents.rds")



```

```{r}


MA_accidents <- readRDS("MA_accidents.rds")


```

```{r}


MA_accidents |> glimpse()


```

```{r}


summary(MA_accidents)


```


```{r}


dim(MA_accidents)


```

```{r}


summary(MA_accidents)


```

```{r}



MA_accidents |> 
    count(Severity) |> 
    mutate(perc = round(n / sum(n),3)) |> 
    arrange(desc(perc))


```

```{r}



MA_accidents <- MA_accidents %>% 
  mutate(Weather_Timestamp = as.Date(Weather_Timestamp))

```



Some features are in the correct data types for use in our analysis.  We will tranform them to the required data types. 



First, Sverity, which is our outcome feature, will be transformed to a factor from an integer data type. 

```{r}




MA_accidents_2 <- MA_accidents |> 
    mutate(Severity = as.factor(Severity))


```



```{r}


class(MA_accidents_2$Severity)


```




We will now extract the year, month,day, and hourm of the day from the Start_Time feature. 





```{r}



MA_accidents_2 <- MA_accidents_2 |> 
    mutate(Year = year(Start_Time)) |> 
    mutate(Month = month(Start_Time)) |> 
    mutate(day_of_week = wday(Start_Time)) |> 
    mutate(hour_of_day = hour(Start_Time))


```


```{r}


MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day) |> 
    head()


```

Day of week starts with Sunday (1) and ends with Saturday (7)

Hour of day starts at 1:00 am (1) and ends at 12:00 am (24)




```{r}


MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day) |> 
    glimpse()



```



```{r}


MA_accidents_2 <- MA_accidents_2 |> 
    mutate(Year = as.factor(Year)) |> 
    mutate(Month = as.factor(Month)) |> 
    mutate(day_of_week = as.factor(day_of_week))



```



```{r}


MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day) |> 
    glimpse()


```

For us to be able to calculate the time duration of accidents we'll first need to onverty the start and endtimes to a date time object that includes year, month, day, hours, minutes, and seconds. 


```{r}


MA_accidents_2 <- MA_accidents_2 |> 
    mutate(Start_Time = ymd_hms(Start_Time)) |> 
    mutate(End_Time = ymd_hms(End_Time))


```

```{r}

MA_accidents_2 |> 
    select(Start_Time, End_Time) |> 
    glimpse()


```


```{r}


MA_accidents_2 |> 
    mutate(accident_duration = round(difftime(End_Time,Start_Time, units = "mins"), 0))|> 
    head()


```

From the output, we see the accident duration is in minutes. 



```{r}

saveRDS(MA_accidents_2, "MA_accidents_2.rds")


```


```{r}


MA_accidents_2 <- readRDS("MA_accidents_2.rds")


```


```{r}



MA_accidents_2$LATITUDE <- MA_accidents_2$Start_Lat

MA_accidents_2$LONGITUDE <- MA_accidents_2$Start_Lng


```


```{r}



MA_accidents_2 <- MA_accidents_2 |> 
    mutate(id = row_number())




```

## US Census Data


```{r}





long_lat_1 <- MA_accidents_2 |> 
    filter(id <=5000) |> 
    select(LATITUDE, LONGITUDE)



```


```{r}


long_lat_2 <- MA_accidents_2 |> 
    filter(id > 5000 & id <=15000) |> 
    select(LATITUDE, LONGITUDE)



```


```{r}




long_lat_3 <- MA_accidents_2 |> 
    filter(id > 15000 & id <=25000) |> 
    select(LATITUDE, LONGITUDE)

```


```{r}


long_lat_4 <- MA_accidents_2 |> 
    filter(id > 25000 & id <=35000) |> 
    select(LATITUDE, LONGITUDE)


```

```{r}


long_lat_5 <- MA_accidents_2 |> 
    filter(id > 35000 & id <=45000) |> 
    select(LATITUDE, LONGITUDE)



```

```{r}


long_lat_6 <- MA_accidents_2 |> 
    filter(id > 45000 & id <=55000) |> 
    select(LATITUDE, LONGITUDE)



```

```{r}



long_lat_7 <- MA_accidents_2 |> 
    filter(id > 55000) |> 
    select(LATITUDE, LONGITUDE)


```



```{r}


Sys.getenv("CENSUS_API_KEY")


```


```{r}


long_lat_1$census_code <- apply(long_lat_1, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))




```


```{r}


saveRDS(long_lat_1 , "long_lat_1.rds")


```

```{r}


long_lat_1  <- readRDS("long_lat_1.rds")


```


```{r}

long_lat_2$census_code <- apply(long_lat_2, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))


```

```{r}

saveRDS(long_lat_2, "long_lat_2.rds")

```

```{r}


long_lat_2 <- readRDS("long_lat_2.rds")


```


```{r}



long_lat_3$census_code <- apply(long_lat_3, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))


```



```{r}


saveRDS(long_lat_3, "long_lat_3.rds")


```


```{r}


long_lat_3 <- readRDS("long_lat_3.rds")



```


```{r}


long_lat_4$census_code <- apply(long_lat_4, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))


```


```{r}



saveRDS(long_lat_4, "long_lat_4.rds")



```


```{r}


long_lat_4 <- readRDS("long_lat_4.rds")


```


```{r}

long_lat_5$census_code <- apply(long_lat_5, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))


```


```{r}


saveRDS(long_lat_5, "long_lat_5.rds")



```


```{r}



long_lat_5 <- readRDS("long_lat_5.rds")


```


```{r}


long_lat_6$census_code <- apply(long_lat_6, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))




```



```{r}

saveRDS(long_lat_6, "long_lat_6.rds")



```


```{r}

long_lat_6 <- readRDS("long_lat_6.rds")


```


```{r}




long_lat_7$census_code <- apply(long_lat_7, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))



```


```{r}

saveRDS(long_lat_7, "long_lat_7.rds")


```


```{r}


long_lat_7 <- readRDS("long_lat_7.rds")

```


```{r}



long_lat <- rbind(long_lat_1, long_lat_2, long_lat_3, long_lat_4, long_lat_5, long_lat_6, long_lat_7)



```


```{r}



long_lat$GEOID <- substr(long_lat$census_code, 1,11)


```


```{r}


saveRDS(long_lat, "long_lat.rds")

```


```{r}


long_lat <- readRDS("long_lat.rds")


```

```{r}


long_lat |>  head()


```


```{r}


census_geo <- long_lat |> 
    select(-LATITUDE, -LONGITUDE)


```

```{r}

saveRDS(census_geo, "census_geo.rds")


```


```{r}

census_geo <- readRDS("census_geo.rds")



```


```{r}




census_geo |> head()
```


```{r}



MA_accidents_3 <- cbind(MA_accidents_2, census_geo)



```

```{r}



saveRDS(MA_accidents_3, "MA_accidents_3.rds")


```


```{r}


MA_accidents_3 <- readRDS("MA_accidents_3.rds")


```

```{r}



MA_accidents_3 |> 
    summary()

```


```{r}


ma_county <- counties("MA", year = 2024)

```

```{r}

saveRDS(ma_county, "ma_county.rds")


```

```{r}


ma_county <- readRDS("ma_county.rds")


```

```{r}


ma_county


```

```{r}


ma_tracts <- tracts("MA", year = 2024)


```

```{r}


saveRDS(ma_tracts, "ma_tracts.rds")


```

```{r}

ma_tracts <- readRDS("ma_tracts.rds")


```

```{r}

ma_tracts


```



