---
title: "Untitled"
format: html
---

# Introduction



# Libraries

```{r}

library(dplyr)
library(ggplot2)
library(plotly)
library(RColorBrewer)
library(forcats)
library(tidycensus)

library(sf)

library(jsonlite)

library(tigris)

library(stringr)

library(leaflet)
library(htmlwidgets)
library(htmltools)
library(gt)
library(DT)
library(GGally)
library(lme4)
library(lmerTest)
library(rstanarm)
library(brms)
library(broom)
library(lubridate)





```


# Data Inspection and Cleaning

## US Accident Data


```{r}


US_accidents <- read.csv("/Users/francislauder/Documents/Data Sets/US Acciidents/US_Accidents_March23.csv")


```


```{r}


saveRDS(US_accidents, "US_accidents.rds")

```

```{r}


US_accidents <- readRDS("US_accidents.rds")


```



```{r}


glimpse(US_accidents)


```



```{r}


MA_accidents <- US_accidents |> 
    filter(State == "MA")


```


```{r}


saveRDS(MA_accidents, "MA_accidents.rds")



```

```{r}


MA_accidents <- readRDS("MA_accidents.rds")


```

```{r}


MA_accidents |> glimpse()


```

```{r}


summary(MA_accidents)


```


```{r}


dim(MA_accidents)


```

```{r}


summary(MA_accidents)


```

```{r}



MA_accidents |> 
    count(Severity) |> 
    mutate(perc = round(n / sum(n),3)) |> 
    arrange(desc(perc))


```

```{r}



MA_accidents <- MA_accidents %>% 
  mutate(Weather_Timestamp = as.Date(Weather_Timestamp))

```



Some features are in the correct data types for use in our analysis.  We will tranform them to the required data types. 



First, Sverity, which is our outcome feature, will be transformed to a factor from an integer data type. 

```{r}




MA_accidents_2 <- MA_accidents |> 
    mutate(Severity = as.factor(Severity))


```



```{r}


class(MA_accidents_2$Severity)


```




We will now extract the year, month,day, and hourm of the day from the Start_Time feature. 





```{r}



MA_accidents_2 <- MA_accidents_2 |> 
    mutate(Year = year(Start_Time)) |> 
    mutate(Month = month(Start_Time)) |> 
    mutate(day_of_week = wday(Start_Time)) |> 
    mutate(hour_of_day = hour(Start_Time))


```


```{r}


MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day) |> 
    head()


```

Day of week starts with Sunday (1) and ends with Saturday (7)

Hour of day starts at 1:00 am (1) and ends at 12:00 am (24)




```{r}


MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day) |> 
    glimpse()



```



```{r}


MA_accidents_2 <- MA_accidents_2 |> 
    mutate(Year = as.factor(Year)) |> 
    mutate(Month = as.factor(Month)) |> 
    mutate(day_of_week = as.factor(day_of_week))



```



```{r}


MA_accidents_2 |> 
    select(Year, Month, day_of_week, hour_of_day) |> 
    glimpse()


```

For us to be able to calculate the time duration of accidents we'll first need to onverty the start and endtimes to a date time object that includes year, month, day, hours, minutes, and seconds. 


```{r}


MA_accidents_2 <- MA_accidents_2 |> 
    mutate(Start_Time = ymd_hms(Start_Time)) |> 
    mutate(End_Time = ymd_hms(End_Time))


```

```{r}

MA_accidents_2 |> 
    select(Start_Time, End_Time) |> 
    glimpse()


```


```{r}


MA_accidents_2 |> 
    mutate(accident_duration = round(difftime(End_Time,Start_Time, units = "mins"), 0))|> 
    head()


```

From the output, we see the accident duration is in minutes. 



```{r}

saveRDS(MA_accidents_2, "MA_accidents_2.rds")


```


```{r}


MA_accidents_2 <- readRDS("MA_accidents_2.rds")


```


```{r}






```


```{r}








```

## US Census Data


```{r}


census_api_key("2f743d3250c4e6afe611f8686d00b9abcd000a05", install = TRUE)






```



```{r}


Sys.getenv("CENSUS_API_KEY")


```